---
title: "LRP2 Lite Testing"
output: html_document
---

```{r}
source("/project/sheynkman/users/megan/github/megan-schertzer/ms_functions.R")

#wget https://github.com/conda-forge/miniforge/releases/download/24.11.3-0/Miniforge3-24.11.3-0-Linux-x86_64.sh
#packageVersion()
# added genomic ranges and tidyverse to sqanti3 env

```

What does sqanti classification file have?
```{r}
all = read_tsv("../results/sqanti/sample_classification.txt")

gtf_file_path = "/project/sheynkman/external_data/GENCODE_v47/gencode.v47.annotation.gtf"

gr <- import(gtf_file_path, format = "gtf")
gtf_data <- as.data.table(gr)

reference_df = gtf_data %>% filter(type == "transcript") %>% select(gene_id, gene_type, gene_name, transcript_id, transcript_type, transcript_name, protein_id)
tail(reference_df)
```



sqanti_filter.R
```{r}
#!/usr/bin/env Rscript

#' SQANTI3 Transcript Filtering Script
#' 
#' R script to filter SQANTI3 transcripts based on quality criteria:
#' - Protein-coding genes only
#' - Remove transcripts with internal priming 
#' - Remove template switching artifacts
#' - Filter by SQANTI structural categories
#'
#' Outputs both filtered files and dropout files with reasons.

# =============================================================================
# Load required libraries
# =============================================================================

suppressPackageStartupMessages({
  library(tidyverse)
  library(rtracklayer)
  library(Biostrings) 
  library(argparse)
  library(data.table)
})

# =============================================================================
# Define SQANTI structural categories
# =============================================================================

STRUCTURAL_CATEGORIES <- list(
  strict = c("novel_not_in_catalog", "novel_in_catalog", 
             "incomplete-splice_match", "full-splice_match"),
  all = c("antisense", "novel_not_in_catalog", "novel_in_catalog",
          "incomplete-splice_match", "full-splice_match", "genic",
          "intergenic", "fusion", "genic_intron")
)

# =============================================================================
# Helper functions
# =============================================================================

#' Save filtered sequences to FASTA
#' @param fasta_file Input FASTA file
#' @param keep_ids IDs to keep
#' @param output_file Output FASTA file
save_filtered_fasta <- function(fasta_file, keep_ids, output_file) {
  # Read FASTA
  sequences <- readDNAStringSet(fasta_file)
  
  # Filter sequences
  keep_mask <- names(sequences) %in% keep_ids
  filtered_sequences <- sequences[keep_mask]
  
  # Write filtered FASTA
  writeXStringSet(filtered_sequences, output_file)
  message(paste0("Saved ", length(filtered_sequences), " sequences to ", basename(output_file)))
}

#' Save filtered GTF
#' @param gtf_file Input GTF file
#' @param keep_ids Transcript IDs to keep
#' @param output_file Output GTF file
save_filtered_gtf <- function(gtf_file, keep_ids, output_file) {
  # Read GTF
  gtf <- import(gtf_file)
  
  # Filter by transcript_id
  filtered_gtf <- gtf[gtf$transcript_id %in% keep_ids]
  
  # Write filtered GTF
  export(filtered_gtf, output_file)
  message(paste0("Saved ", length(unique(filtered_gtf$transcript_id)), " transcripts to ", basename(output_file)))
}

# =============================================================================
# Get environment variables for filtering parameters
# =============================================================================
filter_protein_coding = TRUE
filter_internal_priming = TRUE
filter_RTS = TRUE
percent_polyA_threshold = 60

filter_protein_coding <- aTRUEfilter_protein_coding <- as.logical(Sys.getenv("PROTEIN_CODING_FILTER", "TRUE"))
filter_internal_priming <- as.logical(Sys.getenv("INTERNAL_PRIMING_FILTER", "TRUE"))
filter_RTS <- as.logical(Sys.getenv("TEMPLATE_SWITCHING_FILTER", "TRUE"))
percent_polyA_threshold <- as.numeric(Sys.getenv("PERCENT_POLYA_THRESHOLD", "60"))
structural_level <- Sys.getenv("STRUCTURE_FILTER", "strict")

basename <- Sys.getenv("OUTPUT_BASE_NAME")
dir <- file.path(Sys.getenv("OUTPUT_DIR"), "sqanti") 
gencode_gtf <- Sys.getenv("GENCODE_GTF_FILE")

gencode_gtf <- "/project/sheynkman/external_data/GENCODE_v47/gencode.v47.annotation.gtf"

# =============================================================================
# Check required files
# =============================================================================

classification_file <- "/home/cwp5au/megan/github/LRP2_lite/results/sqanti/sample_classification.txt"
classification_file <- paste0(dir, "/", basename, "_classification.txt")
stopifnot("File not found" = file.exists(classification_file))

sqanti_gtf <- "/home/cwp5au/megan/github/LRP2_lite/results/sqanti/sample_corrected.gtf"
sqanti_gtf <- paste0(dir, "/", basename, "_corrected.gtf")
stopifnot("File not found" = file.exists(sqanti_gtf))
 
sqanti_fasta <- "/home/cwp5au/megan/github/LRP2_lite/results/sqanti/sample_corrected.fasta"
sqanti_fasta <- paste0(dir, "/", basename, "_corrected.fasta")
stopifnot("File not found" = file.exists(sqanti_fasta))

# Create additional output directory
dropout_dir <- file.path(dir, "dropout")
dir.create(dropout_dir, recursive = TRUE, showWarnings = FALSE)

# =============================================================================
# Load and merge inputs
# =============================================================================

# Gencode gtf used to select protein coding genes and add column of readable gene name
gencode    <- import(gencode_gtf, format = "gtf")
gencode_df <- as.data.table(gencode)

gencode_df %<>% 
  filter(type == "transcript") %>% 
  select(associated_gene = gene_id, gene_type, gene_name, associated_transcript = transcript_id, transcript_name, protein_id)

gencode_gene = gencode_df %>% 
  select(associated_gene, gene_type, gene_name) %>% 
  distinct()

# SQANTI classification file- keep only isoforms with ENS gene IDs
sqanti_df_full <- read_tsv(classification_file, show_col_types = FALSE) %>%
  filter(!is.na(associated_gene), str_starts(associated_gene, "ENS"))

sqanti_df_full %<>% 
  left_join(gencode_gene) %>% 
  left_join(gencode_df)

# =============================================================================
# Apply filters sequentially
# =============================================================================

# Initialize dropout tracking
original_ids <- sqanti_df_full$isoform
dropout_tracker <- list()

# Keep only transcripts of protein coding genes
sqanti_df = sqanti_df_full # 653,675
if (filter_protein_coding) {
  ids <- sqanti_df$isoform
  sqanti_df %<>% 
    filter(gene_type == "protein_coding")
  
  kept_ids    <- sqanti_df$isoform
  dropped_ids <- setdiff(ids, kept_ids)
  dropout_tracker[["not_protein_coding"]] <- dropped_ids
  
  message("Protein coding filter: kept ", length(kept_ids), " transcripts, dropped ", length(dropped_ids), " transcripts")
}

# Filter based on downstream polyA sequence to remove internal priming
if (filter_internal_priming) {
  ids <- sqanti_df$isoform
  
  # keep all isoforms below threshold
  keep1 = sqanti_df %>% 
    filter(perc_A_downstream_TTS <= percent_polyA_threshold)

  # Get downstream sequence for FSM above threshold
  protected_seqs = sqanti_df %>%
    filter(perc_A_downstream_TTS > percent_polyA_threshold, structural_category == "full-splice_match") %>%
    distinct(associated_gene, seq_A_downstream_TTS)

  # Keep high polyA isoforms that share sequence with FSM transcript
  keep2 = sqanti_df %>% 
    inner_join(protected_seqs, by = c("associated_gene", "seq_A_downstream_TTS"))

  sqanti_df = bind_rows(keep1, keep2) %>% distinct()
  
  kept_ids    <- sqanti_df$isoform
  dropped_ids <- setdiff(ids, kept_ids)
  dropout_tracker[["internal_priming"]] <- dropped_ids
  
  message("Internal priming filter: kept ", length(kept_ids), " transcripts, dropped ", length(dropped_ids), " transcripts")
}

# Filter out template switching, keep RTS_stage = FALSE (no RTS)
if (filter_RTS) {
  ids <- sqanti_df$isoform
  
  # sum read counts
  # Find column positions
  ratio_tss_pos <- which(colnames(sqanti_df) == "ratio_TSS")
  gene_type_pos <- which(colnames(sqanti_df) == "gene_type")

  # Get columns between them (if any exist)
  if (gene_type_pos - ratio_tss_pos > 1) {
    count_cols <- (ratio_tss_pos + 1):(gene_type_pos - 1)
    sqanti_df$avg_counts <- rowMeans(sqanti_df[, count_cols], na.rm = TRUE)
  } else {
    sqanti_df$avg_counts <- NA  # No columns in between
  }
  
  # filtering considerations
  sqanti_df %<>% filter(
    structural_category == "full-splice_match" |
    RTS_stage == FALSE | 
    (RTS_stage == TRUE & all_canonical == TRUE & avg_counts > 3))
  
  kept_ids <- sqanti_df$isoform
  dropped_ids <- setdiff(ids, kept_ids)
  dropout_tracker[["Template_switching_artifact"]] <- dropped_ids
  
  message("Template switching filter: kept ", length(kept_ids), " transcripts, dropped ", length(dropped_ids), " transcripts")
}
  
# Apply structural category filter
allowed_categories <- STRUCTURAL_CATEGORIES[[structural_level]]
ids <- sqanti_df$isoform
sqanti_df %<>% filter(structural_category %in% allowed_categories)

kept_ids <- sqanti_df$isoform
dropped_ids <- setdiff(ids, kept_ids)
dropout_tracker[["structural_category_filtered"]] <- dropped_ids
message("Structural category filter: kept ", length(kept_ids), " transcripts, dropped ", length(dropped_ids), " transcripts")

# Calculate final results
all_dropout_ids <- setdiff(original_ids, kept_ids)
  
message(paste0("Final: ", length(kept_ids), " kept, ", length(all_dropout_ids), " dropped"))

# =============================================================================
# Write output files and generate summary
# =============================================================================

message("Writing output files...")
write_tsv(sqanti_df, file.path(dir, paste0(basename, "_classification_filtered.txt")))
  
save_filtered_fasta(sqanti_fasta, kept_ids, file.path(dir, paste0(basename, "_corrected_filtered.fasta")))
  
save_filtered_gtf(sqanti_gtf, kept_ids, file.path(dir, paste0(basename, "_corrected_filtered.gtf")))
  
# Save dropout files
dropout_reasons_df <- map_dfr(names(dropout_tracker), ~ {
  tibble(isoform = dropout_tracker[[.x]], dropout_reason = .x)
})

write_tsv(dropout_reasons_df, file.path(dropout_dir, paste0(basename, "_dropout_transcripts.tsv")))
save_filtered_fasta(sqanti_fasta, all_dropout_ids, file.path(dropout_dir, paste0(basename, "_corrected_dropout.fasta")))
save_filtered_gtf(sqanti_gtf, all_dropout_ids, file.path(dropout_dir, paste0(basename, "_corrected_dropout.gtf")))
  
# Print summary
cat("\n=== FILTERING SUMMARY ===\n")
cat(paste0("Sample: ", basename, "\n"))
cat(paste0("Original transcripts: ", length(original_ids), "\n"))
cat(paste0("Final transcripts: ", length(kept_ids), "\n"))
cat(paste0("Retention rate: ", round(100 * length(kept_ids) / length(original_ids), 1), "%\n"))
cat("\nDropout breakdown:\n") 

dropout_summary <- dropout_reasons_df %>%
  count(dropout_reason, sort = TRUE)
  
for (i in seq_len(nrow(dropout_summary))) {
  cat(paste0("  ", dropout_summary$dropout_reason[i], ": ", dropout_summary$n[i], "\n"))
    }

test %>% filter(structural_category == "full-splice_match", gene_type == "protein_coding", gene_name == "RBFOX2")
sqanti_df_full %>% filter(gene_name == "RBFOX2")
sqanti_df_full %>% filter(RTS_stage == TRUE)
sqanti_df %>% filter(RTS_stage == FALSE | structural_category == "full-splice_match")
sqanti_df %>% filter(RTS_stage == TRUE & structural_category != "full-splice_match" & all_canonical == "canonical")





sqanti_df %>% filter(perc_A_downstream_TTS <= percent_polyA_threshold | structural_category == "full-splice_match")
sqanti_df %>% filter(perc_A_downstream_TTS > percent_polyA_threshold & structural_category == "full-splice_match")
sqanti_df %>% filter(associated_gene == "ENSG00000215788.11") %>% group_by(associated_gene, seq_A_downstream_TTS) %>% tally()

dropout_reasons_df %>% group_by(dropout_reason) %>% tally()


```

Why are there so many RT switching in the AML long read?
- Inspect support depth
- canonical or non-canonical junctions
- microhomology length distribution > 6 are more likely to be real

SQANTI’s RT-switching rule is purely sequence-based: if a junction is flanked by microhomology, it gets flagged. Many true biological junctions happen to sit near repeats, so you can get multiple reads flagged even though they’re real isoforms. This is the most common explanation when you see >1 read support.

A lot of these seem real to me
```{r}
# this is my analysis in LRP2_lite
all = read_tsv("../results/sqanti/sample_classification.txt") # 839,763
switch = all %>% filter(RTS_stage == "TRUE") # 542,120

# expression
ggplot(all, aes(x = RTS_stage, y = FL.BioSample_1)) + geom_violin() + theme_bw() + scale_y_log10()
switch %>% filter(FL.BioSample_1 > 1) # 153,179
switch %>% filter(FL.BioSample_1 > 5) # 27,193

# canonical splice sites, GT-AG
switch %>% group_by(all_canonical) %>% tally() # 509,234/542,120 are non-canonical
switch %>% filter(FL.BioSample_1 > 1) %>% group_by(all_canonical) %>% tally() # 138,400/153,179 are non-canonical
switch %>% filter(FL.BioSample_1 > 5) %>% group_by(all_canonical) %>% tally() # 20,923/27,193 are non-canonical

# structural category
switch %>% group_by(structural_category) %>% tally() # 355,038 are novel not in catalog

# consider all factors
switch %>% filter(FL.BioSample_1 > 1, all_canonical == "canonical") %>% group_by(structural_category) %>% tally() # 14,779


```

Compare to Emily's analysis- similar percentage
```{r}
em = read_tsv("~/sheynkman/projects/LRP_AML/02_sqanti/AML_classification.txt") # 1,689,238
em_switch = em %>% filter(RTS_stage == "TRUE") # 929,199

```

Compare to MDS Mouse data
```{r}
mds = read_tsv("~/sheynkman/projects/LRP_Mohi_project/02_sqanti/MDS_classification.txt") # 838,021
mds_switch = mds %>% filter(RTS_stage == "TRUE") # 443,272

# expression
mds_switch %>% filter(FL.BioSample_1 > 1) # 74,490
mds_switch %>% filter(FL.BioSample_1 > 5) # 15,122

# canonical splice sites, GT-AG
mds_switch %>% group_by(all_canonical) %>% tally() # 440,301/443,272 are non-canonical
mds_switch %>% filter(FL.BioSample_1 > 1) %>% group_by(all_canonical) %>% tally() 
mds_switch %>% filter(FL.BioSample_1 > 5) %>% group_by(all_canonical) %>% tally() 

# structural category
mds_switch %>% group_by(structural_category) %>% tally() # most are novel not in catalog

# consider all factors
mds_switch %>% filter(FL.BioSample_1 > 1, all_canonical == "canonical") %>% group_by(structural_category) %>% tally() # 1,359

```


From 05_orf_calling
- I dont think expression is useful here. Create an ORF-id that concatanates the transcript ids but keep each row a transcript. can aggregate on ORF ids later.
- dont call orfs separately, call all together

```{r}
cpat_fasta_file = "/home/cwp5au/megan/github/LRP2_lite/results/orf_calling/sample_cpat.ORF_seqs.fa"
cpat_results_file = "/home/cwp5au/megan/github/LRP2_lite/results/orf_calling/sample_cpat.ORF_prob.tsv"
sample_full_fasta_path = "/home/cwp5au/megan/github/LRP2_lite/results/sqanti/sample_corrected_filtered.fasta"
classification_file = "/home/cwp5au/megan/github/LRP2_lite/results/sqanti/sample_classification_filtered.txt"
sample_gtf_path = "/home/cwp5au/megan/github/LRP2_lite/results/sqanti/sample_corrected_filtered.gtf"
gencode_gtf_path = "/project/sheynkman/external_data/GENCODE_v47/gencode.v47.annotation.gtf"
cpat_dir = "/home/cwp5au/megan/github/LRP2_lite/results/orf_calling"
#sample_gtf = import("../results/sqanti/sample_corrected_filtered.gtf") %>% as.data.frame()
#gencode_gtf = import("/project/sheynkman/external_data/GENCODE_v47/gencode.v47.annotation.gtf") %>% as.data.frame()
basename = "sample"

#!/usr/bin/env Rscript

#' ORF Analysis Script - R Version
#' 
#' Map ORF coordinates to genomic positions
#' Compares to GENCODE start codon annotations
#' Selects the best ORF for each transcript
#'
#' Outputs 

# =============================================================================
# Load required libraries
# =============================================================================

suppressPackageStartupMessages({
  library(tidyverse)
  library(rtracklayer)
  library(Biostrings)
})

# =============================================================================
# Get environment variables
# =============================================================================

basename         <- Sys.getenv("OUTPUT_BASE_NAME")
sqanti_dir       <- file.path(Sys.getenv("OUTPUT_DIR"), "sqanti") 
cpat_dir         <- file.path(Sys.getenv("OUTPUT_DIR"), "orf_calling") 
gencode_gtf_path <- Sys.getenv("GENCODE_GTF_FILE")
coding_threshold <- as.numeric(Sys.getenv("CPAT_CODING_THRESHOLD", "0.364"))

# =============================================================================
# Check required files
# =============================================================================

cpat_fasta_file <- file.path(cpat_dir, paste0(basename, "_cpat.ORF_seqs.fa"))
stopifnot("Input ORF FASTA file not found" = file.exists(cpat_fasta_file))

cpat_results_file <- file.path(cpat_dir, paste0(basename, "_cpat.ORF_prob.tsv"))
stopifnot("CPAT results file not found" = file.exists(cpat_results_file))

sample_full_fasta_path <- file.path(sqanti_dir, paste0(basename, "_corrected_filtered.fasta"))
stopifnot("Sample full fasta file not found" = file.exists(sample_full_fasta_path))

sample_gtf_path <- file.path(sqanti_dir, paste0(basename, "_corrected_filtered.gtf"))
stopifnot("Sample gtf file not found" = file.exists(sample_gtf_path))

classification_file <- paste0(sqanti_dir, "/", basename, "_classification_filtered.txt")
stopifnot("File not found" = file.exists(classification_file))

# =============================================================================
# Helper functions
# =============================================================================

#' Check if ORF ends with stop codon
#' @param orf_fasta_file Path to ORF FASTA file
#' @return Data frame with ID and has_stop_codon columns
check_stop_codons <- function(orf_fasta_file) {

  stop_codons <- c("TAG", "TAA", "TGA")
  orfs        <- readDNAStringSet(orf_fasta_file)
  
  tibble(
    ID       = str_split_fixed(names(orfs), "\t", 2)[,1],
    sequence = as.character(orfs)
  ) %>%
    mutate(seq_length     = nchar(sequence),
           last_codon     = if_else(seq_length >= 3, str_sub(sequence, -3, -1), ""),
           has_stop_codon = last_codon %in% stop_codons) %>%
    select(ID, has_stop_codon)
}

#' Map ORF coordinates to genomic positions
#' This is a simplified version that focuses on the core mapping logic
#' @param orf_coords ORF coordinate data
#' @param sample_gtf GTF data for sample transcripts
#' @param gencode_gtf GTF data for GENCODE reference
#' @param sample_fasta Sample transcript sequences
#' @return Mapped ORF data with genomic coordinates
map_orfs_to_genome <- function(orf_coords, sample_gtf, gencode_gtf, full_fasta) {

  # Extract exon information from sample GTF
  sample_exons = sample_gtf %>%
    filter(type == "exon") %>%
    select(transcript_id, seqnames, start, end, strand) %>%
    mutate(exon_length = end - start + 1) %>%
    arrange(transcript_id, if_else(strand == "+", start, -start))
  
  # Calculate cumulative exon positions for coordinate mapping
  sample_exons %<>%
    group_by(transcript_id) %>%
    mutate(cumulative_length = cumsum(exon_length), prior_length = lag(cumulative_length, default = 0)) %>%
    ungroup()
  
  # Map ORF coordinates to genomic positions and pull number of upstream ATGs- strandedness here is tricky
  mapped_orfs = orf_coords %>%
    left_join(sample_exons, by = c("isoform_id" = "transcript_id")) %>%
    filter(ORF_start > prior_length & ORF_start <= cumulative_length) %>% # which exon has the ORF start?
    left_join(full_fasta, by = "isoform_id") %>%
    mutate(genomic_start = ifelse(strand == "+",
                                  start + (ORF_start - prior_length) - 1,
                                  end - (ORF_start - prior_length) + 1)) %>%
    mutate(upstream_atgs = ifelse(!is.na(full_sequence), 
                                  str_count(substr(full_sequence, 1, ORF_start - 1), "ATG"),
                                  Inf)) %>%
    select(-full_sequence)
  
  # Check for GENCODE start codon matches- looking for any match here
  gencode_start_codons = gencode_gtf %>%
    filter(type == "start_codon") %>% 
    mutate(gencode_pos = ifelse(strand == "+", start, end)) %>%
    select(seqnames, strand, gencode_pos) %>%
    distinct() %>%
    mutate(gencode_match = TRUE)
  
  mapped_orfs %<>% left_join(gencode_start_codons, by = c("seqnames", "strand", "genomic_start" = "gencode_pos")) %>%
    mutate(gencode_match = ifelse(is.na(gencode_match), FALSE, gencode_match))
  
  return(mapped_orfs)
}

#' Select best ORF for each transcript
#' Logic: Prefer GENCODE matches with fewest upstream ATGs, otherwise highest coding score
#' @param mapped_orfs Data frame of mapped ORFs
#' @return Data frame with best ORF per transcript
call_best_orfs <- function(mapped_orfs) {

  # ORF quality scoring parameters but these don't seem to be used
  coding_threshold  # CPAT protein-coding threshold
  atg_shift  <- 10          # Sigmoid parameters for ATG penalty
  atg_growth <- 0.5
  
  # per transcript, if there is a gencode match, pick the one with the fewest upstream ATGs
  mapped_orfs %<>% 
    mutate(atg_penalty     = 1 - 1/(1 + exp(-atg_growth * (upstream_atgs - atg_shift))), 
           composite_score = Coding_prob * (1 - atg_penalty),
           orf_quality     = ifelse(!has_stop_codon, "No Stop Codon", 
                                    ifelse(Coding_prob <= coding_threshold, "Low Quality ORF", "Plausible ORF")))
  
  best_orfs = mapped_orfs %>% 
    filter(orf_quality == "Plausible ORF") %>%
    group_by(isoform_id) %>% 
    arrange(desc(gencode_match), upstream_atgs, desc(Coding_prob)) %>%
    slice_head(n = 1) %>%
    ungroup()
  
  mapped_orfs %<>%
    mutate(orf_quality = ifelse(ID %in% best_orfs$ID, "Clear Best ORF", orf_quality))
  
  return(mapped_orfs)
}

#' Translate ORF sequences to proteins and group by identical sequences
#' @param orfs Data frame with ORF coordinates
#' @param transcript_seqs Named vector of transcript sequences
#' @return Data frame with protein sequences and grouped transcript IDs
group_by_protein_sequence <- function(filtered_results, full_fasta) {

  orf_proteins = filtered_results %>%
    select(isoform_id, ORF_start, ORF_end) %>%
    left_join(full_fasta, by = "isoform_id") %>%
    filter(!is.na(full_sequence)) %>%
    mutate(orf_dna_sequence = substr(full_sequence, ORF_start, ORF_end))
  
  # Vectorized translation for speed
  dna      <- DNAStringSet(orf_proteins$orf_dna_sequence)
  proteins <- translate(dna, if.fuzzy.codon = "solve")
  
  orf_proteins %<>%
    mutate(orf_protein_sequence = as.character(proteins)) %>%
    filter(orf_protein_sequence != "") %>%
    select(isoform_id, orf_protein_sequence)
  
  # Group transcripts by identical protein sequences
  orf_groups = orf_proteins %>%
    group_by(orf_protein_sequence) %>%
    arrange(isoform_id) %>%
    mutate(
      orf_isoform_id = paste(isoform_id, collapse = " | "),
      orf_base_id    = isoform_id[1]  # First transcript as representative
    ) %>% 
    ungroup()
  
  return(orf_groups)
}

# =============================================================================
# Main code broken down into steps
# =============================================================================

# === STEP 1: Load all input data ===
message("\n--- STEP 1: Loading input data ---")
  
# CPAT results
cpat_results = read_tsv(cpat_results_file, show_col_types = FALSE) %>%
  separate(ID, into = c("isoform_id", "orf_rank"), sep = "_ORF_", remove = FALSE)

# Load GTF files
sample_gtf  = import(sample_gtf_path) %>% as.data.frame()
gencode_gtf = import(gencode_gtf_path) %>% as.data.frame()
  
# Load sample fasta with full sequence- include UTRs, filtered transcripts from SQANTI
full = readDNAStringSet(sample_full_fasta_path)
full_fasta = tibble(
  isoform_id    = names(full),
  full_sequence = as.character(full)
)

# SQANTI classification
classification = read_tsv(classification_file, show_col_types = FALSE)
  
# === STEP 2: Map ORFs to genomic coordinates ===
message("\n--- STEP 2: Mapping ORFs to genome ---")

stop_codon_status = check_stop_codons(cpat_fasta_file) # uses ORF (dna) fasta to return new column about stop codon 
orf_coords        = left_join(cpat_results, stop_codon_status, by = "ID")
mapped_orfs       = map_orfs_to_genome(orf_coords, sample_gtf, gencode_gtf, full_fasta)
  
# === STEP 3: Define the clear best ORF and print out all classified ORFs ===
message("\n--- STEP 3: Calling best ORFs and filtering---")
  
mapped_orfs_classified = call_best_orfs(mapped_orfs)
mapped_orfs_classified %<>% left_join(select(classification, 
                                            isoform_id = isoform, 
                                            structural_category, 
                                            gene_id = associated_gene, 
                                            transcript_id = associated_transcript,
                                            gene_name, 
                                            transcript_name))

original_n_transcripts = n_distinct(mapped_orfs_classified$isoform_id)

best_orfs = mapped_orfs_classified %>% 
  filter(orf_quality == "Clear Best ORF")

n_best_orfs = nrow(best_orfs %>% distinct(isoform_id))

write_tsv(mapped_orfs_classified, file.path(cpat_dir, paste0(basename, "_all_cpat_orfs_mapped.tsv")))

# === STEP 4: Group transcripts by ORF protein sequence ===
message("\n--- Grouping by ORF protein sequence ---")
orf_groups = group_by_protein_sequence(best_orfs, full_fasta)
best_orfs %<>% left_join(orf_groups)

write_tsv(best_orfs, file.path(cpat_dir, paste0(basename, "_best_cpat_orfs_mapped.tsv")))


```

Checking strand here
```{r}
# testing
mapped_orfs %>% filter(gencode_match == TRUE) %>% group_by(strand) %>% tally()

# negative stranded gene
sample_exons %>% filter(grepl("^PB\\.27\\.", transcript_id)) %>% arrange(desc(end))
test_neg = orf_coords %>% filter(ID == "PB.27.42_ORF_1") %>%
    left_join(sample_exons, by = c("isoform_id" = "transcript_id")) %>%
    filter(ORF_start > prior_length & ORF_start <= cumulative_length) %>%
    left_join(full_fasta, by = "isoform_id") %>%
    mutate(genomic_start = ifelse(strand == "+",
                                  start + (ORF_start - prior_length) - 1,
                                  end - (ORF_start - prior_length) + 1)) %>%
    mutate(upstream_atgs = ifelse(!is.na(full_sequence), 
                                  str_count(substr(full_sequence, 1, ORF_start - 1), "ATG"),
                                  Inf)) %>%
    select(-full_sequence)

gencode_test = gencode_gtf %>%
    filter(type == "start_codon") %>% 
    mutate(gencode_pos = ifelse(strand == "+", start, end)) %>%
    select(gene_id, transcript_id, seqnames, strand, gencode_pos) %>%
    distinct()

gencode_test %>% filter(strand == "-", gencode_pos == 959240)
gencode_gtf %>% filter(gene_id == "ENSG00000188976.11", type == "transcript")

# positive stranded gene
orf_coords %>% filter(ID == "PB.100.1_ORF_1") %>%
    left_join(sample_exons, by = c("isoform_id" = "transcript_id")) %>%
    filter(ORF_start > prior_length & ORF_start <= cumulative_length) %>%
    left_join(full_fasta, by = "isoform_id") %>%
    mutate(genomic_start = ifelse(strand == "+",
                                  start + (ORF_start - prior_length) - 1,
                                  end - (ORF_start - prior_length) + 1)) %>%
    mutate(upstream_atgs = ifelse(!is.na(full_sequence), 
                                  str_count(substr(full_sequence, 1, ORF_start - 1), "ATG"),
                                  Inf)) %>%
    select(-full_sequence)
```

Refine ORFs
```{r}
#!/usr/bin/env Rscript

# ORF Database Refinement Script - R Version
# Groups transcripts with identical protein sequences and filters by quality

suppressPackageStartupMessages({
  library(tidyverse)
  library(Biostrings)
  library(argparse)
})


#' Translate ORF sequences to proteins and group by identical sequences
#' @param orfs Data frame with ORF coordinates
#' @param transcript_seqs Named vector of transcript sequences
#' @return Data frame with protein sequences and grouped transcript IDs
group_by_protein_sequence <- function(filtered_results, full_fasta) {
  message("Translating ORFs and grouping by protein sequence...")
  
  orf_proteins = filtered_results %>%
    select(isoform_id, ORF_start, ORF_end) %>%
    left_join(full_fasta, by = "isoform_id") %>%
    filter(!is.na(full_sequence)) %>%
    mutate(orf_dna_sequence = substr(full_sequence, ORF_start, ORF_end))
    
  # Vectorized translation for speed
  dna <- DNAStringSet(orf_proteins$orf_dna_sequence)
  proteins <- translate(dna, if.fuzzy.codon = "solve")
  
  orf_proteins %<>%
    mutate(orf_protein_sequence = as.character(proteins)) %>%
    filter(orf_protein_sequence != "") %>%
    select(isoform_id, orf_protein_sequence)
  
  
  # Group transcripts by identical protein sequences
  orf_groups = orf_proteins %>%
    group_by(orf_protein_sequence) %>%
    arrange(isoform_id) %>%
    mutate(
      orf_isoform_id = paste(isoform_id, collapse = "|"),
      orf_base_id = isoform_id[1]  # First transcript as representative
    ) %>% 
    ungroup()
  
  return(orf_groups)
}


#' Calculate total expression for grouped transcripts
#' @param protein_groups Data frame with protein groups
#' @param orfs Data frame with expression data
#' @return Data frame with aggregated expression values
aggregate_expression <- function(protein_groups, orfs) {
  message("Aggregating expression values for protein groups...")
  
  # Create lookup tables for FL and CPM values
  fl_lookup <- setNames(orfs$FL, orfs$pb_acc)
  cpm_lookup <- setNames(orfs$CPM, orfs$pb_acc)
  
  # Aggregate expression for each protein group
  expression_summary <- protein_groups %>%
    mutate(
      FL = map_dbl(pb_accs, ~sum(fl_lookup[.x], na.rm = TRUE)),
      CPM = map_dbl(pb_accs, ~sum(cmp_lookup[.x], na.rm = TRUE))
    )
  
  return(expression_summary)
}

#' Main analysis function
main <- function() {

  message("=== Starting ORF Database Refinement ===")

  # === STEP 1: Load input data ===
  message("\n--- Loading input files ---")
  orfs <- read_tsv(args$orfs, show_col_types = FALSE)
  message("Loaded ", nrow(orfs), " ORFs")
  
  all_ids <- unique(orfs$pb_acc)
  message("Total unique transcripts: ", length(all_ids))
  
  # === STEP 2: Filter ORFs by quality ===
  message("\n--- Filtering ORFs by quality ---")
  # Apply quality filters
  filtered_results = best_orfs %>% filter(orf_quality %in% c("Plausible ORF", "Clear Best ORF"))
  
  # === STEP 4: Group transcripts by protein sequence ===
  message("\n--- Grouping by protein sequence ---")
  orf_groups <- group_by_protein_sequence(filtered_results, full_fasta)
  
  
  
  
  # === STEP 5: Aggregate expression data ===
  message("\n--- Aggregating expression data ---")
  protein_groups_with_expr <- aggregate_expression(protein_groups, kept_orfs)
  
  # === STEP 6: Merge with ORF metadata ===
  message("\n--- Merging with ORF metadata ---")
  
  # Get metadata for representative (base) transcripts
  orf_metadata <- kept_orfs %>%
    select(pb_acc, coding_score, orf_score, orf_calling_confidence, upstream_atgs, gene) %>%
    distinct()
  
  final_results <- protein_groups_with_expr %>%
    left_join(orf_metadata, by = c("base_acc" = "pb_acc")) %>%
    select(pb_accs_str, base_acc, protein_sequence, coding_score, orf_calling_confidence,
           upstream_atgs, orf_score, gene, FL, CPM) %>%
    rename(pb_accs = pb_accs_str)
  
  # === STEP 7: Write final outputs ===
  message("\n--- Writing output files ---")
  
  # Write refined protein FASTA
  refined_fasta <- paste0(args$name, "_orf_refined.fasta")
  fasta_lines <- paste0(">pb|", final_results$base_acc, "|fullname GN=", final_results$gene, "\n",
                       final_results$protein_sequence)
  writeLines(fasta_lines, refined_fasta)
  
  # Write refined metadata TSV
  refined_tsv <- paste0(args$name, "_orf_refined.tsv")
  write_tsv(final_results, refined_tsv)
  
  # === SUMMARY ===
  message("\n=== Refinement Complete ===")
  message("Final outputs:")
  message("- Refined FASTA: ", refined_fasta)
  message("- Refined metadata: ", refined_tsv)
  message("- Dropout FASTA: dropout_", basename(args$name), "_orf.fasta")
  message("- Dropout metadata: dropout_", basename(args$name), "_orf.tsv")
  message("")
  message("Summary:")
  message("- Input ORFs: ", nrow(orfs))
  message("- Quality filtered: ", nrow(kept_orfs))
  message("- Final protein groups: ", nrow(final_results))
  message("- Redundancy reduction: ", round((1 - nrow(final_results)/nrow(kept_orfs)) * 100, 1), "%")
}

# Run main function
if (!interactive()) {
  main()
}
```

Problem is that we have redundancy...some transcripts have more than one probably ORF
```{r}
#' Translate ORF sequences to proteins and group by identical sequences
#' @param orfs Data frame with ORF coordinates
#' @param transcript_seqs Named vector of transcript sequences
#' @return Data frame with protein sequences and grouped transcript IDs
group_by_protein_sequence <- function(filtered_results, full_fasta) {
  message("Translating ORFs and grouping by protein sequence...")
  
  orf_proteins = filtered_results %>%
    select(isoform_id, ORF_start, ORF_end) %>%
    left_join(full_fasta, by = "isoform_id") %>%
    filter(!is.na(full_sequence)) %>%
    mutate(orf_dna_sequence = substr(full_sequence, ORF_start, ORF_end))
    
  # Vectorized translation for speed
  dna <- DNAStringSet(orf_proteins$orf_dna_sequence)
  proteins <- translate(dna, if.fuzzy.codon = "solve")
  
  orf_proteins %<>%
    mutate(orf_protein_sequence = as.character(proteins)) %>%
    filter(orf_protein_sequence != "") %>%
    select(isoform_id, orf_protein_sequence)
  
  
  # Group transcripts by identical protein sequences
  orf_groups = orf_proteins %>%
    group_by(orf_protein_sequence) %>%
    arrange(isoform_id) %>%
    mutate(
      orf_isoform_id = paste(isoform_id, collapse = "|"),
      orf_base_id = isoform_id[1]  # First transcript as representative
    ) %>% 
    ungroup()
  
  return(orf_groups)
}






```


Compare SQANTI and SQANTI protein output
```{r}
trans = read_tsv("/home/cwp5au/sheynkman/projects/LRP_AML/03_filter_sqanti/filtered_AML_classification.txt")
protein = read_tsv("/home/cwp5au/sheynkman/projects/LRP_AML/allAMLvNBM/09_sqanti_protein/AML.sqanti_protein_classification.tsv")

trans %>% filter(associated_gene == "ENSG00000116954.8")
protein %>% filter(tx_gene == "ENSG00000116954.8")

```

Annotate cds and utr based on ORF
```{r}
# Sample GTF
sample_gtf_path = "~/megan/github/LRP2_lite/results/sqanti/sample_corrected_filtered.gtf"
orfs_file = "results/orf_calling/sample_best_cpat_orfs_mapped.tsv"

gtf = import(sample_gtf_path)

#!/usr/bin/env Rscript

#' GTF CDS Feature Addition Script
#' 
#' Creates GTF file that includes ORF regions as CDS features
#' Input: PacBio GTF, ORF calls, pb_gene mapping
#' Output: PacBio GTF with added CDS features (ORFs)

# =============================================================================
# Load required libraries
# =============================================================================

suppressPackageStartupMessages({
  library(tidyverse)
  library(rtracklayer)
  library(magrittr)
})

# =============================================================================
# Get environment variables and check required files
# =============================================================================

basename         <- Sys.getenv("OUTPUT_BASE_NAME")
sqanti_dir       <- file.path(Sys.getenv("OUTPUT_DIR"), "sqanti") 
cpat_dir         <- file.path(Sys.getenv("OUTPUT_DIR"), "orf_calling") 

sample_gtf_path  <- file.path(sqanti_dir, paste0(basename, "_corrected_filtered.gtf"))
orfs_file        <- file.path(cpat_dir, paste0(basename, "_best_cpat_orfs_mapped.tsv"))

stopifnot("Sample GTF file not found" = file.exists(sample_gtf_path))
stopifnot("Best ORFs file not found" = file.exists(orfs_file))

# =============================================================================
# Helper functions
# =============================================================================

#' Find which exon block contains an ORF coordinate
#' @param orf_coord Position in transcript coordinates
#' @param cumulative_lens Cumulative exon lengths
#' @return List with block index and delta from end
get_first_block_index <- function(orf_coord, cumulative_lens) {
  for (i in seq_along(cumulative_lens)) {
    if (orf_coord <= cumulative_lens[i]) {
      delta <- cumulative_lens[i] - orf_coord
      return(list(index = i, delta = delta))
    }
  }
  return(NULL)
}

#' Trim coordinates to ORF range (positive strand)
#' @param i1 Start block index
#' @param delta1 Distance from end of start block
#' @param i2 End block index
#' @param delta2 Distance from end of end block
#' @param coords_df Data frame with start/end coordinates
#' @return Trimmed coordinates
trim_coords_positive <- function(i1, delta1, i2, delta2, coords_df) {
  orf_coords <- coords_df %>%
    slice(i1:i2) %>%
    mutate(
      start = if_else(row_number() == 1, end - delta1, start),
      end = if_else(row_number() == n(), end - delta2, end)
    )
  return(orf_coords)
}

#' Trim coordinates to ORF range (negative strand)
#' @param i1 Start block index
#' @param delta1 Distance from end of start block
#' @param i2 End block index
#' @param delta2 Distance from end of end block
#' @param coords_df Data frame with start/end coordinates
#' @return Trimmed coordinates
trim_coords_negative <- function(i1, delta1, i2, delta2, coords_df) {
  orf_coords <- coords_df %>%
    slice(i1:i2) %>%
    mutate(
      start = if_else(row_number() == n(), start + delta2, start),
      end = if_else(row_number() == 1, start + delta1, end)
    )
  return(orf_coords)
}

#' Generate GTF with CDS features
#' @param sample_gtf GTF file path
#' @param agg_orfs_file Aggregated ORFs file path
#' @param refined_orfs_file Refined ORFs file path
#' @param pb_gene_file PacBio to gene mapping file path
#' @param output_cds Output GTF file path
#' @return NULL (writes output file)
make_pacbio_cds_gtf <- function(sample_gtf, agg_orfs_file, refined_orfs_file, 
                                pb_gene_file, output_cds) {
  
  # === STEP 1: Load GTF and extract exons ===
  message("\n--- STEP 1: Reading GTF file ---")
  
  gtf <- import(sample_gtf) %>%
    as.data.frame() %>%
    filter(type == "exon") %>%
    select(seqnames, start, end, strand, transcript_id) %>%
    rename(chr = seqnames, acc = transcript_id)
  
  message(paste0("Loaded ", nrow(gtf), " exon features"))
  
  # === STEP 2: Load ORF aggregation data ===
  message("\n--- STEP 2: Reading ORF aggregation data ---")
  
  agg_orfs <- read_tsv(agg_orfs_file, show_col_types = FALSE)
  
  # Create mappings from base accession
  base_acc_to_cpm <- setNames(agg_orfs$CPM, agg_orfs$base_acc)
  base_acc_to_transcripts <- setNames(agg_orfs$pb_accs, agg_orfs$base_acc)
  
  # Filter to representative transcripts only
  repr_pbs <- agg_orfs$base_acc
  gtf %<>% filter(acc %in% repr_pbs)
  
  message(paste0("Filtered to ", n_distinct(gtf$acc), " representative transcripts"))
  
  # === STEP 3: Organize coordinates by transcript ===
  message("\n--- STEP 3: Processing transcript coordinates ---")
  
  pbs <- gtf %>%
    group_by(acc) %>%
    arrange(acc, if_else(strand == "+", start, -start)) %>%
    mutate(
      block_len = end - start + 1,
      cum_block_len = cumsum(block_len)
    ) %>%
    summarise(
      chr = first(chr),
      strand = first(strand),
      coords = list(tibble(start = start, end = end)),
      block_lens = list(block_len),
      cum_block_lens = list(cum_block_len),
      .groups = "drop"
    )
  
  # === STEP 4: Load ORF ranges and gene mappings ===
  message("\n--- STEP 4: Loading ORF coordinates and gene mappings ---")
  
  ranges <- read_tsv(refined_orfs_file, show_col_types = FALSE) %>%
    select(pb_acc, orf_start, orf_end)
  
  pb_to_gene <- read_tsv(pb_gene_file, show_col_types = FALSE)
  pb_gene_map <- setNames(pb_to_gene$gene, pb_to_gene$pb_acc)
  
  message(paste0("Processing ", nrow(ranges), " ORF calls"))
  
  # === STEP 5: Map ORFs to genomic coordinates and write output ===
  message("\n--- STEP 5: Mapping ORFs and writing output GTF ---")
  
  output_lines <- list()
  n_processed <- 0
  n_skipped <- 0
  
  for (i in seq_len(nrow(ranges))) {
    acc <- ranges$pb_acc[i]
    orf_start <- ranges$orf_start[i]
    orf_end <- ranges$orf_end[i]
    
    # Check if transcript exists and has gene mapping
    if (!(acc %in% pbs$acc)) {
      n_skipped <- n_skipped + 1
      next
    }
    
    gene <- pb_gene_map[acc]
    if (is.na(gene) || gene == "-") {
      n_skipped <- n_skipped + 1
      next
    }
    
    # Get transcript information
    pb_info <- pbs %>% filter(acc == !!acc)
    chr <- pb_info$chr
    strand <- pb_info$strand
    coords <- pb_info$coords[[1]]
    cum_lens <- pb_info$cum_block_lens[[1]]
    
    # Get CPM and collapsed transcript names
    cpm <- base_acc_to_cpm[acc]
    if (is.na(cpm)) cpm <- 0
    
    all_transcripts <- base_acc_to_transcripts[acc]
    if (is.na(all_transcripts)) all_transcripts <- acc
    all_transcripts <- str_replace_all(all_transcripts, "\\|", ",")
    
    # Truncate long names
    max_name_length <- 50
    if (nchar(all_transcripts) > max_name_length) {
      transcript_list <- str_split(all_transcripts, ",")[[1]]
      all_transcripts <- paste0(transcript_list[1], "_plus_", 
                                length(transcript_list) - 1, "_others")
    }
    
    # Find ORF blocks
    idx1 <- get_first_block_index(orf_start, cum_lens)
    idx2 <- get_first_block_index(orf_end, cum_lens)
    
    if (is.null(idx1) || is.null(idx2)) {
      n_skipped <- n_skipped + 1
      next
    }
    
    # Trim coordinates to ORF range
    if (strand == "+") {
      orf_coords <- trim_coords_positive(idx1$index, idx1$delta, 
                                         idx2$index, idx2$delta, coords)
    } else {
      orf_coords <- trim_coords_negative(idx1$index, idx1$delta, 
                                         idx2$index, idx2$delta, coords)
    }
    
    # Create transcript ID with gene, transcripts, and CPM
    acc_w_gene_w_cpm <- paste0(gene, "|", all_transcripts, "|", cpm)
    attributes <- paste0('gene_id "', gene, '"; transcript_id "', 
                        acc_w_gene_w_cpm, '";')
    
    # Write exon lines
    for (j in seq_len(nrow(coords))) {
      line <- paste(chr, "hg38", "exon", coords$start[j], coords$end[j], 
                   ".", strand, ".", attributes, sep = "\t")
      output_lines <- c(output_lines, line)
    }
    
    # Write CDS lines
    for (j in seq_len(nrow(orf_coords))) {
      line <- paste(chr, "hg38", "CDS", orf_coords$start[j], orf_coords$end[j], 
                   ".", strand, ".", attributes, sep = "\t")
      output_lines <- c(output_lines, line)
    }
    
    n_processed <- n_processed + 1
  }
  
  # Write output file
  writeLines(unlist(output_lines), output_cds)
  
  # === Summary ===
  message("\n=== PROCESSING COMPLETE ===")
  message(paste0("Successfully processed: ", n_processed, " transcripts"))
  message(paste0("Skipped (no gene mapping or not found): ", n_skipped, " transcripts"))
  message(paste0("Output written to: ", output_cds))
  message(paste0("Total GTF lines written: ", length(output_lines)))
}

# =============================================================================
# Main execution
# =============================================================================

message("\n=== GTF CDS FEATURE ADDITION ===")
message(paste0("Sample GTF: ", sample_gtf_path))
message(paste0("Aggregated ORFs: ", agg_orfs_file))
message(paste0("Refined ORFs: ", refined_orfs_file))
message(paste0("PacBio-Gene mapping: ", pb_gene_file))
message(paste0("Output: ", output_cds_path))

make_pacbio_cds_gtf(
  sample_gtf_path,
  agg_orfs_file,
  refined_orfs_file,
  pb_gene_file,
  output_cds_path
)

message("\nScript completed successfully!")

```


```{r}
all_cds_exons %>% filter(new_start > new_end) # PB.10040.43

#keep = best_orfs
best_orfs = keep
best_orfs %>% filter(strand == "+")


get_cds_coords <- function(best_orfs, sample_exons) {
  
  # Join mapped ORFs with their exons- select all CDS exons
  cds_all = best_orfs %>%
    left_join(sample_exons, by = c("isoform_id", "seqnames", "strand")) %>%
    filter(exon_number >= orf_start_exon & exon_number <= orf_end_exon) %>% # get all CDS exons
    group_by(isoform_id) %>%
    arrange(isoform_id, exon_number) %>%
    mutate(
      is_first_orf_exon = (exon_number == min(exon_number)),
      is_last_orf_exon = (exon_number == max(exon_number))
    ) %>%
    ungroup()
  
  # Trim first and last cds exons based on precise cds start and end
  cds_trimmed = cds_all %>%
    mutate(
      new_start = case_when(
        strand == "+" & is_first_orf_exon ~ start + orf_start_offset - 1,
        strand == "-" & is_last_orf_exon ~ end - orf_end_offset + 1,
        TRUE ~ start
      ),
      new_end = case_when(
        strand == "+" & is_last_orf_exon ~ start + orf_end_offset - 1,
        strand == "-" & is_first_orf_exon ~ end - orf_start_offset + 1,
        TRUE ~ end
      )
    ) %>%
    select(ID, isoform_id, orf_isoform_id, seqnames, exon_number, new_start, new_end, strand)
  
  return(cds_trimmed)
}

cds_trimmed %>% filter(new_start == new_end) %>% left_join(gene_mapping)
all_cds_exons = get_cds_coords(best_orfs, sample_exons)
all_cds_exons

# PB.27.11 (-)
gencode_gtf %>% filter(type == "CDS", transcript_name == "NOC2L-201")

# PB.30.2 (+)
gencode_gtf %>% filter(type == "CDS", transcript_name == "ISG15-203")
gencode_gtf %>% filter(transcript_name == "ISG15-203")

# PB.10453.20
mapped_orfs_classified %>% filter(isoform_id == "PB.10453.48")
cds_all %>% filter(isoform_id == "PB.10453.48")
gencode_gtf %>% filter(type == "CDS", transcript_name == "SELENOT-206")


# TNFAIP8-201, PB.15079.14, PB.15079.29, PB.15079.30
gencode_gtf %>% filter(type == "CDS", transcript_name == "TNFAIP8-201")
cds_trimmed %>% filter(isoform_id == "PB.15079.14")
```


Create a single function for CDS mapping- replace
```{r}
sample_gtf_path = "~/megan/github/LRP2_lite/results/sqanti/sample_corrected_filtered.gtf"
sample_gtf  = import(sample_gtf_path) %>% as.data.frame()

gencode_gtf = import("/project/sheynkman/external_data/GENCODE_v47/gencode.v47.annotation.gtf") %>% as.data.frame()

cpat_results_file = "/home/cwp5au/megan/github/LRP2_lite/results/orf_calling/sample_cpat.ORF_prob.tsv"
orf_coords = read_tsv(cpat_results_file, show_col_types = FALSE) %>%
  separate(ID, into = c("isoform_id", "orf_rank"), sep = "_ORF_", remove = FALSE)

map_orfs_to_genome <- function(orf_coords, sample_gtf, gencode_gtf, full_fasta) {
  
  # Extract exon information from sample GTF
  sample_exons = sample_gtf %>%
    filter(type == "exon") %>%
    select(isoform_id = transcript_id, seqnames, start, end, strand) %>%
    mutate(exon_length = end - start + 1) %>%
    arrange(isoform_id, if_else(strand == "+", start, -start))
  
  # Calculate cumulative exon positions for coordinate mapping
  sample_exons %<>%
    group_by(isoform_id) %>%
    mutate(
      exon_number = row_number(),
      cumulative_length = cumsum(exon_length), 
      prior_length = lag(cumulative_length, default = 0)
    ) %>%
    ungroup()
  
  # Count the number of upstream ORFs
  orf_coords %<>% left_join(full_fasta, by = "isoform_id") %>%
    mutate(upstream_atgs = ifelse(!is.na(full_sequence), 
                                  str_count(substr(full_sequence, 1, ORF_start - 1), "ATG"),
                                  Inf)
    ) %>%
    select(-full_sequence)
  
  # Map ORF Start coordinates to genomic positions- strandedness here is tricky
  mapped_starts = orf_coords %>%
    left_join(sample_exons, by = c("isoform_id")) %>%
    filter(ORF_start > prior_length & ORF_start <= cumulative_length) %>% # which exon has the ORF start?
    mutate(orf_genomic_start = ifelse(strand == "+",
                                  start + (ORF_start - prior_length) - 1,
                                  end - (ORF_start - prior_length) + 1),
           orf_start_exon   = exon_number,
           orf_start_offset = ORF_start - prior_length  # position within this exon
    ) %>%
    select(-start, -end, -exon_length, -exon_number, -cumulative_length, -prior_length)
  
  # Map ORF End coordinates to genomic positions
  mapped_complete = mapped_starts %>%
    left_join(sample_exons, by = c("isoform_id", "seqnames", "strand")) %>%
    filter(ORF_end > prior_length & ORF_end <= cumulative_length) %>% # which exon has the ORF end?
    mutate(
      orf_genomic_end = if_else(strand == "+",
                                 start + (ORF_end - prior_length) - 1,
                                 end - (ORF_end - prior_length) + 1),
      orf_end_exon   = exon_number,
      orf_end_offset = ORF_end - prior_length  # position within this exon
    ) %>%
    select(-start, -end, -exon_length, -exon_number, -cumulative_length, -prior_length)
  
  # Check for GENCODE start codon matches- looking for any match here
  gencode_start_codons = gencode_gtf %>%
    filter(type == "start_codon") %>% 
    mutate(gencode_pos = ifelse(strand == "+", start, end)) %>%
    select(seqnames, strand, gencode_pos) %>%
    distinct() %>%
    mutate(gencode_match = TRUE)
  
  mapped_orfs = mapped_complete %>% 
    left_join(gencode_start_codons, by = c("seqnames", "strand", "orf_genomic_start" = "gencode_pos")) %>%
    mutate(gencode_match = ifelse(is.na(gencode_match), FALSE, gencode_match))
  
  return(mapped_orfs)
}


```

Next function to trim CDS to add to gtf- do this after filtering for best ORF? 
- yes because otherwise, will have multiple ORFs per transcript in the gtf
- and this is how it was in the OG pipeline
```{r}
test = best_orfs

#' Get CDS coordinates by trimming exons to ORF boundaries
#' Uses the exon mapping information to extract only CDS regions
#' @param best_orfs Each row has an ORF with mapping info
#' @param sample_gtf Info on all exons
#' @return Data frame with CDS start/end coordinates
get_cds_coords <- function(best_orfs, sample_exons) {
  
  # Join mapped ORFs with their exons- select all CDS exons
  cds_all = best_orfs %>%
    left_join(sample_exons, by = c("isoform_id", "seqnames", "strand")) %>%
    filter(exon_number >= orf_start_exon & exon_number <= orf_end_exon) %>% # get all CDS exons
    group_by(isoform_id) %>%
    arrange(isoform_id, exon_number) %>%
    mutate(
      is_first_orf_exon = (exon_number == min(exon_number)),
      is_last_orf_exon = (exon_number == max(exon_number))
    ) %>%
    ungroup()
  
  # Trim first and last cds exons based on precise cds start and end
  cds_trimmed = cds_all %>%
    mutate(
      new_start = case_when(
        strand == "+" & is_first_orf_exon ~ start + orf_start_offset - 1,
        strand == "-" & is_last_orf_exon ~ start + orf_end_offset - 1,
        TRUE ~ start
      ),
      new_end = case_when(
        strand == "+" & is_last_orf_exon ~ start + orf_end_offset - 1,
        strand == "-" & is_first_orf_exon ~ end - orf_start_offset + 1,
        TRUE ~ end
      )
    ) %>%
    select(ID, isoform_id, seqnames, exon_number, new_start, new_end, strand)
  
  return(cds_trimmed)
}

get_cds_coords(best_orfs, sample_exons)


gencode_gtf %>% group_by(type) %>% tally()
sample_gtf %>% group_by(type) %>% tally()

```

Make the gtf- dont think this needs to be a function
- can we output a gtf of all transcripts and then one that is aggregated at the ORF level with ORF id?


```{r}
basename = "sample"
gene_mapping
all_cds_exons
sample_gtf

mapped_orfs_classified %>% filter(isoform_id == "PB.100.1")


sample_gtf_path = "~/megan/github/LRP2_lite/results/sqanti/sample_corrected_filtered.gtf"
gencode_gtf = import("/project/sheynkman/external_data/GENCODE_v47/gencode.v47.annotation.gtf") %>% as.data.frame()

sample_gtf  = import(sample_gtf_path) %>% as.data.frame()

cpat_dir = "/home/cwp5au/megan/github/LRP2_lite/results/orf_calling"
cpat_fasta_file = "/home/cwp5au/megan/github/LRP2_lite/results/orf_calling/sample_cpat.ORF_seqs.fa"
cpat_results_file = "/home/cwp5au/megan/github/LRP2_lite/results/orf_calling/sample_cpat.ORF_prob.tsv"
sample_full_fasta_path = "/home/cwp5au/megan/github/LRP2_lite/results/sqanti/sample_corrected_filtered.fasta"
classification_file = "/home/cwp5au/megan/github/LRP2_lite/results/sqanti/sample_classification_filtered.txt"


write_gtf_with_cds <- function(sample_gtf, all_cds_exons, gene_mapping) {
  
  # selecting type individually in case custom gtfs have more type columns- want to reset attributes column
  sample_gtf %<>% 
    select(isoform_id = transcript_id, everything()) %>%
    left_join(gene_mapping %>% select(isoform_id, gencode_gene_id = gene_id, gene_name), by = "isoform_id")
  
  # Create transcript lines (one per transcript)
  transcript_lines = sample_gtf %>%
    filter(type == "transcript") %>%
    select(seqnames, type, start, end, strand, source, isoform_id, gencode_gene_id, gene_name)
  
  # Create exon lines
  exon_lines = sample_gtf %>%
    filter(type == "exon") %>%
    select(seqnames, type, start, end, strand, source, isoform_id, gencode_gene_id, gene_name)
  
  # Create CDS lines
  cds_lines = all_cds_exons %>%
    left_join(gene_mapping %>% select(isoform_id, gencode_gene_id = gene_id, gene_name), by = "isoform_id") %>%
    mutate(type = "CDS", source = sample_gtf$source[1]) %>%
    select(seqnames, type, start = new_start, end = new_end, strand, source, isoform_id, gencode_gene_id, gene_name)
  
  # combine types and add attribute column
  updated_gtf = bind_rows(transcript_lines, exon_lines, cds_lines) %>%
    arrange(isoform_id, start) %>%
    mutate(
      attributes = paste0('gene_id "', gencode_gene_id, '"; transcript_id "', isoform_id, '"; gene_name "', gene_name, '";'),
      score = ".", 
      frame = "."
    ) %>%
    select(seqnames, source, type, start, end, score, strand, frame, attributes)

  message(paste0("Combined ", nrow(transcript_lines), " transcript lines, ",
                 nrow(exon_lines), " exon lines, and ", 
                 nrow(cds_lines), " CDS lines"))
  
  return(updated_gtf)
}




```

Compare SQANTI transcript and protein classification output for AML samples
```{r}
transcript = read_tsv("/home/cwp5au/sheynkman/projects/LRP_AML/03_filter_sqanti/filtered_AML_classification.txt")
transcript %>% group_by(structural_category) %>% tally()
transcript %>% group_by(subcategory) %>% tally()
  
protein = read_tsv("/home/cwp5au/sheynkman/projects/LRP_AML/allAMLvNBM/09_sqanti_protein/AML.sqanti_protein_classification.tsv")

transcript %>% filter(str_detect(isoform, "PB\\.1000\\.")) %>% arrange(isoform)
protein %>% filter(str_detect(pb, "PB\\.1000\\.")) %>% arrange(pb)

transcript %>% filter(isoform == "PB.1000.25")
protein %>% filter(pb == "PB.1000.25")


protein_class = protein %>% select(isoform = pb, tx_cat, pr_splice_cat, tx_subcat, pr_splice_subcat, tx_gene, pr_gene, tx_transcripts, pr_transcripts)
transcript_class = transcript %>% select(isoform, structural_category, associated_gene, associated_transcript, subcategory)

df = protein_class %>% left_join(transcript_class)

# original transcript calls versus current
all(df$tx_cat == df$structural_category)
df %>% filter(tx_cat != structural_category) # fusions, 89 dont match

all(df$tx_subcat == df$subcategory)
df %>% filter(tx_subcat != subcategory) # 14,861


# transcript versus protein
all(df$tx_cat == df$pr_splice_cat)
df %>% filter(tx_cat != pr_splice_cat)

all(df$tx_subcat == df$pr_splice_subcat)
df %>% filter(tx_subcat != pr_splice_subcat)

# match to gencode
all(df$tx_gene == df$associated_gene)
df %>% filter(tx_gene != associated_gene) # 104, includes a lot of fusions

# look at a fusion
transcript %>% filter(isoform == "PB.34191.15")
protein %>% filter(pb == "PB.34191.15")


protein %>% filter(is_nmd == TRUE)
transcript %>% group_by(subcategory) %>% tally() # 97,481
protein %>% group_by(tx_subcat) %>% tally() # 32,029
protein %>% group_by(pr_splice_subcat) %>% tally()


protein %>% filter(pb == "PB.1000.25")

protein %>% filter(tx_cat == "full-splice_match") %>% group_by(pr_splice_cat) %>% tally()
protein %>% filter(tx_cat == "incomplete-splice_match") %>% group_by(pr_splice_cat) %>% tally()
protein %>% filter(tx_cat == "novel_in_catalog") %>% group_by(pr_splice_cat) %>% tally()
protein %>% filter(tx_cat == "novel_not_in_catalog") %>% group_by(pr_splice_cat) %>% tally()


```


5'UTR code and protein classification
```{r}
# AML test
gencode_gtf_path = "/project/sheynkman/external_data/GENCODE_v47/gencode.v47.annotation.gtf"
sample_cds_gtf_path = "/project/sheynkman/users/megan/github/LRP2_lite/results/orf_calling/sample_corrected_filtered_CDS.gtf"
protein_sqanti_path = "/project/sheynkman/users/megan/github/LRP2_lite/results/protein_sqanti/sample.sqanti_protein_classification.tsv"
basename = "sample"  

# jurkat test
gencode_gtf_path = "~/datasets/jurkat_zenodo5234651/gencode.v35.annotation.chr22.gtf"
sample_cds_gtf_path = "~/datasets/jurkat_zenodo5234651/jurkat_chr22_with_cds.gtf"
protein_sqanti_path = "/project/sheynkman/users/megan/github/LRP2_lite/results/protein_sqanti/jurkat_chr22.sqanti_protein_classification.tsv"
basename = "jurkat" 

#!/usr/bin/env Rscript

#' Protein Classification Script
#' 
#' Part 1: Classifies Long read derived 5' UTRs as either:
#'   - "subset" (N-terminal truncation, using internal start codon)
#'   - "unique" (novel start site, extending into new sequence)
#'   
#' Part 2: Protein Classification
#' Integrates SQANTI-protein output to classify CDS structures:
#'   - pFSM (Full Protein Match): CDS junctions exactly match reference
#'   - pISM (Incomplete Protein Match): CDS junctions are subset of reference
#'   - pNIC (Novel In Catalog): Novel combination of known N/splice/C
#'   - pNNC (Novel Not in Catalog): At least one novel N/splice/C

# =============================================================================
# Load libraries
# =============================================================================

suppressPackageStartupMessages({
  library(GenomicRanges)
  library(txdbmaker)
  library(Biostrings)
  library(tidyverse)
  library(magrittr)
  library(rtracklayer)
})

# =============================================================================
# Get environment variables and check required files
# =============================================================================

basename           <- Sys.getenv("OUTPUT_BASE_NAME")
cpat_dir           <- file.path(Sys.getenv("OUTPUT_DIR"), "orf_calling") 
protein_sqanti_dir <- file.path(Sys.getenv("OUTPUT_DIR"), "protein_sqanti") 
gencode_gtf_path   <- Sys.getenv("GENCODE_GTF_FILE")

sample_cds_gtf_path   <- file.path(cpat_dir, paste0(basename, "_corrected_filtered_CDS.gtf"))
protein_sqanti_path   <- file.path(protein_sqanti_dir, paste0(basename, ".sqanti_protein_classification.tsv"))

stopifnot("GENCODE gtf file not found" = file.exists(gencode_gtf_path))
stopifnot("Sample gtf file not found" = file.exists(sample_cds_gtf_path))
stopifnot("Protein classification file not found" = file.exists(protein_sqanti_path))

# =============================================================================
# Helper Functions
# =============================================================================

#' Classify multiexonic 5' UTR based on junction chain
#' @param tss Integer: TSS coordinate
#' @param strand Character: '+' or '-'
#' @param junc_chain Character: junction chain string
#' @param gc_chains Character vector: GENCODE junction chains for this gene
#' @return Character: "perfect_subset", "known_protruding", or "novel"
classify_multiexonic_utr = function(tss, strand, junc_chain, gc_chains) {
  if (is.na(tss) || junc_chain == "" || length(gc_chains) == 0) return("novel")

  # Split the concatenated chains
  gc_chains = str_split(gc_chains, "\\|\\|\\|")[[1]]
  
  status = "novel"
  
  for (gc_chain in gc_chains) {
    if (!str_detect(gc_chain, fixed(junc_chain))) next
    
    # Junction chain matches - check if TSS protrudes beyond reference
    if (strand == "+") {
      prefix  = str_split(gc_chain, fixed(junc_chain))[[1]][1]
      coords  = str_split(prefix, "_|-")[[1]][1]
      gc_5end = as.integer(coords[length(coords)])
      
      if (is.na(gc_5end)) next
      
      if ((tss + 9) < gc_5end) {
        status = "known_protruding"
      } else {
        return("perfect_subset")
      }
    } else {
      suffix  = str_split(gc_chain, fixed(junc_chain))[[1]][2]
      coords  = str_split(suffix, "_|-")[[1]][1]
      gc_5end = as.integer(coords[2])
      
      if (is.na(gc_5end)) next
      
      if ((tss - 9) > gc_5end) {
        status = "known_protruding"
      } else {
        return("perfect_subset")
      }
    }
  }
  
  return(status)
}

# =============================================================================
# STEP 1: Process GENCODE reference
# =============================================================================

message("\n--- STEP 1: Processing GENCODE reference ---")

# Load GENCODE GTF
gencode_gtf = import(gencode_gtf_path) %>% as.data.frame()

# Identify protein-coding transcripts (have CDS)
pc_transcripts = gencode_gtf %>%
  filter(type == "CDS") %>%
  pull(transcript_name) %>%
  unique()

# Extract exons and create GRanges for overlap checking
gencode_exons = gencode_gtf %>%
  filter(type == "exon", transcript_name %in% pc_transcripts) %>%
  select(seqnames, start, end, gene_id, transcript_name, strand)

gencode_gr = gencode_exons %>%
  select(seqnames, start, end, strand) %>%
  makeGRangesFromDataFrame()

gencode_gr = GenomicRanges::reduce(gencode_gr) # Reduces overlapping exons into merged genomic intervals

# Create exon chain strings per gene for junction matching (could potentially be tweaked to mimic collapse logic)
gencode_chains = gencode_exons %>%
  arrange(gene_id, transcript_name, start) %>%
  group_by(gene_id, transcript_name) %>%
  summarise(
    exon_chain = paste(paste0(start, "-", end), collapse = "_"),
    .groups = "drop_last"  # Keep gene grouping!
  ) %>%
  summarise(chains = paste(exon_chain, collapse = "|||"), .groups = "drop") 

# =============================================================================
# STEP 2: Process long read transcripts
# =============================================================================

message("\n--- STEP 2: Processing Sample transcripts ---")

sample_gtf = import(sample_cds_gtf_path) %>% as.data.frame()

# Create TxDb and extract 5' UTRs directly
txdb      = makeTxDbFromGFF(sample_cds_gtf_path, format = "gtf")
utr_by_tx = fiveUTRsByTranscript(txdb, use.names = TRUE)

# Get transcript metadata (gene name, chromosome, strand)
tx_metadata = sample_gtf %>%
  filter(type == "transcript") %>%
  select(transcript_id, gene_id, seqnames, strand)

# Get CDS Nterm and TSS info 
sample_cds = sample_gtf %>%
  filter(type == "CDS") %>%
  group_by(transcript_id) %>%
  summarise(
    nterm = ifelse(first(strand) == "+", min(start), max(end)),
    .groups = "drop"
  )

sample_exons = sample_gtf %>%
  filter(type == "exon") %>%
  group_by(transcript_id) %>%
  summarise(
    tss = ifelse(first(strand) == "+", min(start), max(end)),
    .groups = "drop"
  )

# Convert 5' UTR GRangesList to tibble with stats
utr_stats = tibble(
  transcript_id    = names(utr_by_tx),
  num_5utr_exons   = lengths(utr_by_tx),
  utr_length       = sum(width(utr_by_tx))
)

# Create chain strings for 5' UTRs
utr_chains = tibble(
  transcript_id = names(utr_by_tx),
  utr_chain = sapply(utr_by_tx, function(gr) {
    if (length(gr) == 0) return("")
    gr = sort(gr)
    paste(paste0(start(gr), "-", end(gr)), collapse = "_")
  })
) %>%
  mutate(
    junc_chain = case_when(
      utr_chain == "" ~ "",
      !str_detect(utr_chain, "_") ~ "",
      TRUE ~ {
        coords = str_extract_all(utr_chain, "\\d+")
        sapply(coords, function(x) {
          if (length(x) <= 2) return("")
          junctions = as.numeric(x[2:(length(x)-1)])
          paste(junctions, collapse = "_")
        })
      }
    )
  )

# Combine all info
utr_info = tx_metadata %>%
  left_join(sample_exons, by = "transcript_id") %>%
  left_join(sample_cds, by = "transcript_id") %>%
  left_join(utr_stats, by = "transcript_id") %>%
  left_join(utr_chains, by = "transcript_id") %>%
  mutate(
    # Handle transcripts with no UTR
    num_5utr_exons = replace_na(num_5utr_exons, 0),
    utr_length = replace_na(utr_length, 0),
    utr_exon_status = case_when(
      num_5utr_exons == 0 ~ "no_utr",
      num_5utr_exons == 1 ~ "monoexonic",
      TRUE ~ "multiexonic"
    )
  )

# =============================================================================
# STEP 3: Classify 5' UTRs
# =============================================================================

message("\n--- STEP 3: Classifying 5' UTRs ---")

# Check TSS overlap with GENCODE using GRanges 
tss_gr = GRanges(
  seqnames = utr_info$seqnames,
  ranges   = IRanges(start = utr_info$tss, end = utr_info$tss),
  strand   = utr_info$strand
)

overlaps = findOverlaps(tss_gr, gencode_gr)
utr_info$tss_in_gc_exons = seq_len(nrow(utr_info)) %in% queryHits(overlaps)

# For monoexonic: calculate distance to nearest GENCODE exon 
distances       = rep(NA_real_, nrow(utr_info))
non_overlapping = which(!utr_info$tss_in_gc_exons & utr_info$utr_exon_status == "monoexonic")

if (length(non_overlapping) > 0) {
  nearest                    = distanceToNearest(tss_gr[non_overlapping], gencode_gr)
  distances[non_overlapping] = mcols(nearest)$distance
}

utr_info$distance_to_gc = distances

# Join GENCODE chains for multiexonic comparison
utr_info = utr_info %>%
  left_join(gencode_chains, by = "gene_id")

# Classify monoexonic UTRs
utr_info = utr_info %>%
  mutate(
    mono_class = case_when(
      utr_exon_status != "monoexonic" ~ NA_character_,
      tss_in_gc_exons ~ "within",
      distance_to_gc >= 10 ~ "protruding",
      TRUE ~ "within"
    )
  )

# Classify multiexonic UTRs (needs rowwise for junction matching)
multi_idx = which(utr_info$utr_exon_status == "multiexonic")

if (length(multi_idx) > 0) {
  multi_results = utr_info[multi_idx, ] %>%
    rowwise() %>%
    mutate(
      multi_class = classify_multiexonic_utr(tss, strand, junc_chain, chains)
    ) %>%
    ungroup() %>%
    pull(multi_class)
  
  utr_info$multi_class            = NA_character_
  utr_info$multi_class[multi_idx] = multi_results
}

# Final classification 
utr_results = utr_info %>%
  mutate(
    junc_cat = case_when(
      is.na(nterm) ~ "no_orf",
      utr_exon_status == "no_utr" ~ "no_utr",
      utr_exon_status == "monoexonic" ~ mono_class,
      utr_exon_status == "multiexonic" ~ multi_class,
      TRUE ~ "unknown"
    ),
    
    # Map to final categories
    utr_cat = case_when(
      junc_cat %in% c("no_orf", "no_utr", "unknown") ~ junc_cat,
      junc_cat %in% c("within", "perfect_subset") ~ "subset",
      junc_cat %in% c("protruding", "known_protruding", "novel") ~ "unique",
      TRUE ~ "unknown"
    )
  ) %>%
  select(isoform_id = transcript_id, num_5utr_exons, utr_exon_status, tss_in_gc_exons, junc_cat, utr_cat)

# Merge with SQANTI protein classification
sqanti_class = read_tsv(protein_sqanti_path, show_col_types = FALSE)

# Adjust C term differences since CPAT includes stop codon in coords and gencode does not
sqanti_class %<>% mutate(
  pr_cterm_diff = ifelse(!is.na(pr_cterm_diff), pr_cterm_diff + 3, pr_cterm_diff),
  pr_cterm_gene_diff = ifelse(!is.na(pr_cterm_gene_diff), pr_cterm_gene_diff - 3, pr_cterm_gene_diff),
  pr_chang = ifelse(!is.na(pr_chang), pr_chang - 3, pr_chang))

utr_output = sqanti_class %>%
  full_join(utr_results, by = "isoform_id")

# write out intermediate file for troubleshooting
#write_tsv(utr_output, paste0("../results/protein_sqanti/", basename, "_5utr_classification.tsv"))


# 5' UTR summary statistics
message("\n--- 5'UTR Classification complete! ---")
message(sprintf("Total transcripts: %d", nrow(utr_output)))
message(sprintf("  subset: %d", sum(utr_output$utr_cat == "subset", na.rm = TRUE)))
message(sprintf("  unique: %d", sum(utr_output$utr_cat == "unique", na.rm = TRUE)))
message(sprintf("  no_orf: %d", sum(utr_output$utr_cat == "no_orf", na.rm = TRUE)))
message(sprintf("  no_utr: %d", sum(utr_output$utr_cat == "no_utr", na.rm = TRUE)))


# =============================================================================
# STEP 4: Protein classification
# =============================================================================

message("\n--- STEP 4: Protein classification ---")

protein_classifications = utr_output %>%
  mutate(
    protein_classification = case_when(
      
      ## Compared to the specific best-matching reference isoform
      # pr_nterm_diff: difference in ORF start positions (ATG)
      # pr_cterm_diff: difference in ORF end positions (stop codon)
      
      ## Compared to the nearest position across ALL isoforms of that gene
      # pr_nterm_gene_diff: distance to nearest CDS start among all isoforms
      # pr_cterm_gene_diff: distance to nearest CDS end among all isoforms
      
      # -----------------------------------------------------------------------------------------------------------
      # SQANTI_Protein, Full Splice Match - CDS splice junctions match exactly (known_splice), but precise N and C term may differ 
      # -----------------------------------------------------------------------------------------------------------
      
      # Case 1: Both ATG and stop codon match reference isoform exactly
      # FPM = Full protein match
      pr_splice_cat == "full-splice_match" & pr_splice_subcat == "multi-exon" & 
        pr_nterm_diff == 0 & pr_cterm_diff == 0 ~ 
        "pFPM,known_nterm_known_splice_known_cterm",
      
      # Case 2: Novel combination - ATG and/or STOP differ from reference isoform, but both exist in other isoforms of the gene
      pr_splice_cat == "full-splice_match" & pr_splice_subcat == "multi-exon" & 
        (pr_nterm_diff != 0 | pr_cterm_diff != 0) & pr_nterm_gene_diff == 0 & pr_cterm_gene_diff == 0 ~ 
        "pNIC,combo_nterm_cterm",
      
      # Case 3: ATG matches the reference exactly, but STOP codon is novel
      pr_splice_cat == "full-splice_match" & pr_splice_subcat == "multi-exon" & 
        pr_nterm_diff == 0 & pr_cterm_diff != 0 ~ 
        "pNNC,known_nterm_known_splice_novel_cterm",
      
      # Case 4 subtypes: ATG doesn't match any isoform (Implicit: pr_nterm_gene_diff != 0), known STOP
      
      # 4a: Protein extends upstream - novel N-terminal amino acids
      pr_splice_cat == "full-splice_match" & pr_splice_subcat == "multi-exon" & 
        pr_nterm_diff != 0 & pr_cterm_diff == 0 & pr_nhang > 0 ~ 
        "pNNC,novel_nterm_known_splice_known_cterm",
      
      # 4b: ATG is downstream/within reference CDS, using internal start codon = N-terminal truncation
      # IPM = Incomplete Protein Match
      pr_splice_cat == "full-splice_match" & pr_splice_subcat == "multi-exon" & 
        pr_nterm_diff != 0 & pr_cterm_diff == 0 & pr_nhang <= 0 & utr_cat == "subset" ~ 
        "pIPM,nterm_truncation",
      
      # 4c: Novel TSS context - ATG position is novel but protein doesn't extend
      # Transcript has extended 5' UTR (novel TSS) creating novel regulatory context
      pr_splice_cat == "full-splice_match" & pr_splice_subcat == "multi-exon" & 
        pr_nterm_diff != 0 & pr_cterm_diff == 0 & pr_nhang <= 0 & utr_cat == "unique" ~ 
        "pNNC,novel_nterm_known_splice_known_cterm",
      
      # Case 5 subtypes: Both ATG and stop codon differ from reference isoform
      
      # 5a: ATG matches another isoform (known), stop is novel
      pr_splice_cat == "full-splice_match" & pr_splice_subcat == "multi-exon" & 
        pr_nterm_diff != 0 & pr_cterm_diff != 0 & pr_nterm_gene_diff == 0 ~ 
        "pNNC,known_nterm_known_splice_novel_cterm",
      
      # 5b: ATG doesn't match any isoform + internal start = truncation
      pr_splice_cat == "full-splice_match" & pr_splice_subcat == "multi-exon" & 
        pr_nterm_diff != 0 & pr_cterm_diff != 0 & pr_nterm_gene_diff != 0 & utr_cat == "subset" ~ 
        "pIPM,nterm_truncation",
      
      # 5c: Novel ATG (protein or TSS context) but end matches known isoform
      pr_splice_cat == "full-splice_match" & pr_splice_subcat == "multi-exon" & 
        pr_nterm_diff != 0 & pr_cterm_diff != 0 & pr_nterm_gene_diff != 0 & 
        utr_cat == "unique" & pr_cterm_gene_diff == 0 ~ 
        "pNNC,novel_nterm_known_splice_known_cterm",
      
      # 5d: Novel ATG and novel STOP
      pr_splice_cat == "full-splice_match" & pr_splice_subcat == "multi-exon" & 
        pr_nterm_diff != 0 & pr_cterm_diff != 0 & pr_nterm_gene_diff != 0 & 
        utr_cat == "unique" & pr_cterm_gene_diff != 0 ~ 
        "pNNC,novel_nterm_known_splice_novel_cterm",
      
      # FSM orphan catch-all
      pr_splice_cat == "full-splice_match" & pr_splice_subcat == "multi-exon" ~ 
        "orphan_fsm",
      
      # ---------------------------------------------------------------------------------------
      # SQANTI_Protein, Incomplete Splice Match (ISM) - CDS junctions are a subset of reference
      # (e.g., 5' fragment, 3' fragment, intron retention in CDS)
      # ---------------------------------------------------------------------------------------
      
      # Case 1: ATG matches reference isoform, stop codon is novel
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff == 0 & pr_cterm_diff != 0 & pr_cterm_gene_diff != 0 ~ 
        "pNNC,known_nterm_known_splice_novel_cterm",
      
      # Case 2: Novel combination - ATG matches reference isoform, stop codon matches another isoform
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff == 0 & pr_cterm_diff != 0 & pr_cterm_gene_diff == 0 ~ 
        "pNIC,combo_nterm_cterm",
      
      # Case 3: Novel combination - Both ATG and stop codon match (incomplete CDS splicing only)
      # Likely intron retention within CDS causing unique splicing pattern
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff == 0 & pr_cterm_diff == 0 ~ 
        "pNIC,known_nterm_combo_splice_known_cterm",
      
      # Case 4: Novel combination - ATG differs from reference but matches another isoform, stop codon matches
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff != 0 & pr_cterm_diff == 0 & pr_nterm_gene_diff == 0 ~ 
        "pNIC,combo_nterm_cterm",
      
      # Case 5 subtypes - ATG doesnt match any isoform (novel), stop codon matches
      
      # Case 5a: ATG is downstream (internal start codon) - N-terminal truncation
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff != 0 & pr_cterm_diff == 0 & pr_nterm_gene_diff != 0 & 
        pr_nhang <= 0 & utr_cat == "subset" ~ 
        "pIPM,nterm_truncation",
      
      # Case 5b: Novel TSS context - ATG position novel but protein doesn't extend
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff != 0 & pr_cterm_diff == 0 & pr_nterm_gene_diff != 0 & 
        pr_nhang <= 0 & utr_cat == "unique" ~ 
        "pNNC,novel_nterm_known_splice_known_cterm",
      
      # Case 5c: Protein extends upstream - novel N-terminal amino acids
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff != 0 & pr_cterm_diff == 0 & pr_nterm_gene_diff != 0 & pr_nhang > 0 ~ 
        "pNNC,novel_nterm_known_splice_known_cterm",
      
      # Case 6: Novel Combination - Both ATG and stop match other isoforms
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff != 0 & pr_cterm_diff != 0 & pr_nterm_gene_diff == 0 & pr_cterm_gene_diff == 0 ~ 
        "pNIC,combo_nterm_cterm",
      
      # Case 7: ATG matches another isoform, stop codon is novel
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff != 0 & pr_cterm_diff != 0 & pr_nterm_gene_diff == 0 & pr_cterm_gene_diff != 0 ~ 
        "pNNC,known_nterm_known_splice_novel_cterm",
      
      # Case 8 subtypes - ATG doesn't match any isoform, stop codon varies
      
      # Case 8a: Internal start codon - truncation
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff != 0 & pr_cterm_diff != 0 & pr_nterm_gene_diff != 0 & utr_cat == "subset" ~ 
        "pIPM,nterm_truncation",
      
      # Case 8b: Novel ATG (protein or TSS context), stop matches another isoform
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff != 0 & pr_cterm_diff != 0 & pr_nterm_gene_diff != 0 & 
        utr_cat == "unique" & pr_cterm_gene_diff == 0 ~ 
        "pNNC,novel_nterm_known_splice_known_cterm",
      
      # Case 8c: Both ATG and stop codon are novel (don't match any isoform)
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff != 0 & pr_cterm_diff != 0 & pr_nterm_gene_diff != 0 & 
        utr_cat == "unique" & pr_cterm_gene_diff != 0 ~ 
        "pNNC,novel_nterm_known_splice_novel_cterm",
      
      # ISM orphan catch-all
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") ~ 
        "orphan_ism",
      
      # ------------------------------------------------------------------------------------------
      # SQANTI_Protein, Novel in Catalog (NIC) - combo_splice (no reference isoform to compare to)
      # ------------------------------------------------------------------------------------------
      
      # Case 1: Both ATG and stop codon match an annotated isoform
      pr_splice_cat == "novel_in_catalog" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_gene_diff == 0 & pr_cterm_gene_diff == 0 ~ 
        "pNIC,known_nterm_combo_splice_known_cterm",
      
      # Case 2: ATG matches an isoform, stop codon is novel
      pr_splice_cat == "novel_in_catalog" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_gene_diff == 0 & pr_cterm_gene_diff != 0 ~ 
        "pNNC,known_nterm_combo_splice_novel_cterm",
      
      # Case 3: ATG doesn't match - truncation
      pr_splice_cat == "novel_in_catalog" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_gene_diff != 0 & utr_cat == "subset" ~ 
        "pIPM,nterm_truncation",
      
      # Case 4: Novel ATG (protein or TSS context), stop codon matches
      pr_splice_cat == "novel_in_catalog" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_gene_diff != 0 & utr_cat == "unique" & pr_cterm_gene_diff == 0 ~ 
        "pNNC,novel_nterm_combo_splice_known_cterm",
      
      # Case 5: Both ATG and stop codon are novel
      pr_splice_cat == "novel_in_catalog" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_gene_diff != 0 & utr_cat == "unique" & pr_cterm_gene_diff != 0 ~ 
        "pNNC,novel_nterm_combo_splice_novel_cterm",
      
      # NIC orphan catch-all
      pr_splice_cat == "novel_in_catalog" & !str_detect(pr_splice_subcat, "mono-exon") ~ 
        "orphan_nic",
      
      # ----------------------------------------------------------------------------------------------
      # SQANTI_Protein, Novel Not in Catalog (NNC) - novel_splice (no reference isoform to compare to)
      # ----------------------------------------------------------------------------------------------
    
      # Case 1: Both ATG and stop codon match an annotated isoform
      pr_splice_cat == "novel_not_in_catalog" & pr_nterm_gene_diff == 0 & pr_cterm_gene_diff == 0 ~ 
        "pNNC,known_nterm_novel_splice_known_cterm",
      
      # Case 2: ATG matches an isoform, stop codon is novel
      pr_splice_cat == "novel_not_in_catalog" & pr_nterm_gene_diff == 0 & pr_cterm_gene_diff != 0 ~ 
        "pNNC,known_nterm_novel_splice_novel_cterm",
      
      # Case 3: ATG doesn't match - truncation
      pr_splice_cat == "novel_not_in_catalog" & pr_nterm_gene_diff != 0 & utr_cat == "subset" ~ 
        "pIPM,nterm_truncation",
      
      # Case 4: Novel ATG (protein or TSS context), stop codon matches
      pr_splice_cat == "novel_not_in_catalog" & pr_nterm_gene_diff != 0 & 
        utr_cat == "unique" & pr_cterm_gene_diff == 0 ~ 
        "pNNC,novel_nterm_novel_splice_known_cterm",
      
      # Case 5: Both ATG and stop codon are novel
      pr_splice_cat == "novel_not_in_catalog" & pr_nterm_gene_diff != 0 & 
        utr_cat == "unique" & pr_cterm_gene_diff != 0 ~ 
        "pNNC,novel_nterm_novel_splice_novel_cterm",
      
      # NNC orphan catch-all
      pr_splice_cat == "novel_not_in_catalog" ~ 
        "orphan_nnc",
      
      # ------------------------------
      # Mono-exon cases
      # ------------------------------
      
      # Perfect mono-exon match
      (pr_splice_subcat == "mono-exon" | pr_splice_subcat == "mono-exon_by_intron_retention") & 
        pr_splice_cat == "full-splice_match" & pr_nterm_diff == 0 & pr_cterm_diff == 0 ~ 
        "pFPM,mono-exon",
      
      # Intergenic mono-exon
      (pr_splice_subcat == "mono-exon" | pr_splice_subcat == "mono-exon_by_intron_retention") & 
        pr_splice_cat == "intergenic" ~ 
        "intergenic,mono-exon",
      
      # Other mono-exon cases (orphan)
      (pr_splice_subcat == "mono-exon" | pr_splice_subcat == "mono-exon_by_intron_retention") ~ 
        "orphan_monoexon,mono-exon",
      
      # ------------------------------
      # Miscellaneous multi-exon
      # ------------------------------
      
      pr_splice_subcat == "multi-exon" & pr_splice_cat == "intergenic" ~ 
        "intergenic,multi-exon",
      
      pr_splice_subcat == "multi-exon" & pr_splice_cat == "genic" ~ 
        "genic,multi-exon",
      
      pr_splice_subcat == "multi-exon" & pr_splice_cat == "antisense" ~ 
        "antisense,multi-exon",
      
      pr_splice_subcat == "multi-exon" & pr_splice_cat == "fusion" ~ 
        "fusion,multi-exon",
      
      # ------------------------------
      # Unclassified
      # ------------------------------
      
      TRUE ~ ""
    )
  ) %>%
  # Split protein_classification into base and subset
  separate(protein_classification, 
           into = c("protein_classification_base", "protein_classification_subset"), 
           sep = ",", 
           fill = "right", 
           remove = FALSE)

message("\n--- Protein classification complete! ---")
message("\nBy classification base:")
message(sprintf("  pFPM: %d", sum(pb_classifications$protein_classification_base == "pFPM", na.rm = TRUE)))
message(sprintf("  pIPM: %d", sum(pb_classifications$protein_classification_base == "pIPM", na.rm = TRUE)))
message(sprintf("  pNIC: %d", sum(pb_classifications$protein_classification_base == "pNIC", na.rm = TRUE)))
message(sprintf("  pNNC: %d", sum(pb_classifications$protein_classification_base == "pNNC", na.rm = TRUE)))
message(sprintf("  intergenic: %d", sum(pb_classifications$protein_classification_base == "intergenic", na.rm = TRUE)))
message(sprintf("  genic: %d", sum(pb_classifications$protein_classification_base == "genic", na.rm = TRUE)))
message(sprintf("  antisense: %d", sum(pb_classifications$protein_classification_base == "antisense", na.rm = TRUE)))
message(sprintf("  fusion: %d", sum(pb_classifications$protein_classification_base == "fusion", na.rm = TRUE)))
message(sprintf("  orphan: %d", sum(grepl("orphan", pb_classifications$protein_classification_base), na.rm = TRUE)))

message("\nAnalysis complete!")
```

```{r}
sample = read_tsv("~/megan/github/LRP2_lite/results/protein_sqanti/sample_5utr_classification.tsv")
sample %>% group_by(utr_exon_status, junc_cat) %>% tally()
sample %>% group_by(utr_cat) %>% tally()

# check ORF file for orf coords, incldues stop codon
orfs = read_tsv("~/megan/github/LRP2_lite/results/orf_calling/sample_best_orfs_mapped.tsv")


names(sqanti_class)
test = sqanti_class %>% mutate(
  pr_cterm_diff = ifelse(!is.na(pr_cterm_diff), pr_cterm_diff + 3, pr_cterm_diff),
  pr_cterm_gene_diff = ifelse(!is.na(pr_cterm_gene_diff), pr_cterm_gene_diff - 3, pr_cterm_gene_diff),
  pr_chang = ifelse(!is.na(pr_chang), pr_chang - 3, pr_chang))

test %>% group_by(pr_cterm_diff) %>% tally() %>% arrange(desc(n)) # 75710, equals 0 so no diff and 51276 NA (NIC and NNC)
test %>% group_by(pr_cterm_gene_diff) %>% tally() %>% arrange(desc(n)) # 99117, equals 0 so no diff and 650 NA
test %>% group_by(pr_chang) %>% tally() %>% arrange(desc(n)) #=
test

```


```{r}
# add fasta to orf calling module
best_orfs = read_tsv("../results/orf_calling/sample_best_orfs_mapped.tsv", show_col_types = FALSE)
best_orfs %>% filter(gene_id == "ENSG00000157191.20")
unique_orf = best_orfs %>% distinct(gene_id, gene_name, orf_isoform_id, orf_protein_sequence)
protein_seqs = AAStringSet(unique_orf$orf_protein_sequence)
names(protein_seqs) = paste0(unique_orf$orf_isoform_id, "|", unique_orf$gene_id, "|", unique_orf$gene_name)
writeXStringSet(protein_seqs, "../results/orf_calling/sample_")


```

What is the agreement between transcript and protein annotation?
Issue in the protein sqanti code
I think it selects one transcript even if multiple transcripts match
some duplication in gene names
```{r}
sqanti = read_tsv("~/megan/github/LRP2_lite/results/sqanti/sample_classification_filtered.txt")
sqanti %>% filter(associated_gene == "ENSG00000157191.20")

protein_classifications = read_tsv("~/megan/github/LRP2_lite/results/protein_sqanti/sample.sqanti_protein_classification.tsv") # 148,829
protein_classifications %>% filter(tx_gene != pr_gene) # 1,717
protein_classifications %>% filter(tx_gene == pr_gene)
protein_classifications %>% separate_rows(pr_gene, sep = ",")
protein_classifications %>% separate_rows(tx_gene, sep = ",")
protein_classifications %>% filter(grepl(",", tx_gene))
protein_classifications %>% filter(grepl(",", pr_gene))

# add section in protein classification script
protein_classifications = read_tsv("~/megan/github/LRP2_lite/results/protein_sqanti/sample_protein_classification.tsv")
gencode_gtf_path = "/project/sheynkman/external_data/GENCODE_v47/gencode.v47.annotation.gtf"
gencode_gtf = import(gencode_gtf_path) %>% as.data.frame()

message("\n--- STEP 5: Reconcile gene ids and names ---")

gene_mapping = gencode_gtf %>% distinct(gene_id, gene_name)
ensg_lookup  = setNames(gene_mapping$gene_name, gene_mapping$gene_id)

# Just do direct assignment - no functions at all
protein_classifications$tx_gene_name = ensg_lookup[protein_classifications$tx_gene]
protein_classifications$pr_gene_name = ensg_lookup[protein_classifications$pr_gene]

# For the columns with commas, manually fix them (this is much faster than ifelse)
has_comma_tx = which(grepl(",", protein_classifications$tx_gene))
has_comma_pr = which(grepl(",", protein_classifications$pr_gene))

message("Transcripts mapping to multiple genes at tx level: ", length(has_comma_tx))
message("Transcripts mapping to multiple genes at pr level: ", length(has_comma_pr))

for (i in has_comma_tx) {
  ensgs = strsplit(protein_classifications$tx_gene[i], ",")[[1]]
  protein_classifications$tx_gene_name[i] = paste(unique(ensg_lookup[ensgs]), collapse = ",")
}

for (i in has_comma_pr) {
  ensgs = strsplit(protein_classifications$pr_gene[i], ",")[[1]]
  protein_classifications$pr_gene_name[i] = paste(unique(ensg_lookup[ensgs]), collapse = ",")
}

# Reconciliation of gene ids
matching = protein_classifications$tx_gene == protein_classifications$pr_gene
protein_classifications$reconciled_gene_id = protein_classifications$pr_gene  
protein_classifications$reconciled_gene_name = protein_classifications$pr_gene_name

# Only need to reconcile non-matching rows
non_matching = which(!matching)

message("Rows that don't match and need reconciliation:", length(non_matching), "\n")

for (i in non_matching) {
  tx_ids  = strsplit(protein_classifications$tx_gene[i], ",")[[1]]
  pr_ids  = strsplit(protein_classifications$pr_gene[i], ",")[[1]]
  
  tx_name = strsplit(protein_classifications$tx_gene_name[i], ",")[[1]]
  pr_name = strsplit(protein_classifications$pr_gene_name[i], ",")[[1]]
  
  match_id   = intersect(tx_ids, pr_ids)
  match_name = intersect(tx_name, pr_name)
  
  if (length(match_id) > 0) {
    protein_classifications$reconciled_gene_id[i] = match_id[1]
  } else {
    protein_classifications$reconciled_gene_id[i] = pr_ids[1]
  }
  
  if (length(match_name) > 0) {
    protein_classifications$reconciled_gene_name[i] = match_name[1]
  } else {
    protein_classifications$reconciled_gene_name[i] = pr_name[1]
  }
}

# check examples- check na
protein_classifications %>% 
  filter(grepl(",", tx_gene)) %>%
  select(tx_gene, pr_gene, tx_gene_name, pr_gene_name, reconciled_gene_id, reconciled_gene_name)

protein_classifications %>% 
  filter(grepl(",", pr_gene)) %>%
  select(tx_gene, pr_gene, tx_gene_name, pr_gene_name, reconciled_gene_id, reconciled_gene_name)

```

Filtering SQANTI protein
```{r}
min_junctions_after_stop_codon = 0
pclass_base_to_keep = c('pFPM', 'pNIC')
sample_cds_fasta_path = "~/megan/github/LRP2_lite/results/orf_calling/sample_best_orfs_collapsed.fa"


# Subclass lookup for short names, important for browser tracks
subclass_lookup = c(
  'known_nterm_novel_splice_known_cterm' = 'kn_ns_kc',
  'known_nterm_known_splice_known_cterm' = 'kn_ks_kc',
  'known_nterm_combo_splice_known_cterm' = 'kn_cs_kc',
  'known_nterm_novel_splice_novel_cterm' = 'kn_ns_nc',
  'known_nterm_combo_splice_novel_cterm' = 'kn_cs_nc',
  'known_nterm_known_splice_novel_cterm' = 'kn_ks_nc',
  'novel_nterm_known_splice_known_cterm' = 'nn_ks_kc',
  'novel_nterm_known_splice_novel_cterm' = 'nn_ks_nc',
  'novel_nterm_novel_splice_known_cterm' = 'nn_ns_kc',
  'novel_nterm_combo_splice_novel_cterm' = 'nn_cs_nc',
  'novel_nterm_combo_splice_known_cterm' = 'nn_cs_kc',
  'novel_nterm_novel_splice_novel_cterm' = 'nn_ns_nc',
  'combo_nterm_cterm'   = 'combo',
  'nterm_truncation'    = 'ntrunc',
  'multi-exon'          = 'multi',
  'mono-exon'           = 'mono'
)

full_protein = protein_classifications %>%
  filter(!is.na(protein_classification)) %>%
  mutate(
    filter_status = case_when(
      
      # Check 1: Filter if high NMD signal (applies to ALL)
      num_junc_after_stop_codon > min_junctions_after_stop_codon ~ "NMD",
      
      # Check 2: Filter if problematic patterns
      grepl('intergenic|antisense|fusion|orphan|genic', protein_classification) ~ "atypical",
      
      # Check 3: Filter if NOT in keep base classification list
      !(protein_classification_base %in% pclass_base_to_keep) ~ "pc_base",
      
      # Keep everything else
      TRUE ~ "keep"
    ),
    
    # abbreviate
    psubclass_short = subclass_lookup[protein_classification_subset]

  )

# Output full table with filter status
full_protein %<>%
  dplyr::select(isoform_id, tx_gene, pr_gene, reconciled_gene_id, reconciled_gene_name, num_junc_after_stop_codon, num_5utr_exons,
         tclass = tx_cat, pclass = protein_classification_base, psubclass = protein_classification_subset, psubclass_short, filter_status)
  
write_tsv(full_protein, file.path(protein_sqanti_dir, paste0(basename, "_protein_classification_all.tsv")))

# Output filtered text file
filtered_protein = full_protein %>% filter(filter_status == "keep")
write_tsv(filtered_protein, file.path(protein_sqanti_dir, paste0(basename, "_protein_high_confidence.txt")))

# Output filtered gtf 
gtf          = import(sample_cds_gtf_path)
gtf_filtered = gtf[gtf$transcript_id %in% filtered_protein$isoform_id]
gtf_meta     = as.data.frame(mcols(gtf_filtered)) # pull metadata

new_meta = filtered_protein %>% 
  mutate(transcript_id = paste(reconciled_gene_name, isoform_id, pclass, sep = "|")) %>%
  select(isoform_id, gene_id = reconciled_gene_id, gene_name = reconciled_gene_name, transcript_id)

updated_meta = gtf_meta %>% 
  select(source, type, score, phase, isoform_id = transcript_id, ORF_id) %>%
  left_join(new_meta, by = "isoform_id") %>%
  select(-ORF_id, everything(), ORF_id)

mcols(gtf_filtered) = updated_meta
export(gtf_filtered, file.path(protein_sqanti_dir, paste0(basename, "_protein_high_confidence.gtf")), format="gtf")

# Output filtered fasta, and adjust header pb|PB.1001.26|fullname GN=MYCBP
fa = readDNAStringSet(sample_cds_fasta_path)

filtered_fa_seqs = data.frame(
  header   = names(fa),
  sequence = as.character(fa),
  length   = width(fa),
  row.names = NULL
) %>%
  separate(header,
           into = c("isoform_id", "old_gene_id", "old_gene_name"),
           sep = "\\|") %>%
  separate_rows(isoform_id, sep = ",") %>%
  select(isoform_id, sequence, length) %>%
  distinct() %>%
  filter(isoform_id %in% filtered_protein$isoform_id)

# combine with reconciled naming, update orf ids post filtering
fasta_out = filtered_fa_seqs %>% 
  left_join(select(filtered_protein, isoform_id, reconciled_gene_id, reconciled_gene_name)) %>%
  group_by(sequence, length) %>%
  arrange(isoform_id) %>%
  summarize(orf_isoform_id = paste(isoform_id, collapse = ","), gene_name = paste(unique(reconciled_gene_name), collapse = ",")) %>%
  ungroup()

protein_seqs        = AAStringSet(fasta_out$sequence)
names(protein_seqs) = paste0(fasta_out$orf_isoform_id, "|", fasta_out$gene_name)
writeXStringSet(protein_seqs, file.path(protein_sqanti_dir, paste0(basename, "_protein_high_confidence.fa")))

```




```{r}


# testing
test %>% group_by(protein_classification_base, filter_status) %>% tally()
test %>% group_by(protein_classification_subset) %>% tally()

protein_classifications %>% filter(protein_classification_base == "pIPM") %>% arrange(num_junc_after_stop_codon)
protein_classifications %>% filter(num_junc_after_stop_codon > min_junctions_after_stop_codon) %>%
  group_by(protein_classification_base) %>% tally()

protein_classifications %>% filter(protein_classification_base == "pFPM" & num_junc_after_stop_codon > min_junctions_after_stop_codon)
full_protein %>% filter(is.na(protein_classification_subset))
full_protein %>% filter(is_nmd == FALSE) %>% group_by(num_junc_after_stop_codon) %>% tally()
full_protein %>% filter(filter_status == "NMD" & is_nmd == TRUE) # 25562

# numbers are off because of AA sequences mapping to different genes?
fasta_out
test = fa[keep_seqs]
data.frame(header = names(test), sequence = as.character(test), row.names = NULL) %>%
  group_by(sequence) %>% tally() %>% arrange(desc(n))
data.frame(header = names(test), sequence = as.character(test), row.names = NULL) %>% 
  filter(sequence == "MSGRGKGGKGGKGGAKRHRKVRDNGTKARRARRGGVKRSGYTRGVKVNVRDAVTYTHAKRKTVTAMDVVYAKRGRTYGGG")

```



Jurkat testing
```{r}
og_gtf = import("~/datasets/jurkat_zenodo5234651/jurkat_chr22.collapsed.gff")
gtf = import("../jurkat_test_results/pacbio_isoseq/jurkat_chr22_merged.collapsed.gff")
og_gtf %>% as.data.frame() %>% filter(type == "transcript") %>% arrange(start, end)
gtf %>% as.data.frame() %>% filter(type == "transcript") %>% arrange(start, end)

sample = read_tsv("../results/sqanti/sample_classification.txt")


og_sqanti = read_tsv("~/datasets/jurkat_zenodo5234651/jurkat_chr22_classification.txt")

sqanti = read_tsv("../jurkat_test_results/sqanti/jurkat_chr22_classification.txt") #diff_to_TSS, diff_to_TTS, diff_to_gene_TSS signs are all flipped

og_sqanti
sqanti %>% arrange(isoform)

# orf calling
new = read_tsv("../jurkat_test_results/orf_calling/jurkat_chr22_cpat.ORF_prob.tsv")
old = read_tsv("~/datasets/jurkat_zenodo5234651/jurkat_chr22.ORF_prob.tsv")
all.equal(new, old) # for cpat results, this is expected

new = read_tsv("../jurkat_test_results/orf_calling/jurkat_chr22_best_orfs_mapped.tsv")
old = read_tsv("~/datasets/jurkat_zenodo5234651/jurkat_chr22_best_orf.tsv")
new %>% select(-ID) %>% arrange(isoform_id)
old %>% arrange(pb_acc)

PB.1.14
```


```{r}
metadata      = fread(metadata_path)
sample_lookup = setNames(metadata$name, metadata$bam_id)

classification_file = "~/megan/github/LRP2_lite/results/sqanti_transcript/sample_classification.txt"
sqanti_df = read_tsv(classification_file, show_col_types = FALSE)

colnames(sqanti_df) = sub("^FL\\.", "", colnames(sqanti_df))
colnames(sqanti_df) = ifelse(colnames(sqanti_df) %in% names(sample_lookup),
                           paste0(sample_lookup[colnames(sqanti_df)], "_counts"),
                           colnames(sqanti_df))

counts_cols = grep("_counts$", colnames(sqanti_df), value = TRUE)
sqanti_df %<>%
  mutate(across(all_of(counts_cols), 
                ~ (.x / sum(.x)) * 1e6,
                .names = "{gsub('_counts', '_cpm', .col)}"))

sqanti_df_full = sqanti_df %>%
  filter(!is.na(associated_gene), str_starts(associated_gene, "ENS"))

sqanti_df_full %<>% 
  left_join(gencode_gene, by = "associated_gene") %>% 
  left_join(gencode_df, by = c("associated_gene", "associated_transcript", "gene_type", "gene_name"))

# combine with hashids, and selct tpm columns
hashids = read_tsv(mapping_file)

sqanti_ids = sqanti_df_full %>% 
  select(isoform_id = isoform, ensg_gene_id = associated_gene, gene_name, enst_transcript_id = associated_transcript, transcript_name, gene_type, protein_id, ends_with("_counts"), ends_with("_cpm"))

all_ids = hashids %>% select(isoform_id = transcript_id, hash_id) %>%
  inner_join(sqanti_ids, by = "isoform_id") 
all_ids %>% distinct(isoform_id)
all_ids %>% distinct(ensg_gene_id, hash_id)

write_tsv(all_ids, "~/megan/github/LRP2_lite/results/sqanti_transcript/sample_all_hashids_with_cpm.txt")

```

```{r}
sample_cds_fasta_path  <- "../results/orf_calling/sample_best_orfs_collapsed.fa"

readDNAStringSet(sample_cds_fasta_path)
readAAStringSet(sample_cds_fasta_path)


```


```{r}

filtered_protein = read_tsv("../results/protein_sqanti/sample_protein_high_confidence.txt")
fa = readAAStringSet("../results/orf_calling/sample_best_orfs_collapsed.fa")
cpms = read_tsv("~/megan/github/LRP2_lite/results/sqanti_transcript/sample_hashids_with_cpm_filtered.txt")

# update orf ids post filtering and summarize cpm
filtered_fa_seqs %>%
  group_by(sequence, length) %>%
  arrange(isoform_id) %>%
  summarize(orf_isoform_id = paste(isoform_id, collapse = ","), 
            gene_id = paste(unique(reconciled_gene_id), collapse = ","),
            gene_name = paste(unique(reconciled_gene_name), collapse = ","),
            hash_id = paste(hash_id, collapse = ","),
            across(ends_with("_counts"), sum),
            across(ends_with("_cpm"), sum)) %>%
  ungroup()


```


DGE
```{r}
library(edgeR)
library(DRIMSeq)
count_file_path = "../results/sqanti_transcript/sample_hashids_with_cpm_filtered.txt"
orf_count_file_path = "../results/protein_sqanti/sample_hashids_with_cpm_ORF.txt"
metadata_file_path = "../config/sample_metadata.txt"

control_group <- as.character("NBM")
experimental_group <- as.character("AML")

# =============================================================================
# Configure transcript-level data for edgeR
# =============================================================================

message("\n--- Reading transcript and ORF count files and sample sheet ---")

# Gene and transcript df
counts_raw         = as.data.frame(fread(count_file_path, header = TRUE))
count_cols         = grep("_counts$", colnames(counts_raw), value = TRUE)
cpm_cols           = grep("_cpm$", colnames(counts_raw), value = TRUE)
count_sample_names = sub("_counts$", "", count_cols)

# orf df- assumed that samples are in the same order as in the transcript level df above
orf_counts_raw         = as.data.frame(fread(orf_count_file_path, header = TRUE))
orf_count_cols         = grep("_counts$", colnames(orf_counts_raw), value = TRUE)
cpm_cols               = grep("_cpm$", colnames(orf_counts_raw), value = TRUE)
orf_count_sample_names = sub("_counts$", "", orf_count_cols)

# Check that transcript and ORF samples order matches
if (!identical(count_sample_names, orf_count_sample_names)) {
  stop("Sample order differs between transcript and ORF count files!\n",
       "Transcript samples: ", paste(count_sample_names, collapse = ", "), "\n",
       "ORF samples: ", paste(orf_count_sample_names, collapse = ", "))
}

# reorder sample sheet
sample_metadata = as.data.frame(fread(metadata_file_path, header = TRUE))
sample_metadata = sample_metadata[match(count_sample_names, sample_metadata$name), ] # reorders

# Check for any mismatches
if (any(is.na(sample_metadata$name))) {
  stop("Some count columns don't have matching samples in metadata file!")
}

# When creating the factor, set control as reference (first level)
control   = control_group
experimental = experimental_group

group        = factor(sample_metadata$group, levels = c(control, experimental))
sample_names = sample_metadata$name

# =============================================================================
# Differential Gene Expression with edgeR
# =============================================================================

cat("Running gene-level differential expression analysis...\n")

# Aggregate transcripts to genes using ensg_gene_id
counts_gene_matrix = counts_raw %>%
  select(ensg_gene_id, all_of(count_cols)) %>%
  group_by(ensg_gene_id) %>%
  summarise(across(everything(), sum)) %>%
  column_to_rownames("ensg_gene_id") %>%
  as.data.frame()

# Create DGEList object
dge_raw = DGEList(counts = counts_gene_matrix, group = group)
colnames(dge_raw$counts) = sample_names
dge_raw$samples$samples  = sample_names

if (!identical(colnames(dge_raw$counts), dge_raw$samples$samples)) {
  stop("Count matrix columns and sample names don't match!")
}

# Filter, normalize, and estimate dispersion
dge = dge_raw[filterByExpr(dge_raw), , keep.lib.sizes = TRUE]
dge = calcNormFactors(dge, method = "TMM")
dge = estimateDisp(dge, robust = TRUE)

# Run differential expression test
result_dge  = exactTest(dge, pair = c(control, experimental)) # so that logFC>0 means higher in AML
dge_results = topTags(result_dge, n = Inf)$table

# Summary
cat(sprintf("Genes with FDR < 0.05: %d\n", sum(dge_results$FDR < 0.05)))
print(summary(decideTests(result_dge)))

# Save results- combine with gene ids and cpm results
gene_cpms = counts_raw %>%
  select(ensg_gene_id, gene_name, all_of(cpm_cols)) %>%
  group_by(ensg_gene_id, gene_name) %>%
  summarise(across(everything(), sum)) %>%
  ungroup() %>%
  distinct()

dge_results %>%
  rownames_to_column("ensg_gene_id") %>%
  left_join(gene_cpms) %>%
  write_tsv(file.path(multisample_analysis_dir, paste0(basename, "_DGE_edgeR_results.txt")))

cpm(dge_raw) %>%
  as.data.frame() %>%
  rownames_to_column("ensg_gene_id") %>%
  left_join(gene_cpms) %>%
  write_tsv(file.path(multisample_analysis_dir, paste0(basename, "_DGE_edgeR_raw_CPM_matrix.txt")))

cpm(dge) %>%
  as.data.frame() %>%
  rownames_to_column("ensg_gene_id") %>%
  left_join(gene_cpms) %>%
  write_tsv(file.path(multisample_analysis_dir, paste0(basename, "_DGE_edgeR_normalized_CPM_matrix.txt")))

# Generate MD plot
pdf(file.path(multisample_analysis_dir, paste0(basename, "_DGE_MD_plot.pdf")), width = 8, height = 6)
plotMD(result_dge)
abline(h = c(-1, 1), col = "blue")
dev.off()

# =============================================================================
# Differential Transcript Expression with edgeR
# =============================================================================
cat("Running transcript-level differential expression analysis...\n")

# Extract transcript count matrix
counts_transcript_matrix = counts_raw %>%
  select(isoform_id, all_of(count_cols)) %>%
  column_to_rownames("isoform_id") %>%
  as.data.frame()

# Create DGEList object
dte_raw = DGEList(counts = counts_transcript_matrix, group = group)
colnames(dte_raw$counts) = sample_names
dte_raw$samples$samples  = sample_names

if (!identical(colnames(dte_raw$counts), dte_raw$samples$samples)) {
  stop("Count matrix columns and sample names don't match!")
}

# Filter, normalize, and estimate dispersion
dte = dte_raw[filterByExpr(dte_raw), , keep.lib.sizes = TRUE]
dte = calcNormFactors(dte, method = "TMM")
dte = estimateDisp(dte, robust = TRUE)

# Run differential expression test
result_dte  = exactTest(dte, pair = c(control, experimental))
dte_results = topTags(result_dte, n = Inf)$table

# Summary
cat(sprintf("Transcripts with FDR < 0.05: %d\n", sum(dte_results$FDR < 0.05)))
print(summary(decideTests(result_dte)))

# Save results
transcript_cpms = counts_raw %>%
  select(isoform_id, hash_id, ensg_gene_id, gene_name, enst_transcript_id, transcript_name, all_of(cpm_cols)) %>%
  distinct()

dte_results %>%
  rownames_to_column("isoform_id") %>%
  left_join(transcript_cpms) %>%
  write_tsv(file.path(multisample_analysis_dir, paste0(basename, "_DTE_edgeR_results.txt")))

cpm(dte_raw) %>%
  as.data.frame() %>%
  rownames_to_column("isoform_id") %>%
  left_join(transcript_cpms) %>%
  write_tsv(file.path(multisample_analysis_dir, paste0(basename, "_DTE_edgeR_raw_CPM_matrix.txt")))

cpm(dte) %>%
  as.data.frame() %>%
  rownames_to_column("isoform_id") %>%
  left_join(transcript_cpms) %>%
  write_tsv(file.path(multisample_analysis_dir, paste0(basename, "_DTE_edgeR_normalized_CPM_matrix.txt")))

# Generate MD plot
pdf(file.path(multisample_analysis_dir, paste0(basename, "_DTE_MD_plot.pdf")), width = 8, height = 6)
plotMD(result_dte)
abline(h = c(-1, 1), col = "blue")
dev.off()

# =============================================================================
# Differential ORF Expression with edgeR
# =============================================================================
cat("Running ORF-level differential expression analysis...\n")

# Extract protein count matrix
counts_orf_matrix = orf_counts_raw %>%
  select(orf_isoform_id, all_of(orf_count_cols)) %>%
  column_to_rownames("orf_isoform_id") %>%
  as.data.frame()

# Create DGEList object
dpe_raw = DGEList(counts = counts_orf_matrix, group = group)

# Replace entire samples data frame from gene analysis, keep the original library sizes
dpe_raw$samples          = dge_raw$samples
colnames(dpe_raw$counts) = sample_names

if (!identical(colnames(dpe_raw$counts), dpe_raw$samples$samples)) {
  stop("Count matrix columns and sample names don't match!")
}

# Filter, normalize, and estimate dispersion
dpe = dpe_raw[filterByExpr(dpe_raw), , keep.lib.sizes = TRUE]
dpe = calcNormFactors(dpe, method = "TMM")
dpe = estimateDisp(dpe, robust = TRUE)

# Run differential expression test
result_dpe  = exactTest(dpe, pair = c(control, experimental))
dpe_results = topTags(result_dpe, n = Inf)$table

# Summary
cat(sprintf("ORFs with FDR < 0.05: %d\n", sum(dpe_results$FDR < 0.05)))
print(summary(decideTests(result_dpe)))

# Save results as TSV with IDs as columns
orf_cpms = orf_counts_raw %>%
  select(orf_isoform_id, hash_id, gene_id, gene_name, all_of(cpm_cols)) %>%
  distinct()

dpe_results %>%
  rownames_to_column("orf_isoform_id") %>%
  left_join(orf_cpms) %>%
  write_tsv(file.path(multisample_analysis_dir, paste0(basename, "_DE_ORF_edgeR_results.txt")))

cpm(dpe_raw) %>%
  as.data.frame() %>%
  rownames_to_column("orf_isoform_id") %>%
  left_join(orf_cpms) %>%
  write_tsv(file.path(multisample_analysis_dir, paste0(basename, "_DE_ORF_edgeR_raw_CPM_matrix.txt")))

cpm(dpe) %>%
  as.data.frame() %>%
  rownames_to_column("orf_isoform_id") %>%
  left_join(orf_cpms) %>%
  write_tsv(file.path(multisample_analysis_dir, paste0(basename, "_DE_ORF_edgeR_normalized_CPM_matrix.txt")))

# Generate MD plot
pdf(file.path(multisample_analysis_dir, paste0(basename, "_DE_ORF_MD_plot.pdf")), width = 8, height = 6)
plotMD(result_dpe)
abline(h = c(-1, 1), col = "blue")
dev.off()

# =============================================================================
# Differential Transcript Usage with DRIMSeq
# =============================================================================
cat("Running differential transcript usage analysis with DRIMSeq...\n")

counts_drimseq = counts_raw %>%
  select(gene_id = ensg_gene_id,      # DRIMSeq needs this named "gene_id"
         feature_id = isoform_id,     # DRIMSeq needs this named "feature_id"
         all_of(count_cols)) %>%
  rename_with(~ sub("_counts$", "", .), .cols = all_of(count_cols))

# Create sample table for DRIMSeq
sample_table = data.frame(
  sample_id = sample_names,
  group = group
)

# Create and filter DRIMSeq object
#counts_drimseq = head(x = as.data.frame(counts_drimseq), n = 200)
d = dmDSdata(counts = as.data.frame(counts_drimseq), samples = sample_table)

cat(sprintf("Before filtering: %d genes, %d transcripts\n", 
            length(unique(counts(d)$gene_id)), 
            nrow(counts(d))))

d = dmFilter(d, 
             min_samps_gene_expr    = nrow(sample_table)/2, # requires gene to be expressed in half of samples
             min_gene_expr          = 10, # based on gene level counts
             min_samps_feature_expr = 0, # isoform
             min_feature_expr       = 0) # isoform

cat(sprintf("After filtering: %d genes, %d transcripts\n", 
            length(unique(counts(d)$gene_id)), 
            nrow(counts(d))))

# Design, fit, and test
design_full = model.matrix(~ group, data = samples(d))
set.seed(123)
d = dmPrecision(d, design = design_full)
d = dmFit(d, design = design_full)
d = dmTest(d, coef = paste0("group", experimental))

# Extract results
dtu_gene_results = results(d) %>%
  dplyr::rename(lr_gene         = lr, 
                df_gene         = df,
                pvalue_gene     = pvalue, 
                adj_pvalue_gene = adj_pvalue)

dtu_tx_results = results(d, level = "feature") %>%
  dplyr::rename(isoform_id            = feature_id,
                lr_transcript         = lr,
                df_transcript        = df,
                pvalue_transcript     = pvalue,
                adj_pvalue_transcript = adj_pvalue)

# Get proportions from DRIMSeq and calculate means and deltas
props = proportions(d)

props_long = props %>%
  pivot_longer(cols = -c(gene_id, feature_id), 
               names_to = "sample_id", 
               values_to = "proportion") %>%
  left_join(sample_table, by = "sample_id")

mean_props = props_long %>%
  group_by(gene_id, feature_id, group) %>%
  summarise(mean_proportion = mean(proportion, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = group, 
              values_from = mean_proportion,
              names_prefix = "mean_prop_") %>%
  mutate(delta_proportion = .data[[paste0("mean_prop_", levels(group)[2])]] - 
                            .data[[paste0("mean_prop_", levels(group)[1])]])

dtu_summary = dtu_tx_results %>%
  left_join(dtu_gene_results, by = "gene_id") %>%
  left_join(props, by = c("gene_id", "isoform_id" = "feature_id")) %>%
  rename_with(~paste0(., "_prop"), .cols = all_of(sample_names)) %>%
  left_join(mean_props, by = c("gene_id", "isoform_id" = "feature_id")) %>%
  left_join(transcript_cpms, by = c("gene_id" = "ensg_gene_id", "isoform_id")) %>%
  select(isoform_id, enst_transcript_id, transcript_name, lr_transcript, pvalue_transcript, adj_pvalue_transcript, ends_with("_cpm"), ends_with("_prop"), starts_with("mean_prop_"), delta_proportion,
         gene_id, gene_name, lr_gene, df_gene, pvalue_gene, adj_pvalue_gene, hash_id)

# Summary
cat(sprintf("Genes with DTU (adj_pvalue < 0.05): %d\n", 
            sum(dtu_gene_results$adj_pvalue_gene < 0.05, na.rm = TRUE)))
cat(sprintf("Transcripts with differential usage (adj_pvalue < 0.05): %d\n", 
            sum(dtu_tx_results$adj_pvalue_transcript < 0.05, na.rm = TRUE)))

# Save results
write_tsv(dtu_summary, file.path(multisample_analysis_dir, paste0(basename, "_DTU_transcript_DRIMSeq_summary.txt")))

# =============================================================================
# Differential ORF Usage with DRIMSeq
# =============================================================================
cat("Running differential ORF usage analysis with DRIMSeq...\n")

# Prepare DRIMSeq format from already-loaded ORF counts
counts_drimseq_orf = orf_counts_raw %>%
  select(gene_id,
         feature_id = orf_isoform_id,
         all_of(orf_count_cols)) %>%
  rename_with(~ sub("_counts$", "", .), .cols = all_of(orf_count_cols))

# Create sample table for DRIMSeq (reuse from transcript analysis)
sample_table_orf = data.frame(
  sample_id = sample_names,
  group = group
)

# Create and filter DRIMSeq object
d_orf = dmDSdata(counts = as.data.frame(counts_drimseq_orf), samples = sample_table_orf)

cat(sprintf("Before filtering (ORF DU): %d genes, %d ORFs\n", 
            length(unique(counts(d_orf)$gene_id)), 
            nrow(counts(d_orf))))

d_orf = dmFilter(d_orf, 
                 min_samps_gene_expr = nrow(sample_table_orf)/2,
                 min_gene_expr = 10,
                 min_samps_feature_expr = 0,
                 min_feature_expr = 0)

cat(sprintf("After filtering (ORF DU): %d genes, %d ORFs\n", 
            length(unique(counts(d_orf)$gene_id)), 
            nrow(counts(d_orf))))

# Design, fit, and test
design_full_orf = model.matrix(~ group, data = samples(d_orf))
set.seed(123)
d_orf = dmPrecision(d_orf, design = design_full_orf)
d_orf = dmFit(d_orf, design = design_full_orf)
d_orf = dmTest(d_orf, coef = paste0("group", experimental))

# Extract results
dpu_gene_results = results(d_orf) %>%
  dplyr::rename(lr_gene         = lr, 
         df_gene         = df,
         pvalue_gene     = pvalue, 
         adj_pvalue_gene = adj_pvalue)

dpu_orf_results = results(d_orf, level = "feature") %>%
  dplyr::rename(orf_isoform_id = feature_id,
                lr_orf         = lr,
                df_orf         = df,
                pvalue_orf     = pvalue,
                adj_pvalue_orf = adj_pvalue)

# Get proportions from DRIMSeq and calculate means and deltas
orf_usage_table = orf_cpms %>% 
  group_by(gene_id, gene_name) %>%
  mutate(across(ends_with("_cpm"), 
                ~ . / sum(., na.rm = TRUE),
                .names = "{.col}_prop")) %>%
  ungroup() %>%
  rename_with(~ str_replace(., "_cpm_prop$", "_prop"), ends_with("_cpm_prop"))

props_orf = proportions(d_orf)

props_orf_long = props_orf %>%
  pivot_longer(cols = -c(gene_id, feature_id), 
               names_to = "sample_id", 
               values_to = "proportion") %>%
  left_join(sample_table, by = "sample_id") %>% 
  distinct(gene_id, feature_id, proportion, group)

group_orf_props = props_orf_long %>%
  pivot_wider(names_from = group, 
              values_from = proportion,
              names_prefix = "group_prop_") %>%
  mutate(delta_proportion = .data[[paste0("group_prop_", levels(group)[2])]] - 
                            .data[[paste0("group_prop_", levels(group)[1])]])

dpu_summary = dpu_orf_results %>%
  left_join(dpu_gene_results, by = "gene_id") %>%
  left_join(group_orf_props, by = c("gene_id", "orf_isoform_id" = "feature_id")) %>%
  left_join(orf_usage_table, by = c("gene_id", "orf_isoform_id")) %>%
  select(orf_isoform_id, hash_id, lr_orf, pvalue_orf, adj_pvalue_orf, ends_with("_cpm"), ends_with("_prop"), starts_with("group_prop_"), delta_proportion,
         gene_id, gene_name, lr_gene, df_gene, pvalue_gene, adj_pvalue_gene)

# Summary
cat(sprintf("Genes with DPU (adj_pvalue < 0.05): %d\n", 
            sum(dpu_gene_results$adj_pvalue_gene < 0.05, na.rm = TRUE)))
cat(sprintf("ORFs with differential usage (adj_pvalue < 0.05): %d\n", 
            sum(dpu_orf_results$adj_pvalue_orf < 0.05, na.rm = TRUE)))

# Save results
write_tsv(dpu_summary, file.path(multisample_analysis_dir, paste0(basename, "_DU_ORF_DRIMSeq_summary.txt")))

```

```{r}
options(scipen = 999)

transript_usage_table = transcript_cpms %>% 
  group_by(ensg_gene_id, gene_name) %>%
  mutate(across(ends_with("_cpm"), 
                ~ . / sum(., na.rm = TRUE),
                .names = "{.col}_prop")) %>%
  ungroup() %>%
  rename_with(~ str_replace(., "_cpm_prop$", "_prop"), ends_with("_cpm_prop"))

props = proportions(d) # these are calculated per group not per sample

props_long = props %>%
  pivot_longer(cols = -c(gene_id, feature_id), 
               names_to = "sample_id", 
               values_to = "proportion") %>%
  left_join(sample_table, by = "sample_id") %>% 
  distinct(gene_id, feature_id, proportion, group)

group_props = props_long %>%
  pivot_wider(names_from = group, 
              values_from = proportion,
              names_prefix = "group_prop_") %>%
  mutate(delta_proportion = .data[[paste0("group_prop_", levels(group)[2])]] - 
                            .data[[paste0("group_prop_", levels(group)[1])]])

dtu_summary = dtu_tx_results %>%
  left_join(dtu_gene_results, by = "gene_id") %>%
  left_join(group_props, by = c("gene_id", "isoform_id" = "feature_id")) %>%
  left_join(transript_usage_table, by = c("gene_id" = "ensg_gene_id", "isoform_id")) %>%
  select(isoform_id, enst_transcript_id, transcript_name, hash_id, lr_transcript, pvalue_transcript, adj_pvalue_transcript, ends_with("_cpm"), ends_with("_prop"), starts_with("group_prop_"), delta_proportion,
         gene_id, gene_name, lr_gene, df_gene, pvalue_gene, adj_pvalue_gene)

```

Long read leafcutter
- input gtf and count file (sqanti transcript)
- gtf to psl function
- psl coords function
- isoform to junction
- minicutter
- Give sub isoform counts to diff splicing code, might need to tweak naming (could use hashing)

839763 sample_corrected.psl
```{r}
# test updated gtf to psl
new_psl = gtf_to_psl("/home/cwp5au/megan/github/LRP2_lite/results/sqanti_transcript/sample_corrected.gtf")

old_psl = fread("/home/cwp5au/megan/github/LRP2_lite/results/sqanti_transcript/sample_corrected.psl")

new = as.data.frame(new_psl)
old = as.data.frame(old_psl)
identical(old, new)
all.equal(old, new)

colnames(old) <- NULL
colnames(new) <- NULL
all.equal(old, new, check.attributes = FALSE) # TRUE
identical(as.matrix(old), as.matrix(new)) # TRUE
```

Data wrangling for leafcutter
```{r}
transcript_gtf = "/home/cwp5au/megan/github/LRP2_lite/results/sqanti_transcript/sample_corrected_filtered.gtf"
psl = gtf_to_psl("/home/cwp5au/megan/github/LRP2_lite/results/sqanti_transcript/sample_corrected_filtered.gtf")
#write_tsv(psl, "/home/cwp5au/megan/github/LRP2_lite/results/sqanti_transcript/sample_corrected_filtered.psl", col_names = FALSE)
fread("/home/cwp5au/megan/github/LRP2_lite/results/sqanti_transcript/sample_corrected_filtered.psl")
coords = psl_to_coords("/home/cwp5au/megan/github/LRP2_lite/results/sqanti_transcript/sample_corrected_filtered.psl")
metadata_file_path = "../config/sample_metadata.txt"
count_file_path = "../results/sqanti_transcript/sample_hashids_with_cpm_filtered.txt"


intron_coords = coords$intron_coords


counts = fread(count_file_path) %>% 
  select(trans_id = isoform_id, ends_with("_counts"), ends_with("_cpm"))

isoform_to_junction <- function(intron_coords, counts){
  
  # reshape to long to handle multiple samples
  counts_long = counts %>%
    pivot_longer(
    cols = -trans_id,
    names_to = c("sample", ".value"),
    names_pattern = "(.+)_(counts|cpm)"
  ) %>%
  replace_na(list(counts = 0, cpm = 0))
  
  # join isoform->junction map, then aggregate to junction per sample
  jc = intron_coords %>% left_join(counts_long, by = "trans_id") %>% 
    group_by(junc_id, chr, intron_start, intron_end, gene_id, strand, sample) %>%
    dplyr::summarise(isoforms = paste0(unique(trans_id), collapse = ","), 
                     lr_junc_count = sum(counts, na.rm = TRUE),
                     lr_junc_cpm   = sum(cpm, na.rm = TRUE),
                     .groups = "drop") %>%
    dplyr::arrange(chr, intron_start, junc_id, sample)
  
  return(jc)
  
}

lr_junctions = isoform_to_junction(intron_coords, counts)

# get single counts per junction
lr_junctions_collapsed = lr_junctions %>% 
  group_by(chr, intron_start, intron_end, junc_id, strand) %>%
  summarize(readcount = sum(lr_junc_count, na.rm = TRUE)) %>%
  ungroup() %>%
  relocate(strand, .after = last_col())

clusters = minicutter(juncs = lr_junctions_collapsed, plot_summary = TRUE, min_usage_ratio = 0.01)
as_clusters = clusters %>% 
  select(chr = chrom, start, end, junc_id = name, cluster_idx, strand) %>%
  group_by(cluster_idx) %>%
  filter(n() > 1) %>%
  ungroup()

# format lr junctions for overlap
junc_to_iso = lr_junctions %>% distinct(junc_id, isoforms)

# overlap clusters to get isoform mapping
# as_clusters %>% 
#   left_join(junc_to_iso, by = c("junc_id")) %>%
#   separate_rows(isoforms, sep = ",") %>%
#   group_by(chr, cluster_idx, isoforms, strand) %>%
#   arrange(start) %>%
#   summarise(blocknumber = n(), 
#             blockstarts = paste0(unique(start), collapse = ","),
#             blockends = paste0(unique(end), collapse = ",")) %>%
#   ungroup()

# need to sum per sample per subisoform, this gets tricky
# filter(cluster_idx == "6407")
by_isoform = as_clusters %>%
  left_join(junc_to_iso, by = c("junc_id")) %>%
  separate_rows(isoforms, sep = ",") %>%
  group_by(chr, cluster_idx, isoforms, strand) %>%
  arrange(start, end) %>%
  summarise(intron_starts = paste0(unique(start), collapse = ","),
            intron_ends = paste0(unique(end), collapse = ",")) %>%
  ungroup() 

# include counts here to sum
by_isoform %<>% 
  left_join(select(counts, -ends_with("_cpm")), by = c("isoforms" = "trans_id"))

subisoform = by_isoform %>%
  group_by(chr, cluster_idx, strand, intron_starts, intron_ends) %>%
  summarise(isoforms = paste0(unique(isoforms), collapse = ","),
            across(ends_with("_counts"), sum),
            .groups = "drop") %>%
  mutate(subisoform_id = paste0(chr, ":", intron_starts, ":", intron_ends, ":clu_", cluster_idx)) %>%
  relocate((subisoform_id))

write_tsv(subisoform, file = file.path(multisample_analysis_dir, paste0(basename, "_lr_leafcutter_subisoform_clusters.txt")), col_names = TRUE)

# write out space delimited file with row names for diff splicing script
out = subisoform %>%
  select(subisoform_id, ends_with("_counts")) %>%
  rename_with(~gsub("_counts$", "", .), ends_with("_counts")) %>%
  column_to_rownames("subisoform_id")

write.table(out, 
            file = gzfile(file.path(multisample_analysis_dir, paste0(basename, "_lr_leafcutter_perind_numers.counts.gz"))),
            sep = " ",
            row.names = TRUE,
            col.names = TRUE,
            quote = FALSE)

# create the group text file required for ds (use sample sheet)
sample_metadata = as.data.frame(fread(metadata_file_path, header = TRUE))
sample_metadata %>% select(name, group) %>%
  write_tsv(file.path(multisample_analysis_dir, paste0(basename, "_groups_file.txt")), col_names = FALSE)




```





```{r}
# checking
intron_coords %>% distinct(junc_id) # 204,605
lr_junctions %>% filter(junc_id == "chr1_1013576_1013984")
lr_junctions_collapsed %>% filter(junc_id == "chr1_1013576_1013984")

```

# check cds gtf
```{r}
gtf_data = import("~/megan/AML_project/LRP2_lite/results/protein_sqanti/all_AML_protein_high_confidence.gtf", format = "gtf") %>%
    as.data.table() 

gtf_data %>% filter(type == "CDS")
```

<<<<<<< HEAD:scripts/LRP2_lite_testing.Rmd
January 14, 2026
# filter cpat
Fix: remove requirement for stop codon if gencode match
Fix: cpat filter no longer requires sqanti classification file. the gtf is required to have gene_id and transcript_id attributes and those will be used.
Fix: Moving the orf collapsing script to the protein classification module so that CPAT filter is simpler. We dont do anything other than classify and filter cpat orfs.

Check the SREK1 example in AML data
```{r}
# requirements for 03_filter_cpat

basename         <- "all_AML"
sqanti_dir       <- file.path("/home/cwp5au/megan/AML_project/LRP2_lite/results", "sqanti_transcript") 
cpat_dir         <- file.path("/home/cwp5au/megan/AML_project/LRP2_lite/results", "orf_calling") 
gencode_gtf_path <- "/project/sheynkman/external_data/GENCODE_v47/gencode.v47.annotation.gtf"
coding_threshold <- as.numeric("0.364")

cpat_fasta_file        <- file.path(cpat_dir, paste0(basename, "_cpat.ORF_seqs.fa"))
cpat_results_file      <- file.path(cpat_dir, paste0(basename, "_cpat.ORF_prob.tsv"))
sample_full_fasta_path <- file.path(sqanti_dir, paste0(basename, "_corrected_filtered.fasta"))
sample_gtf_path        <- file.path(sqanti_dir, paste0(basename, "_corrected_filtered.gtf"))

```



```{r}
call_best_orfs <- function(mapped_orfs) {
  
  # ORF quality scoring parameters but these don't seem to be used
  coding_threshold  # CPAT protein-coding threshold
  atg_shift  <- 10          # Sigmoid parameters for ATG penalty
  atg_growth <- 0.5
  
  # per transcript, if there is a gencode match, pick the one with the fewest upstream ATGs
  mapped_orfs %<>% 
    mutate(atg_penalty     = 1 - 1/(1 + exp(-atg_growth * (upstream_atgs - atg_shift))), 
           composite_score = Coding_prob * (1 - atg_penalty),
           orf_quality     = case_when(
             Coding_prob <= coding_threshold ~ "Low Quality ORF",
             !has_stop_codon & !gencode_match ~ "No Stop Codon",
             TRUE ~ "Plausible ORF"
           ))
  
  best_orfs = mapped_orfs %>% 
    filter(orf_quality == "Plausible ORF") %>%
    group_by(isoform_id) %>% 
    arrange(desc(gencode_match), upstream_atgs, desc(Coding_prob)) %>%
    slice_head(n = 1) %>%
    ungroup()
  
  mapped_orfs %<>%
    mutate(orf_quality = ifelse(ID %in% best_orfs$ID, "Clear Best ORF", orf_quality))
  
  return(mapped_orfs)
}

full_mapped_orfs = mapped_orfs
mapped_orfs %<>% filter(isoform_id == "PB.18418.1")

test = call_best_orfs(full_mapped_orfs)
test %>% filter(isoform_id == "PB.18418.1") # no stop codon but gencode match
test %>% filter(isoform_id == "PB.18418.82") # NMD
test %>% filter(isoform_id == "PB.18418.57") # no stop codon but gencode match
test %>% filter(isoform_id == "PB.18418.18") # NMD
test %>% filter(isoform_id == "PB.18418.52") # NMD

test %>% filter(isoform_id == "PB.18418.272")
```

```{r}
# === STEP 1: Load all input data ===
message("\n--- STEP 1: Loading input data ---")

# CPAT results
cpat_results = read_tsv(cpat_results_file, show_col_types = FALSE) %>%
  separate(ID, into = c("isoform_id", "orf_rank"), sep = "_ORF_", remove = FALSE)

# Load GTF files
sample_gtf  = import(sample_gtf_path) %>% as.data.frame()
gencode_gtf = import(gencode_gtf_path) %>% as.data.frame()

# Load sample fasta with full DNA sequence- include UTRs, filtered transcripts from SQANTI
full = readDNAStringSet(sample_full_fasta_path)
full_fasta = tibble(
  isoform_id    = names(full),
  full_sequence = as.character(full)
)

# SQANTI classification
# classification = read_tsv(classification_file, show_col_types = FALSE)
# gene_mapping = classification %>% 
#   select(isoform_id = isoform, 
#          structural_category, 
#          gene_id = associated_gene, 
#          transcript_id = associated_transcript, 
#          any_of(c("gene_name", "transcript_name")))

```


```{r}
# === STEP 4: Define the best ORF per transcript ===
message("\n--- STEP 4: Calling best ORFs and filtering---")

mapped_orfs_classified = call_best_orfs(mapped_orfs)
#mapped_orfs_classified %<>% left_join(gene_mapping, by = "isoform_id")

all_orfs       = mapped_orfs_classified
plausible_orfs = mapped_orfs_classified %>% filter(orf_quality == "Clear Best ORF" | orf_quality == "Plausible ORF")
best_orfs      = mapped_orfs_classified %>% filter(orf_quality == "Clear Best ORF")

write_tsv(best_orfs, file.path(cpat_dir, paste0(basename, "_best_orfs_mapped.tsv")))
write_tsv(all_orfs, file.path(cpat_dir, paste0(basename, "_all_orfs_mapped.tsv")))

# === STEP 5: Write gtf for best ORFs, including CDS and exon types, no collapsing here ===
message("\n--- STEP 6: Writing GTF of best ORFs with CDS and exon types ---")
all_cds_exons = get_cds_coords(best_orfs, sample_exons)
# gene_mapping %<>% 
#   left_join(select(best_orfs, isoform_id, orf_isoform_id)) %>%
#   mutate(orf_isoform_id = ifelse(is.na(orf_isoform_id), "noORF", orf_isoform_id))

all_cds_exons %<>% 
  left_join(distinct(sample_gtf, transcript_id, gene_id), by = c("isoform_id" = "transcript_id"))
updated_gtf = write_gtf_with_cds(sample_gtf, all_cds_exons)

write.table(updated_gtf, file.path(cpat_dir, paste0(basename, "_corrected_filtered_CDS.gtf")), 
            sep = "\t", 
            quote = FALSE, 
            row.names = FALSE, 
            col.names = FALSE)

# =============================================================================
# Summary
# =============================================================================

starting_transcripts = nrow(all_sample_transcripts)
n_transcripts_orfs   = n_distinct(all_orfs$isoform_id)
percent_orfs         = round((n_transcripts_orfs/starting_transcripts) * 100, 4)

n_cpat_orfs          = nrow(all_orfs)
n_plausible_orfs     = nrow(plausible_orfs)
percent_plausible    = round((n_plausible_orfs/n_cpat_orfs) * 100, 1)

n_best_orfs          = nrow(best_orfs %>% distinct(isoform_id))
n_noORFs             = starting_transcripts - n_best_orfs
percent_best         = round((n_best_orfs/starting_transcripts) * 100, 1)
percent_noORFs       = round((n_noORFs/starting_transcripts) * 100, 1)


message("\n=== CPAT FILTER ANALYSIS COMPLETE ===")
message(paste0("- CPAT called at least one ORF for ", n_transcripts_orfs, " / ", starting_transcripts, " (", percent_orfs,"%)", " transcripts."))
message(paste0("- Based on coding potential, ", n_plausible_orfs, " / ", n_cpat_orfs, " (", percent_plausible,"%)"," of CPAT ORFs were classified as plausible ORFs.")) 
message(paste0("- ", n_noORFs, " / ", starting_transcripts, " (", percent_noORFs,"%)", " transcripts do not have a plausible ORF.")) 
message(paste0("- After further filtering, a 'Clear Best ORF' was identified for ", n_best_orfs, " / ", starting_transcripts, " (", percent_best,"%)", " transcripts.")) 

message("\n=== CPAT FILTER OUTPUT FILES ===")
message(paste0("All CPAT ORFs with Quality Metrics: ", basename, "_all_cpat_orfs_mapped.tsv"))
message(paste0("The single best plausible ORF per transcript: ", basename, "_best_cpat_orfs_mapped.tsv"))
message(paste0("GTF contains exon type for all transcripts and CDS transcripts with a best plausible ORF: ", basename, "_corrected_filtered_CDS.gtf"))

```


January 15, 2026

Check hash id uniqueness on 12 million isoseq transcripts
```{r}
source("~/megan/github/megan-schertzer/ms_functions.R")

# 12 million transcripts
hash = fread("/home/cwp5au/megan/AML_project/260112_lrp2_test/S1_PACBIO_ISOSEQ/M3_ISOSEQ_COLLAPSE/AML_hashids_mapping.txt") # 12,767,372

hash_gene = hash %>% 
  group_by(gene_id, hash_id) %>%
  summarize(pb_ids = paste0(transcript_id, collapse = ",")) %>%
  ungroup() # 11,591,258

hash_gene %>% distinct(hash_id) # 11,535,642

hash_collapsed = hash_gene %>% 
  group_by(hash_id) %>%
  summarize(
    n_id = n(),
    gene_id = paste0(gene_id, collapse = ","),
    pb_ids = paste0(pb_ids, collapse = ";")) %>%
  ungroup() %>%
  arrange(desc(n_id))

redundant = hash_collapsed %>% filter(n_id > 1) # 4515

write_tsv(redundant, "/home/cwp5au/megan/AML_project/260112_lrp2_test/S1_PACBIO_ISOSEQ/M3_ISOSEQ_COLLAPSE/AML_redundant_hashids.txt")

redundant = read_tsv("/home/cwp5au/megan/AML_project/260112_lrp2_test/S1_PACBIO_ISOSEQ/M3_ISOSEQ_COLLAPSE/AML_redundant_hashids.txt")
redundant %>% select(hash_id, n_id) 

# s46b9dd2b0ba88d13:e46b9dd2b0ba88d13 has 51k matches?
hash_gene %>% filter(hash_id == "s46b9dd2b0ba88d13:e46b9dd2b0ba88d13")

# more likely, 1.6 million transcripts
hash = fread("/home/cwp5au/megan/AML_project/LRP2_lite/results/sqanti_transcript/all_AML_hashids_mapping.txt") # 1,650,770

hash_gene = hash %>% 
  group_by(gene_id, hash_id) %>%
  summarize(pb_ids = paste0(transcript_id, collapse = ",")) %>%
  ungroup() # 1,459,435

hash_gene %>% distinct(hash_id) # 1,423,084
redundant_hash = hash_gene %>% filter(hash_id == "s46b9dd2b0ba88d13:e46b9dd2b0ba88d13") 
hash_gene %>% filter(hash_id != "s46b9dd2b0ba88d13:e46b9dd2b0ba88d13") # 1,423,877
other_redundant = hash_gene %>% filter(hash_id != "s46b9dd2b0ba88d13:e46b9dd2b0ba88d13") %>%
  group_by(hash_id) %>%
  tally() %>%
  filter(n > 1) # 794

sqanti = fread("/home/cwp5au/megan/AML_project/LRP2_lite/results/sqanti_transcript/all_AML_classification.txt") # 1,650,770
gtf = import("/home/cwp5au/megan/AML_project/LRP2_lite/results/pacbio_isoseq/all_AML_merged.collapsed.gff") %>%
  as.data.frame()

# look at very redundant hash- these are monoexons
redundant_hash = hash_gene %>% filter(hash_id == "s46b9dd2b0ba88d13:e46b9dd2b0ba88d13") %>% separate_rows(pb_ids)
sqanti %>% filter(isoform %in% c(redundant_hash$pb_ids)) %>% group_by(subcategory) %>% tally()

# look at other examples
hash_gene %>% filter(hash_id == "s001ef32b6bfe4937:e4b4acc7c44d4fa89") # PB.55218.3, PB.55220.18
sqanti %>% filter(isoform %in% c("PB.55218.3", "PB.55220.18")) # intron chain matches but strand differs, one looks like internal priming, considering adding strand

gtf %>% filter(transcript_id == "PB.55218.3")
gtf %>% filter(transcript_id == "PB.55220.18")

hash_gene %>% filter(hash_id == "s007ca1c358e0ef36:ee81cf9caa746d6c7")
sqanti %>% filter(isoform %in% c("PB.13551.879", "PB.13554.1")) # same, strandedness
gtf %>% filter(transcript_id == "PB.13551.879")
gtf %>% filter(transcript_id == "PB.13554.1")

# check broadly for strandedness
other = other_redundant %>% 
  left_join(hash_gene) %>%
  mutate(isoform = str_extract(pb_ids, "^[^,]+")) %>%
  select(hash_id, isoform) %>%
  left_join(sqanti, by = "isoform")

other %>% group_by(exons) %>% tally() # 1398 have 2 exons
strand_check = other %>%
  group_by(hash_id) %>%
  summarize(
    strand_consistent = n_distinct(strand) == 1,
    strands = paste(unique(strand), collapse = ","),
    n_isoforms = n()
  )

strand_check %>% group_by(strand_consistent) %>% tally() # 702 false, 92 true
same_strand = strand_check %>% filter(strand_consistent == TRUE)

# if not different strand, what are these?
other %>% filter(hash_id %in% same_strand$hash_id)
hash_gene %>% filter(hash_id == "s0d12b371f0a94ca4:ef25ad78c1a983880") 
sqanti %>% filter(isoform %in% c("PB.13787.1971", "PB.3675.78")) 

gtf %>% filter(transcript_id == "PB.13787.1971") # different chr but actually same intron chain
gtf %>% filter(transcript_id == "PB.3675.78")

gtf %>% filter(transcript_id == "PB.33036.226") # different chr but actually same intron chain
gtf %>% filter(transcript_id == "PB.48200.6")

gtf %>% filter(transcript_id == "PB.55666.91") # chrX/Y same junction chain
gtf %>% filter(transcript_id == "PB.57711.28")

# look at chr
other %>% filter(hash_id %in% same_strand$hash_id) %>%
  group_by(hash_id) %>%
  summarise(chroms = paste(unique(chrom), collapse = ","),
            exons_n = paste(exons, collapse = ",")) %>%
  ungroup() %>%
  filter(chroms != "chrX,chrY") # 34

```


January 19, 2025

Test original sqanti protein
conda activate sqanti3

python scripts/sqanti3_protein_input_full_gtf.py \
  /home/cwp5au/megan/AML_project/LRP2_lite/results/orf_calling/all_AML_corrected_filtered_CDS.gtf \
  /home/cwp5au/megan/AML_project/LRP2_lite/results/orf_calling/all_AML_best_orfs_mapped.tsv \
  /project/sheynkman/external_data/GENCODE_v47/gencode.v47.annotation.gtf \
  -d /home/cwp5au/megan/AML_project/LRP2_lite/results/test \
  -p all_AML_og


python scripts/sqanti3_protein.py \
  /home/cwp5au/megan/AML_project/LRP2_lite/results/orf_calling/all_AML_corrected_filtered_CDS.gtf \
  /project/sheynkman/external_data/GENCODE_v47/gencode.v47.annotation.gtf \
  -d /home/cwp5au/megan/AML_project/LRP2_lite/results/test \
  -p all_AML_new

Results are the same. Here is a summary of the changes.
- Removed dependency on separate ORF file by extracting CDS information directly from GTF
- new function: extract_cds_info_from_gtf()
- before, the script was using transcript coords from best orf tsv and then converting to genomic coords. this was unnecessary since all of that info is in the gtf.
- Added sorted() to gene concatenation for deterministic output order
```{r}
# compare the outputs
og = read_tsv("/home/cwp5au/megan/AML_project/LRP2_lite/results/test/all_AML_og.sqanti_protein_classification.tsv")
new = read_tsv("/home/cwp5au/megan/AML_project/LRP2_lite/results/test/all_AML_new.sqanti_protein_classification.tsv")

og_test = og %>% filter(isoform_id != "PB.44482.4")
new_test = new %>% filter(isoform_id != "PB.44482.4")

all.equal(og_test, new_test)

og_test
new_test

rows_differ <- !apply(og == new, 1, all, na.rm = TRUE)
rows_differ

og[rows_differ, ]

df_old <- og %>% mutate(row_id = row_number())
df_new <- new %>% mutate(row_id = row_number())

# Anti-join to find differences
diff_rows <- anti_join(og_test, new_test, by = colnames(og_test))

# Comprehensive comparison function
check_differences <- function(df1, df2) {
  cat("=== Dimension Check ===\n")
  cat("df1:", dim(df1), "\n")
  cat("df2:", dim(df2), "\n")
  
  cat("\n=== Identical Check ===\n")
  print(identical(df1, df2))
  
  cat("\n=== all.equal Check ===\n")
  print(all.equal(df1, df2))
  
  cat("\n=== Column-by-Column Check ===\n")
  for (col in colnames(df1)) {
    if (col %in% colnames(df2)) {
      if (!identical(df1[[col]], df2[[col]])) {
        cat("Column", col, "DIFFERS\n")
        
        # Count differences
        n_diff <- sum(df1[[col]] != df2[[col]], na.rm = TRUE)
        cat("  Number of different values:", n_diff, "\n")
        
        # Show example
        diff_idx <- which(df1[[col]] != df2[[col]])[1]
        if (length(diff_idx) > 0 && !is.na(diff_idx)) {
          cat("  Example - Row", diff_idx, "\n")
          cat("    df1:", df1[[col]][diff_idx], "\n")
          cat("    df2:", df2[[col]][diff_idx], "\n")
        }
      }
    } else {
      cat("Column", col, "MISSING in df2\n")
    }
  }
}

# Run it
check_differences(og_test, new_test)
identical(colnames(og_test), colnames(new_test))

```

Edits to Protein classification script
- do all collapsing in this module now
- since sqanti protein re-aligns with gencode, we can get gene name info from here
- removes the need to read in sqanti transcript
- output full protein fasta for mass spec
- add bed visualization for CDS
- use base orf id which is the most abundant transcript with that ORF

```{r}
# requirements for 04_protein_classfication

basename                        <- "all_AML"
sqanti_transcript_dir           <- file.path("/home/cwp5au/megan/AML_project/LRP2_lite/results", "sqanti_transcript")
cpat_dir                        <- file.path("/home/cwp5au/megan/AML_project/LRP2_lite/results", "orf_calling")
protein_sqanti_dir              <- file.path("/home/cwp5au/megan/AML_project/LRP2_lite/results", "protein_sqanti")
gencode_gtf_path                <- "/project/sheynkman/external_data/GENCODE_v47/gencode.v47.annotation.gtf"

sample_dna_fasta_path  <- file.path(sqanti_transcript_dir, paste0(basename, "_corrected_filtered.fasta"))
sample_cds_gtf_path    <- file.path(cpat_dir, paste0(basename, "_corrected_filtered_CDS.gtf"))
mapped_orfs            <- file.path(cpat_dir, paste0(basename, "_all_orfs_mapped.tsv"))
protein_sqanti_path    <- "/home/cwp5au/megan/AML_project/LRP2_lite/results/test/all_AML_og.sqanti_protein_classification.tsv"
cpm_file_path          <- file.path(sqanti_transcript_dir, paste0(basename, "_hashids_with_cpm_filtered.txt"))

PROTEIN_CLASS_KEEP="FPM,NPC,NPE"
min_junctions_after_stop_codon <- as.numeric(0)
pclass_base_to_keep            <- strsplit(PROTEIN_CLASS_KEEP, ",")[[1]]

```

# Done
```{r}
full_protein = protein_classifications %>%
  mutate(
    filter_status = case_when(
      
      # Check 1: Filter if no ORF found
      utr_cat == "no_orf" ~ "no_ORF",
      
      # Check 2: Filter if high NMD signal (applies to ALL)
      num_junc_after_stop_codon > min_junctions_after_stop_codon ~ "NMD",
      
      # Check 3: Filter if problematic patterns
      grepl('intergenic|antisense|fusion|orphan|genic', protein_classification) ~ "sqanti_atypical",
      
      # Check 3: Filter if NOT in keep base classification list
      !(protein_classification_base %in% pclass_base_to_keep) ~ "sqanti_classification",
      
      # Keep everything else
      TRUE ~ "high_confidence"
    ),
    
    # abbreviate
    psubclass_short = subclass_lookup[protein_classification_subset]
    
  )

# Output full table with filter status
full_protein %<>%
  left_join(distinct(sample_gtf, gene_id, transcript_id), by = "transcript_id") %>%
  dplyr::select(transcript_id, gene_id, reference_gene_id = reconciled_gene_id, reference_gene_name = reconciled_gene_name, tx_transcripts, pr_transcripts, 
                num_junc_after_stop_codon, num_nt_after_stop_codon, num_5utr_exons, tss_in_gc_exons, utr_cat, 
                tclass = tx_cat, pclass = protein_classification_base, psubclass = protein_classification_subset, psubclass_short, filter_status)

write_tsv(full_protein, file.path(protein_sqanti_dir, paste0(basename, "_protein_classification_all_isoforms.txt")))

```


# add group by ORF to protein classifcation
# Done
```{r}
group_by_protein_sequence <- function(orf_info, full_fasta) {
  
  orf_proteins = orf_info %>%
    left_join(full_fasta, by = "transcript_id") %>%
    filter(!is.na(full_dna_sequence)) %>%
    mutate(orf_dna_sequence = substr(full_dna_sequence, ORF_start, ORF_end))
  
  # Vectorized translation for speed
  dna      <- DNAStringSet(orf_proteins$orf_dna_sequence)
  proteins <- translate(dna, if.fuzzy.codon = "solve")
  
  orf_proteins %<>%
    mutate(orf_aa_sequence = as.character(proteins)) %>%
    filter(orf_aa_sequence != "") %>%
    select(-orf_dna_sequence, -full_dna_sequence)
  
  # Group transcripts by identical protein sequences, base id should be high confidence isoform with the highest expression
  orf_groups = orf_proteins %>%
    mutate(avg_cpm = rowMeans(select(., contains("cpm")), na.rm = TRUE)) %>%
    mutate(filter_status = factor(filter_status, levels = c("high_confidence", "NMD", "sqanti_classification", "sqanti_atypical"))) %>%
    group_by(orf_aa_sequence, gene_id) %>%
    arrange(filter_status, desc(avg_cpm), .by_group = TRUE) %>%
    mutate(
      orf_all_isoform_id = paste(transcript_id, collapse = ","),
      #orf_hc_isoform_id  = na_if(paste(transcript_id[filter_status == "high_confidence"], collapse = ","), ""),
      orf_base_id        = transcript_id[1]  # First transcript as representative
    ) %>% 
    ungroup()
    
  return(orf_groups)
}

# Generate a matching amino acid fasta for mass spec reference- this will include lower confidence sequences such as NMD
full = readDNAStringSet(sample_dna_fasta_path) # dna fasta
full_fasta = tibble(
  transcript_id = names(full),
  full_dna_sequence = as.character(full)
)

# count matrix
counts = read_tsv(cpm_file_path) %>%
  select(transcript_id = 1, contains("cpm"))

# orf coords
orf_info = read_tsv(mapped_orfs) %>%
  filter(orf_quality == "Clear Best ORF") %>%
  select(transcript_id = isoform_id, ORF_start, ORF_end) %>%
  left_join(counts, by = c("transcript_id")) %>%
  left_join(select(full_protein, 
                   gene_id, transcript_id, pclass, reference_gene_id, reference_gene_name, filter_status), 
            by = "transcript_id")

# group by orf
orf_groups = group_by_protein_sequence(orf_info, full_fasta)

# write fasta for all protein amino acid sequences for mass spec
orf_groups %<>%
  mutate(header = paste0(
    transcript_id,
    "|", gene_id, 
    "|", reference_gene_name,
    "|", pclass,
    "|", filter_status))

protein_seqs        = AAStringSet(orf_groups$orf_aa_sequence)
names(protein_seqs) = orf_groups$header
writeXStringSet(protein_seqs, file.path(protein_sqanti_dir, paste0(basename, ".predicted.proteome.all_best_orfs.fa")))

```

# Write high confidence GTF and count file
# Done
```{r}
# table of high confidence orfs with cpm, for multisample analysis
hc_orf_info = orf_info %>% filter(filter_status == "high_confidence") 
hc_orf_groups = group_by_protein_sequence(hc_orf_info, full_fasta)

name_map = hc_orf_groups %>% distinct(orf_base_id, reference_gene_name)
hc_collapsed = hc_orf_groups %>% 
  select(-avg_cpm) %>%
  group_by(orf_all_isoform_id, orf_base_id, gene_id) %>%
  summarize(across(contains("cpm"), sum, na.rm = TRUE)) %>%
  ungroup() %>%
  left_join(name_map, by = "orf_base_id") %>%
  select(orf_base_id, orf_all_isoform_id, gene_id, reference_gene_name, everything())

write_tsv(hc_collapsed, file.path(protein_sqanti_dir, paste0(basename, ".predicted.proteome.high_confidence_ORF_cpm.txt")))

# write a corresponding ORF centric gtf
gtf = import(sample_cds_gtf_path) %>%
  as.data.frame()

new_orf_attributes = hc_collapsed %>% 
  mutate(avg_orf_cpm = rowMeans(across(contains("cpm")), na.rm = TRUE)) %>%
  group_by(gene_id) %>%
  mutate(
    gene_total_cpm = sum(avg_orf_cpm, na.rm = TRUE),
    avg_orf_ratio = round(avg_orf_cpm / gene_total_cpm, 3)
  ) %>%
  ungroup() %>%
  select(orf_base_id, orf_all_isoform_id, reference_gene_name, avg_orf_ratio)

hc_gtf = gtf %>% 
  filter(type %in% c("transcript", "CDS")) %>%
  filter(transcript_id %in% hc_collapsed$orf_base_id)

orf_boundaries = hc_gtf %>%
  filter(type == "CDS") %>%
  group_by(transcript_id) %>%
  summarize(
    orf_start = min(start),
    orf_end = max(end)
  )

hc_gtf %<>% 
  left_join(orf_boundaries, by = "transcript_id") %>%
  mutate(
    start = if_else(type == "transcript" & !is.na(orf_start), orf_start, start),
    end   = if_else(type == "transcript" & !is.na(orf_end), orf_end, end)
  ) %>%
  select(-orf_start, -orf_end) %>%
  rename(orf_base_id = transcript_id) %>%
  left_join(new_orf_attributes, by = c("orf_base_id")) %>%
  mutate(name = paste0(orf_base_id, "|",
                       reference_gene_name, "|", 
                       avg_orf_ratio)) %>%
  select(-orf_all_isoform_id)

# calculate ratio
gr_updated = makeGRangesFromDataFrame(hc_gtf, keep.extra.columns = TRUE)
export(gr_updated, file.path(protein_sqanti_dir, paste0(basename, ".predicted.proteome.high_confidence_ORF.gtf")), format = "gtf")



```


# Convert to bed12 for IGV/UCSC visualization

```{r}
gtf_to_bed12 <- function(gtf_path, output_bed, color_by = NULL) {
  
  # Import GTF
  gtf <- import(gtf_path)
  df <- as.data.frame(gtf)
  dt <- as.data.table(df)
  
  # Get transcripts
  tx <- dt[type == "transcript"]
  
  # Determine what to use for blocks
  has_exon <- any(dt$type == "exon")
  has_cds <- any(dt$type == "CDS")
  
  if(has_exon) {
    ex <- dt[type == "exon"]      # Use exons for blocks (shows UTRs)
  } else {
    ex <- dt[type == "CDS"]       # Use CDS for blocks (CDS only)
  }
  
  cds <- dt[type == "CDS"]        # Always get CDS for thick regions
  
  # Set keys for fast joins
  setkey(ex, orf_base_id)
  if(nrow(cds) > 0) setkey(cds, orf_base_id)
  
  # Sort transcripts by gene, then by ratio (descending)
  if(!is.null(color_by)) {
    tx[, sort_val := as.numeric(get(color_by))]
    tx <- tx[order(gene_id, -sort_val)]
  }
  
  # Pre-compute colors
  if(!is.null(color_by)) {
    vals <- as.numeric(tx[[color_by]])
    col_idx <- round((1 - vals) * 99) + 1
    col_idx[is.na(col_idx)] <- 50
    hex_colors <- viridis::viridis(100)[col_idx]
    rgb_matrix <- col2rgb(hex_colors)
    tx$color <- paste(rgb_matrix[1,], rgb_matrix[2,], rgb_matrix[3,], sep=",")
  } else {
    tx$color <- "0,0,0"
  }
  
  # Build BED12 rows
  bed_list <- vector("list", nrow(tx))
  
  for(i in 1:nrow(tx)) {
    t <- tx[i]
    e <- ex[.(t$orf_base_id)][order(start)]
    
    if(nrow(e) == 0) next
    
    # Get CDS boundaries for thick regions
    if(nrow(cds) > 0) {
      t_cds <- cds[.(t$orf_base_id)]
      if(nrow(t_cds) > 0) {
        thick_start <- min(t_cds$start) - 1
        thick_end <- max(t_cds$end)
      } else {
        # Has exons but no CDS = all thin (UTR only)
        thick_start <- t$start - 1
        thick_end <- t$start - 1
      }
    } else {
      # No CDS at all = all thin
      thick_start <- t$start - 1
      thick_end <- t$start - 1
    }
    
    # Build block strings
    block_sizes <- paste(e$width, collapse=",")
    block_starts <- paste(e$start - t$start, collapse=",")
    
    # Build BED12 line
    bed_list[[i]] <- c(
      as.character(t$seqnames), 
      as.character(t$start-1), 
      as.character(t$end), 
      as.character(t$name), 
      "0", 
      as.character(t$strand),
      as.character(thick_start), 
      as.character(thick_end), 
      as.character(t$color), 
      as.character(nrow(e)),
      block_sizes,
      block_starts
    )
  }
  
  # Remove NULLs and write
  bed <- do.call(rbind, bed_list[!sapply(bed_list, is.null)])
  write.table(bed, output_bed, sep="\t", quote=F, row.names=F, col.names=F)
  
}

gtf_to_bed12(gtf_path, output_bed, color_by = "avg_orf_ratio")

```










```{r}
# Output filtered fasta, and adjust header pb|PB.1001.26|fullname GN=MYCBP
fa = readAAStringSet(sample_cds_fasta_path)

filtered_fa_seqs = data.frame(
  header   = names(fa),
  sequence = as.character(fa),
  length   = width(fa),
  row.names = NULL
) %>%
  separate(header,
           into = c("isoform_id", "old_gene_id", "old_gene_name"),
           sep = "\\|") %>%
  separate_rows(isoform_id, sep = ",") %>%
  select(isoform_id, sequence, length) %>%
  distinct() %>%
  filter(isoform_id %in% filtered_protein$isoform_id)

# combine with reconciled naming
filtered_fa_seqs %<>% 
  left_join(select(filtered_protein, isoform_id, reconciled_gene_id, reconciled_gene_name))
  
# create collapsed cpm table
cpms = read_tsv(cpm_file_path)
filtered_fa_seqs %<>% left_join(cpms, by = "isoform_id")

# update orf ids post filtering and summarize cpm
fasta_out = filtered_fa_seqs %>%
  group_by(sequence, length) %>%
  arrange(isoform_id) %>%
  summarize(orf_isoform_id = paste(isoform_id, collapse = ","), 
            gene_id = paste(unique(reconciled_gene_id), collapse = ","),
            gene_name = paste(unique(reconciled_gene_name), collapse = ","),
            hash_id = paste(hash_id, collapse = ","),
            across(ends_with("_counts"), sum),
            across(ends_with("_cpm"), sum)) %>%
  ungroup()

fasta_out %>% select(orf_isoform_id, hash_id, gene_id, gene_name, ends_with("_counts"), ends_with("_cpm")) %>%
  write_tsv(file.path(protein_sqanti_dir, paste0(basename, "_hashids_with_cpm_ORF.txt")))

protein_seqs        = AAStringSet(fasta_out$sequence)
names(protein_seqs) = paste0(fasta_out$orf_isoform_id, "|", fasta_out$gene_name)
writeXStringSet(protein_seqs, file.path(protein_sqanti_dir, paste0(basename, "_protein_high_confidence.fa")))
```




```{r}

# Output filtered text file
#filtered_protein = full_protein %>% filter(filter_status == "keep")
#write_tsv(filtered_protein, file.path(protein_sqanti_dir, paste0(basename, "_protein_high_confidence.txt")))

# Output filtered gtf 
gtf          = import(sample_cds_gtf_path)
gtf_filtered = gtf[gtf$transcript_id %in% filtered_protein$isoform_id]
gtf_meta     = as.data.frame(mcols(gtf_filtered)) # pull metadata

new_meta = filtered_protein %>% 
  mutate(transcript_id = paste(reconciled_gene_name, isoform_id, pclass, sep = "|")) %>%
  select(isoform_id, gene_id = reconciled_gene_id, gene_name = reconciled_gene_name, transcript_id)

updated_meta = gtf_meta %>% 
  select(source, type, score, phase, isoform_id = transcript_id, ORF_id) %>%
  left_join(new_meta, by = "isoform_id") %>%
  select(-ORF_id, everything(), ORF_id)

mcols(gtf_filtered) = updated_meta
export(gtf_filtered, file.path(protein_sqanti_dir, paste0(basename, "_protein_high_confidence.gtf")), format="gtf")


```




```{r}
# === STEP 5: Collapse transcripts by identical ORF protein sequence ===
message("\n--- STEP 5: Collapsing by ORF protein sequence ---")

orf_groups = group_by_protein_sequence(best_orfs, full_fasta)
best_orfs %<>% left_join(orf_groups, by = "isoform_id")

write_tsv(best_orfs, file.path(cpat_dir, paste0(basename, "_best_orfs_mapped.tsv")))

if ("gene_name" %in% colnames(best_orfs)) {
  unique_orf = best_orfs %>% distinct(gene_id, gene_name, orf_isoform_id, orf_protein_sequence)
} else {
  unique_orf = best_orfs %>% distinct(gene_id, orf_isoform_id, orf_protein_sequence)
}

protein_seqs        = AAStringSet(unique_orf$orf_protein_sequence)
names(protein_seqs) = if ("gene_name" %in% colnames(unique_orf)) {
  paste0(unique_orf$orf_isoform_id, "|", unique_orf$gene_id, "|", unique_orf$gene_name)
} else {
  paste0(unique_orf$orf_isoform_id, "|", unique_orf$gene_id)
}
writeXStringSet(protein_seqs, file.path(cpat_dir, paste0(basename, "_best_orfs_collapsed.fa")))

# === STEP 6: Write updated gtfs considering best ORFs ===
message("\n--- STEP 6: Writing updated GTF ---")
all_cds_exons = get_cds_coords(best_orfs, sample_exons)
gene_mapping %<>% 
  left_join(select(best_orfs, isoform_id, orf_isoform_id)) %>%
  mutate(orf_isoform_id = ifelse(is.na(orf_isoform_id), "noORF", orf_isoform_id))

updated_gtf   = write_gtf_with_cds(sample_gtf, all_cds_exons, gene_mapping)

write.table(updated_gtf, file.path(cpat_dir, paste0(basename, "_corrected_filtered_CDS.gtf")), 
            sep = "\t", 
            quote = FALSE, 
            row.names = FALSE, 
            col.names = FALSE)

```


February 2, 2026
Working on feature branch with sqanti transcript R code
- remove dropout folder, add filtering rationle in table
- add bed conversion function
- ensure that code works with new hash code

```{r}
dir                     <- "/home/cwp5au/megan/AML_project/LRP2_lite/results/sqanti_transcript"
basename                <- "all_AML"
gencode_gtf             <- "/project/sheynkman/external_data/GENCODE_v47/gencode.v47.annotation.gtf"
metadata_path           <- "/home/cwp5au/megan/AML_project/LRP2_lite/config/sample_metadata.txt"


filter_protein_coding   <- as.logical(Sys.getenv("PROTEIN_CODING_FILTER", "TRUE"))
filter_internal_priming <- as.logical(Sys.getenv("INTERNAL_PRIMING_FILTER", "TRUE"))
filter_RTS              <- as.logical(Sys.getenv("TEMPLATE_SWITCHING_FILTER", "TRUE"))
percent_polyA_threshold <- as.numeric(Sys.getenv("PERCENT_POLYA_THRESHOLD", "60"))
tclass_to_keep          <- Sys.getenv("TRANSCRIPT_CLASS_KEEP", "FSM,ISM,NIC,NNC")

classification_file = paste0(dir, "/", basename, "_classification.txt")
sqanti_gtf          = paste0(dir, "/", basename, "_corrected.gtf")
sqanti_fasta        = paste0(dir, "/", basename, "_corrected.fasta")
mapping_file        = paste0(dir, "/", basename, "_hashids_mapping.txt")

```

```{r}
df = read_tsv(classification_file, show_col_types = FALSE)
df %>% filter(!is.na(associated_gene), str_starts(associated_gene, "ENS")) #1,206,432
df %>% filter(!is.na(associated_gene), !str_starts(associated_gene, "novelGene"))



# sample gtf
gtf = import(sqanti_gtf) %>%
  as.data.frame()

filtered_gtf = gtf[gtf$transcript_id %in% kept_ids, ]
gtf_ids = filtered_gtf %>% distinct(gene_id, transcript_id)

new_attributes = all_ids %>%
  filter(isoform_id %in% kept_ids) %>%
  select(transcript_id = isoform_id, everything()) %>%
  left_join(gtf_ids) %>%
  mutate(avg_cpm = rowMeans(across(contains("cpm")), na.rm = TRUE)) %>%
  group_by(gene_id) %>%
  mutate(
    gene_total_cpm = sum(avg_cpm, na.rm = TRUE),
    avg_ratio = round(avg_cpm / gene_total_cpm, 3)
  ) %>%
  ungroup() %>%
  select(transcript_id, reference_gene_name = gene_name, avg_ratio)

filtered_gtf %<>% 
  left_join(new_attributes, by = c("transcript_id")) %>%
  mutate(name = paste0(transcript_id, "|",
                       reference_gene_name, "|", 
                       avg_ratio))

# calculate ratio
gr_updated = makeGRangesFromDataFrame(filtered_gtf, keep.extra.columns = TRUE)
export(gr_updated, file.path(dir, paste0(basename, "_corrected_filtered.gtf")), format = "gtf")

```





=======

>>>>>>> dev:bin/LRP2_lite_testing.Rmd


















