---
title: "LRP2 Lite Testing"
output: html_document
---

```{r}
source("/project/sheynkman/users/megan/github/megan-schertzer/ms_functions.R")

#wget https://github.com/conda-forge/miniforge/releases/download/24.11.3-0/Miniforge3-24.11.3-0-Linux-x86_64.sh
#packageVersion()
# added genomic ranges and tidyverse to sqanti3 env

```

What does sqanti classification file have?
```{r}
all = read_tsv("../results/sqanti/sample_classification.txt")

gtf_file_path = "/project/sheynkman/external_data/GENCODE_v47/gencode.v47.annotation.gtf"

gr <- import(gtf_file_path, format = "gtf")
gtf_data <- as.data.table(gr)

reference_df = gtf_data %>% filter(type == "transcript") %>% select(gene_id, gene_type, gene_name, transcript_id, transcript_type, transcript_name, protein_id)
tail(reference_df)
```



sqanti_filter.R
```{r}
#!/usr/bin/env Rscript

#' SQANTI3 Transcript Filtering Script
#' 
#' R script to filter SQANTI3 transcripts based on quality criteria:
#' - Protein-coding genes only
#' - Remove transcripts with internal priming 
#' - Remove template switching artifacts
#' - Filter by SQANTI structural categories
#'
#' Outputs both filtered files and dropout files with reasons.

# =============================================================================
# Load required libraries
# =============================================================================

suppressPackageStartupMessages({
  library(tidyverse)
  library(rtracklayer)
  library(Biostrings) 
  library(argparse)
  library(data.table)
})

# =============================================================================
# Define SQANTI structural categories
# =============================================================================

STRUCTURAL_CATEGORIES <- list(
  strict = c("novel_not_in_catalog", "novel_in_catalog", 
             "incomplete-splice_match", "full-splice_match"),
  all = c("antisense", "novel_not_in_catalog", "novel_in_catalog",
          "incomplete-splice_match", "full-splice_match", "genic",
          "intergenic", "fusion", "genic_intron")
)

# =============================================================================
# Helper functions
# =============================================================================

#' Save filtered sequences to FASTA
#' @param fasta_file Input FASTA file
#' @param keep_ids IDs to keep
#' @param output_file Output FASTA file
save_filtered_fasta <- function(fasta_file, keep_ids, output_file) {
  # Read FASTA
  sequences <- readDNAStringSet(fasta_file)
  
  # Filter sequences
  keep_mask <- names(sequences) %in% keep_ids
  filtered_sequences <- sequences[keep_mask]
  
  # Write filtered FASTA
  writeXStringSet(filtered_sequences, output_file)
  message(paste0("Saved ", length(filtered_sequences), " sequences to ", basename(output_file)))
}

#' Save filtered GTF
#' @param gtf_file Input GTF file
#' @param keep_ids Transcript IDs to keep
#' @param output_file Output GTF file
save_filtered_gtf <- function(gtf_file, keep_ids, output_file) {
  # Read GTF
  gtf <- import(gtf_file)
  
  # Filter by transcript_id
  filtered_gtf <- gtf[gtf$transcript_id %in% keep_ids]
  
  # Write filtered GTF
  export(filtered_gtf, output_file)
  message(paste0("Saved ", length(unique(filtered_gtf$transcript_id)), " transcripts to ", basename(output_file)))
}

# =============================================================================
# Get environment variables for filtering parameters
# =============================================================================
filter_protein_coding = TRUE
filter_internal_priming = TRUE
filter_RTS = TRUE
percent_polyA_threshold = 60

filter_protein_coding <- aTRUEfilter_protein_coding <- as.logical(Sys.getenv("PROTEIN_CODING_FILTER", "TRUE"))
filter_internal_priming <- as.logical(Sys.getenv("INTERNAL_PRIMING_FILTER", "TRUE"))
filter_RTS <- as.logical(Sys.getenv("TEMPLATE_SWITCHING_FILTER", "TRUE"))
percent_polyA_threshold <- as.numeric(Sys.getenv("PERCENT_POLYA_THRESHOLD", "60"))
structural_level <- Sys.getenv("STRUCTURE_FILTER", "strict")

basename <- Sys.getenv("OUTPUT_BASE_NAME")
dir <- file.path(Sys.getenv("OUTPUT_DIR"), "sqanti") 
gencode_gtf <- Sys.getenv("GENCODE_GTF_FILE")

gencode_gtf <- "/project/sheynkman/external_data/GENCODE_v47/gencode.v47.annotation.gtf"

# =============================================================================
# Check required files
# =============================================================================

classification_file <- "/home/cwp5au/megan/github/LRP2_lite/results/sqanti/sample_classification.txt"
classification_file <- paste0(dir, "/", basename, "_classification.txt")
stopifnot("File not found" = file.exists(classification_file))

sqanti_gtf <- "/home/cwp5au/megan/github/LRP2_lite/results/sqanti/sample_corrected.gtf"
sqanti_gtf <- paste0(dir, "/", basename, "_corrected.gtf")
stopifnot("File not found" = file.exists(sqanti_gtf))
 
sqanti_fasta <- "/home/cwp5au/megan/github/LRP2_lite/results/sqanti/sample_corrected.fasta"
sqanti_fasta <- paste0(dir, "/", basename, "_corrected.fasta")
stopifnot("File not found" = file.exists(sqanti_fasta))

# Create additional output directory
dropout_dir <- file.path(dir, "dropout")
dir.create(dropout_dir, recursive = TRUE, showWarnings = FALSE)

# =============================================================================
# Load and merge inputs
# =============================================================================

# Gencode gtf used to select protein coding genes and add column of readable gene name
gencode    <- import(gencode_gtf, format = "gtf")
gencode_df <- as.data.table(gencode)

gencode_df %<>% 
  filter(type == "transcript") %>% 
  select(associated_gene = gene_id, gene_type, gene_name, associated_transcript = transcript_id, transcript_name, protein_id)

gencode_gene = gencode_df %>% 
  select(associated_gene, gene_type, gene_name) %>% 
  distinct()

# SQANTI classification file- keep only isoforms with ENS gene IDs
sqanti_df_full <- read_tsv(classification_file, show_col_types = FALSE) %>%
  filter(!is.na(associated_gene), str_starts(associated_gene, "ENS"))

sqanti_df_full %<>% 
  left_join(gencode_gene) %>% 
  left_join(gencode_df)

# =============================================================================
# Apply filters sequentially
# =============================================================================

# Initialize dropout tracking
original_ids <- sqanti_df_full$isoform
dropout_tracker <- list()

# Keep only transcripts of protein coding genes
sqanti_df = sqanti_df_full # 653,675
if (filter_protein_coding) {
  ids <- sqanti_df$isoform
  sqanti_df %<>% 
    filter(gene_type == "protein_coding")
  
  kept_ids    <- sqanti_df$isoform
  dropped_ids <- setdiff(ids, kept_ids)
  dropout_tracker[["not_protein_coding"]] <- dropped_ids
  
  message("Protein coding filter: kept ", length(kept_ids), " transcripts, dropped ", length(dropped_ids), " transcripts")
}

# Filter based on downstream polyA sequence to remove internal priming
if (filter_internal_priming) {
  ids <- sqanti_df$isoform
  
  # keep all isoforms below threshold
  keep1 = sqanti_df %>% 
    filter(perc_A_downstream_TTS <= percent_polyA_threshold)

  # Get downstream sequence for FSM above threshold
  protected_seqs = sqanti_df %>%
    filter(perc_A_downstream_TTS > percent_polyA_threshold, structural_category == "full-splice_match") %>%
    distinct(associated_gene, seq_A_downstream_TTS)

  # Keep high polyA isoforms that share sequence with FSM transcript
  keep2 = sqanti_df %>% 
    inner_join(protected_seqs, by = c("associated_gene", "seq_A_downstream_TTS"))

  sqanti_df = bind_rows(keep1, keep2) %>% distinct()
  
  kept_ids    <- sqanti_df$isoform
  dropped_ids <- setdiff(ids, kept_ids)
  dropout_tracker[["internal_priming"]] <- dropped_ids
  
  message("Internal priming filter: kept ", length(kept_ids), " transcripts, dropped ", length(dropped_ids), " transcripts")
}

# Filter out template switching, keep RTS_stage = FALSE (no RTS)
if (filter_RTS) {
  ids <- sqanti_df$isoform
  
  # sum read counts
  # Find column positions
  ratio_tss_pos <- which(colnames(sqanti_df) == "ratio_TSS")
  gene_type_pos <- which(colnames(sqanti_df) == "gene_type")

  # Get columns between them (if any exist)
  if (gene_type_pos - ratio_tss_pos > 1) {
    count_cols <- (ratio_tss_pos + 1):(gene_type_pos - 1)
    sqanti_df$avg_counts <- rowMeans(sqanti_df[, count_cols], na.rm = TRUE)
  } else {
    sqanti_df$avg_counts <- NA  # No columns in between
  }
  
  # filtering considerations
  sqanti_df %<>% filter(
    structural_category == "full-splice_match" |
    RTS_stage == FALSE | 
    (RTS_stage == TRUE & all_canonical == TRUE & avg_counts > 3))
  
  kept_ids <- sqanti_df$isoform
  dropped_ids <- setdiff(ids, kept_ids)
  dropout_tracker[["Template_switching_artifact"]] <- dropped_ids
  
  message("Template switching filter: kept ", length(kept_ids), " transcripts, dropped ", length(dropped_ids), " transcripts")
}
  
# Apply structural category filter
allowed_categories <- STRUCTURAL_CATEGORIES[[structural_level]]
ids <- sqanti_df$isoform
sqanti_df %<>% filter(structural_category %in% allowed_categories)

kept_ids <- sqanti_df$isoform
dropped_ids <- setdiff(ids, kept_ids)
dropout_tracker[["structural_category_filtered"]] <- dropped_ids
message("Structural category filter: kept ", length(kept_ids), " transcripts, dropped ", length(dropped_ids), " transcripts")

# Calculate final results
all_dropout_ids <- setdiff(original_ids, kept_ids)
  
message(paste0("Final: ", length(kept_ids), " kept, ", length(all_dropout_ids), " dropped"))

# =============================================================================
# Write output files and generate summary
# =============================================================================

message("Writing output files...")
write_tsv(sqanti_df, file.path(dir, paste0(basename, "_classification_filtered.txt")))
  
save_filtered_fasta(sqanti_fasta, kept_ids, file.path(dir, paste0(basename, "_corrected_filtered.fasta")))
  
save_filtered_gtf(sqanti_gtf, kept_ids, file.path(dir, paste0(basename, "_corrected_filtered.gtf")))
  
# Save dropout files
dropout_reasons_df <- map_dfr(names(dropout_tracker), ~ {
  tibble(isoform = dropout_tracker[[.x]], dropout_reason = .x)
})

write_tsv(dropout_reasons_df, file.path(dropout_dir, paste0(basename, "_dropout_transcripts.tsv")))
save_filtered_fasta(sqanti_fasta, all_dropout_ids, file.path(dropout_dir, paste0(basename, "_corrected_dropout.fasta")))
save_filtered_gtf(sqanti_gtf, all_dropout_ids, file.path(dropout_dir, paste0(basename, "_corrected_dropout.gtf")))
  
# Print summary
cat("\n=== FILTERING SUMMARY ===\n")
cat(paste0("Sample: ", basename, "\n"))
cat(paste0("Original transcripts: ", length(original_ids), "\n"))
cat(paste0("Final transcripts: ", length(kept_ids), "\n"))
cat(paste0("Retention rate: ", round(100 * length(kept_ids) / length(original_ids), 1), "%\n"))
cat("\nDropout breakdown:\n") 

dropout_summary <- dropout_reasons_df %>%
  count(dropout_reason, sort = TRUE)
  
for (i in seq_len(nrow(dropout_summary))) {
  cat(paste0("  ", dropout_summary$dropout_reason[i], ": ", dropout_summary$n[i], "\n"))
    }

test %>% filter(structural_category == "full-splice_match", gene_type == "protein_coding", gene_name == "RBFOX2")
sqanti_df_full %>% filter(gene_name == "RBFOX2")
sqanti_df_full %>% filter(RTS_stage == TRUE)
sqanti_df %>% filter(RTS_stage == FALSE | structural_category == "full-splice_match")
sqanti_df %>% filter(RTS_stage == TRUE & structural_category != "full-splice_match" & all_canonical == "canonical")





sqanti_df %>% filter(perc_A_downstream_TTS <= percent_polyA_threshold | structural_category == "full-splice_match")
sqanti_df %>% filter(perc_A_downstream_TTS > percent_polyA_threshold & structural_category == "full-splice_match")
sqanti_df %>% filter(associated_gene == "ENSG00000215788.11") %>% group_by(associated_gene, seq_A_downstream_TTS) %>% tally()

dropout_reasons_df %>% group_by(dropout_reason) %>% tally()


```

Why are there so many RT switching in the AML long read?
- Inspect support depth
- canonical or non-canonical junctions
- microhomology length distribution > 6 are more likely to be real

SQANTI’s RT-switching rule is purely sequence-based: if a junction is flanked by microhomology, it gets flagged. Many true biological junctions happen to sit near repeats, so you can get multiple reads flagged even though they’re real isoforms. This is the most common explanation when you see >1 read support.

A lot of these seem real to me
```{r}
# this is my analysis in LRP2_lite
all = read_tsv("../results/sqanti/sample_classification.txt") # 839,763
switch = all %>% filter(RTS_stage == "TRUE") # 542,120

# expression
ggplot(all, aes(x = RTS_stage, y = FL.BioSample_1)) + geom_violin() + theme_bw() + scale_y_log10()
switch %>% filter(FL.BioSample_1 > 1) # 153,179
switch %>% filter(FL.BioSample_1 > 5) # 27,193

# canonical splice sites, GT-AG
switch %>% group_by(all_canonical) %>% tally() # 509,234/542,120 are non-canonical
switch %>% filter(FL.BioSample_1 > 1) %>% group_by(all_canonical) %>% tally() # 138,400/153,179 are non-canonical
switch %>% filter(FL.BioSample_1 > 5) %>% group_by(all_canonical) %>% tally() # 20,923/27,193 are non-canonical

# structural category
switch %>% group_by(structural_category) %>% tally() # 355,038 are novel not in catalog

# consider all factors
switch %>% filter(FL.BioSample_1 > 1, all_canonical == "canonical") %>% group_by(structural_category) %>% tally() # 14,779


```

Compare to Emily's analysis- similar percentage
```{r}
em = read_tsv("~/sheynkman/projects/LRP_AML/02_sqanti/AML_classification.txt") # 1,689,238
em_switch = em %>% filter(RTS_stage == "TRUE") # 929,199

```

Compare to MDS Mouse data
```{r}
mds = read_tsv("~/sheynkman/projects/LRP_Mohi_project/02_sqanti/MDS_classification.txt") # 838,021
mds_switch = mds %>% filter(RTS_stage == "TRUE") # 443,272

# expression
mds_switch %>% filter(FL.BioSample_1 > 1) # 74,490
mds_switch %>% filter(FL.BioSample_1 > 5) # 15,122

# canonical splice sites, GT-AG
mds_switch %>% group_by(all_canonical) %>% tally() # 440,301/443,272 are non-canonical
mds_switch %>% filter(FL.BioSample_1 > 1) %>% group_by(all_canonical) %>% tally() 
mds_switch %>% filter(FL.BioSample_1 > 5) %>% group_by(all_canonical) %>% tally() 

# structural category
mds_switch %>% group_by(structural_category) %>% tally() # most are novel not in catalog

# consider all factors
mds_switch %>% filter(FL.BioSample_1 > 1, all_canonical == "canonical") %>% group_by(structural_category) %>% tally() # 1,359

```


From 05_orf_calling
- I dont think expression is useful here. Create an ORF-id that concatanates the transcript ids but keep each row a transcript. can aggregate on ORF ids later.
- dont call orfs separately, call all together

```{r}
cpat_fasta_file = "/home/cwp5au/megan/github/LRP2_lite/results/orf_calling/sample_cpat.ORF_seqs.fa"
cpat_results_file = "/home/cwp5au/megan/github/LRP2_lite/results/orf_calling/sample_cpat.ORF_prob.tsv"
sample_full_fasta_path = "/home/cwp5au/megan/github/LRP2_lite/results/sqanti/sample_corrected_filtered.fasta"
classification_file = "/home/cwp5au/megan/github/LRP2_lite/results/sqanti/sample_classification_filtered.txt"
sample_gtf_path = "/home/cwp5au/megan/github/LRP2_lite/results/sqanti/sample_corrected_filtered.gtf"
gencode_gtf_path = "/project/sheynkman/external_data/GENCODE_v47/gencode.v47.annotation.gtf"
cpat_dir = "/home/cwp5au/megan/github/LRP2_lite/results/orf_calling"
#sample_gtf = import("../results/sqanti/sample_corrected_filtered.gtf") %>% as.data.frame()
#gencode_gtf = import("/project/sheynkman/external_data/GENCODE_v47/gencode.v47.annotation.gtf") %>% as.data.frame()
basename = "sample"

#!/usr/bin/env Rscript

#' ORF Analysis Script - R Version
#' 
#' Map ORF coordinates to genomic positions
#' Compares to GENCODE start codon annotations
#' Selects the best ORF for each transcript
#'
#' Outputs 

# =============================================================================
# Load required libraries
# =============================================================================

suppressPackageStartupMessages({
  library(tidyverse)
  library(rtracklayer)
  library(Biostrings)
})

# =============================================================================
# Get environment variables
# =============================================================================

basename         <- Sys.getenv("OUTPUT_BASE_NAME")
sqanti_dir       <- file.path(Sys.getenv("OUTPUT_DIR"), "sqanti") 
cpat_dir         <- file.path(Sys.getenv("OUTPUT_DIR"), "orf_calling") 
gencode_gtf_path <- Sys.getenv("GENCODE_GTF_FILE")
coding_threshold <- as.numeric(Sys.getenv("CPAT_CODING_THRESHOLD", "0.364"))

# =============================================================================
# Check required files
# =============================================================================

cpat_fasta_file <- file.path(cpat_dir, paste0(basename, "_cpat.ORF_seqs.fa"))
stopifnot("Input ORF FASTA file not found" = file.exists(cpat_fasta_file))

cpat_results_file <- file.path(cpat_dir, paste0(basename, "_cpat.ORF_prob.tsv"))
stopifnot("CPAT results file not found" = file.exists(cpat_results_file))

sample_full_fasta_path <- file.path(sqanti_dir, paste0(basename, "_corrected_filtered.fasta"))
stopifnot("Sample full fasta file not found" = file.exists(sample_full_fasta_path))

sample_gtf_path <- file.path(sqanti_dir, paste0(basename, "_corrected_filtered.gtf"))
stopifnot("Sample gtf file not found" = file.exists(sample_gtf_path))

classification_file <- paste0(sqanti_dir, "/", basename, "_classification_filtered.txt")
stopifnot("File not found" = file.exists(classification_file))

# =============================================================================
# Helper functions
# =============================================================================

#' Check if ORF ends with stop codon
#' @param orf_fasta_file Path to ORF FASTA file
#' @return Data frame with ID and has_stop_codon columns
check_stop_codons <- function(orf_fasta_file) {

  stop_codons <- c("TAG", "TAA", "TGA")
  orfs        <- readDNAStringSet(orf_fasta_file)
  
  tibble(
    ID       = str_split_fixed(names(orfs), "\t", 2)[,1],
    sequence = as.character(orfs)
  ) %>%
    mutate(seq_length     = nchar(sequence),
           last_codon     = if_else(seq_length >= 3, str_sub(sequence, -3, -1), ""),
           has_stop_codon = last_codon %in% stop_codons) %>%
    select(ID, has_stop_codon)
}

#' Map ORF coordinates to genomic positions
#' This is a simplified version that focuses on the core mapping logic
#' @param orf_coords ORF coordinate data
#' @param sample_gtf GTF data for sample transcripts
#' @param gencode_gtf GTF data for GENCODE reference
#' @param sample_fasta Sample transcript sequences
#' @return Mapped ORF data with genomic coordinates
map_orfs_to_genome <- function(orf_coords, sample_gtf, gencode_gtf, full_fasta) {

  # Extract exon information from sample GTF
  sample_exons = sample_gtf %>%
    filter(type == "exon") %>%
    select(transcript_id, seqnames, start, end, strand) %>%
    mutate(exon_length = end - start + 1) %>%
    arrange(transcript_id, if_else(strand == "+", start, -start))
  
  # Calculate cumulative exon positions for coordinate mapping
  sample_exons %<>%
    group_by(transcript_id) %>%
    mutate(cumulative_length = cumsum(exon_length), prior_length = lag(cumulative_length, default = 0)) %>%
    ungroup()
  
  # Map ORF coordinates to genomic positions and pull number of upstream ATGs- strandedness here is tricky
  mapped_orfs = orf_coords %>%
    left_join(sample_exons, by = c("isoform_id" = "transcript_id")) %>%
    filter(ORF_start > prior_length & ORF_start <= cumulative_length) %>% # which exon has the ORF start?
    left_join(full_fasta, by = "isoform_id") %>%
    mutate(genomic_start = ifelse(strand == "+",
                                  start + (ORF_start - prior_length) - 1,
                                  end - (ORF_start - prior_length) + 1)) %>%
    mutate(upstream_atgs = ifelse(!is.na(full_sequence), 
                                  str_count(substr(full_sequence, 1, ORF_start - 1), "ATG"),
                                  Inf)) %>%
    select(-full_sequence)
  
  # Check for GENCODE start codon matches- looking for any match here
  gencode_start_codons = gencode_gtf %>%
    filter(type == "start_codon") %>% 
    mutate(gencode_pos = ifelse(strand == "+", start, end)) %>%
    select(seqnames, strand, gencode_pos) %>%
    distinct() %>%
    mutate(gencode_match = TRUE)
  
  mapped_orfs %<>% left_join(gencode_start_codons, by = c("seqnames", "strand", "genomic_start" = "gencode_pos")) %>%
    mutate(gencode_match = ifelse(is.na(gencode_match), FALSE, gencode_match))
  
  return(mapped_orfs)
}

#' Select best ORF for each transcript
#' Logic: Prefer GENCODE matches with fewest upstream ATGs, otherwise highest coding score
#' @param mapped_orfs Data frame of mapped ORFs
#' @return Data frame with best ORF per transcript
call_best_orfs <- function(mapped_orfs) {

  # ORF quality scoring parameters but these don't seem to be used
  coding_threshold  # CPAT protein-coding threshold
  atg_shift  <- 10          # Sigmoid parameters for ATG penalty
  atg_growth <- 0.5
  
  # per transcript, if there is a gencode match, pick the one with the fewest upstream ATGs
  mapped_orfs %<>% 
    mutate(atg_penalty     = 1 - 1/(1 + exp(-atg_growth * (upstream_atgs - atg_shift))), 
           composite_score = Coding_prob * (1 - atg_penalty),
           orf_quality     = ifelse(!has_stop_codon, "No Stop Codon", 
                                    ifelse(Coding_prob <= coding_threshold, "Low Quality ORF", "Plausible ORF")))
  
  best_orfs = mapped_orfs %>% 
    filter(orf_quality == "Plausible ORF") %>%
    group_by(isoform_id) %>% 
    arrange(desc(gencode_match), upstream_atgs, desc(Coding_prob)) %>%
    slice_head(n = 1) %>%
    ungroup()
  
  mapped_orfs %<>%
    mutate(orf_quality = ifelse(ID %in% best_orfs$ID, "Clear Best ORF", orf_quality))
  
  return(mapped_orfs)
}

#' Translate ORF sequences to proteins and group by identical sequences
#' @param orfs Data frame with ORF coordinates
#' @param transcript_seqs Named vector of transcript sequences
#' @return Data frame with protein sequences and grouped transcript IDs
group_by_protein_sequence <- function(filtered_results, full_fasta) {

  orf_proteins = filtered_results %>%
    select(isoform_id, ORF_start, ORF_end) %>%
    left_join(full_fasta, by = "isoform_id") %>%
    filter(!is.na(full_sequence)) %>%
    mutate(orf_dna_sequence = substr(full_sequence, ORF_start, ORF_end))
  
  # Vectorized translation for speed
  dna      <- DNAStringSet(orf_proteins$orf_dna_sequence)
  proteins <- translate(dna, if.fuzzy.codon = "solve")
  
  orf_proteins %<>%
    mutate(orf_protein_sequence = as.character(proteins)) %>%
    filter(orf_protein_sequence != "") %>%
    select(isoform_id, orf_protein_sequence)
  
  # Group transcripts by identical protein sequences
  orf_groups = orf_proteins %>%
    group_by(orf_protein_sequence) %>%
    arrange(isoform_id) %>%
    mutate(
      orf_isoform_id = paste(isoform_id, collapse = " | "),
      orf_base_id    = isoform_id[1]  # First transcript as representative
    ) %>% 
    ungroup()
  
  return(orf_groups)
}

# =============================================================================
# Main code broken down into steps
# =============================================================================

# === STEP 1: Load all input data ===
message("\n--- STEP 1: Loading input data ---")
  
# CPAT results
cpat_results = read_tsv(cpat_results_file, show_col_types = FALSE) %>%
  separate(ID, into = c("isoform_id", "orf_rank"), sep = "_ORF_", remove = FALSE)

# Load GTF files
sample_gtf  = import(sample_gtf_path) %>% as.data.frame()
gencode_gtf = import(gencode_gtf_path) %>% as.data.frame()
  
# Load sample fasta with full sequence- include UTRs, filtered transcripts from SQANTI
full = readDNAStringSet(sample_full_fasta_path)
full_fasta = tibble(
  isoform_id    = names(full),
  full_sequence = as.character(full)
)

# SQANTI classification
classification = read_tsv(classification_file, show_col_types = FALSE)
  
# === STEP 2: Map ORFs to genomic coordinates ===
message("\n--- STEP 2: Mapping ORFs to genome ---")

stop_codon_status = check_stop_codons(cpat_fasta_file) # uses ORF (dna) fasta to return new column about stop codon 
orf_coords        = left_join(cpat_results, stop_codon_status, by = "ID")
mapped_orfs       = map_orfs_to_genome(orf_coords, sample_gtf, gencode_gtf, full_fasta)
  
# === STEP 3: Define the clear best ORF and print out all classified ORFs ===
message("\n--- STEP 3: Calling best ORFs and filtering---")
  
mapped_orfs_classified = call_best_orfs(mapped_orfs)
mapped_orfs_classified %<>% left_join(select(classification, 
                                            isoform_id = isoform, 
                                            structural_category, 
                                            gene_id = associated_gene, 
                                            transcript_id = associated_transcript,
                                            gene_name, 
                                            transcript_name))

original_n_transcripts = n_distinct(mapped_orfs_classified$isoform_id)

best_orfs = mapped_orfs_classified %>% 
  filter(orf_quality == "Clear Best ORF")

n_best_orfs = nrow(best_orfs %>% distinct(isoform_id))

write_tsv(mapped_orfs_classified, file.path(cpat_dir, paste0(basename, "_all_cpat_orfs_mapped.tsv")))

# === STEP 4: Group transcripts by ORF protein sequence ===
message("\n--- Grouping by ORF protein sequence ---")
orf_groups = group_by_protein_sequence(best_orfs, full_fasta)
best_orfs %<>% left_join(orf_groups)

write_tsv(best_orfs, file.path(cpat_dir, paste0(basename, "_best_cpat_orfs_mapped.tsv")))


```

Checking strand here
```{r}
# testing
mapped_orfs %>% filter(gencode_match == TRUE) %>% group_by(strand) %>% tally()

# negative stranded gene
sample_exons %>% filter(grepl("^PB\\.27\\.", transcript_id)) %>% arrange(desc(end))
test_neg = orf_coords %>% filter(ID == "PB.27.42_ORF_1") %>%
    left_join(sample_exons, by = c("isoform_id" = "transcript_id")) %>%
    filter(ORF_start > prior_length & ORF_start <= cumulative_length) %>%
    left_join(full_fasta, by = "isoform_id") %>%
    mutate(genomic_start = ifelse(strand == "+",
                                  start + (ORF_start - prior_length) - 1,
                                  end - (ORF_start - prior_length) + 1)) %>%
    mutate(upstream_atgs = ifelse(!is.na(full_sequence), 
                                  str_count(substr(full_sequence, 1, ORF_start - 1), "ATG"),
                                  Inf)) %>%
    select(-full_sequence)

gencode_test = gencode_gtf %>%
    filter(type == "start_codon") %>% 
    mutate(gencode_pos = ifelse(strand == "+", start, end)) %>%
    select(gene_id, transcript_id, seqnames, strand, gencode_pos) %>%
    distinct()

gencode_test %>% filter(strand == "-", gencode_pos == 959240)
gencode_gtf %>% filter(gene_id == "ENSG00000188976.11", type == "transcript")

# positive stranded gene
orf_coords %>% filter(ID == "PB.100.1_ORF_1") %>%
    left_join(sample_exons, by = c("isoform_id" = "transcript_id")) %>%
    filter(ORF_start > prior_length & ORF_start <= cumulative_length) %>%
    left_join(full_fasta, by = "isoform_id") %>%
    mutate(genomic_start = ifelse(strand == "+",
                                  start + (ORF_start - prior_length) - 1,
                                  end - (ORF_start - prior_length) + 1)) %>%
    mutate(upstream_atgs = ifelse(!is.na(full_sequence), 
                                  str_count(substr(full_sequence, 1, ORF_start - 1), "ATG"),
                                  Inf)) %>%
    select(-full_sequence)
```

Refine ORFs
```{r}
#!/usr/bin/env Rscript

# ORF Database Refinement Script - R Version
# Groups transcripts with identical protein sequences and filters by quality

suppressPackageStartupMessages({
  library(tidyverse)
  library(Biostrings)
  library(argparse)
})


#' Translate ORF sequences to proteins and group by identical sequences
#' @param orfs Data frame with ORF coordinates
#' @param transcript_seqs Named vector of transcript sequences
#' @return Data frame with protein sequences and grouped transcript IDs
group_by_protein_sequence <- function(filtered_results, full_fasta) {
  message("Translating ORFs and grouping by protein sequence...")
  
  orf_proteins = filtered_results %>%
    select(isoform_id, ORF_start, ORF_end) %>%
    left_join(full_fasta, by = "isoform_id") %>%
    filter(!is.na(full_sequence)) %>%
    mutate(orf_dna_sequence = substr(full_sequence, ORF_start, ORF_end))
    
  # Vectorized translation for speed
  dna <- DNAStringSet(orf_proteins$orf_dna_sequence)
  proteins <- translate(dna, if.fuzzy.codon = "solve")
  
  orf_proteins %<>%
    mutate(orf_protein_sequence = as.character(proteins)) %>%
    filter(orf_protein_sequence != "") %>%
    select(isoform_id, orf_protein_sequence)
  
  
  # Group transcripts by identical protein sequences
  orf_groups = orf_proteins %>%
    group_by(orf_protein_sequence) %>%
    arrange(isoform_id) %>%
    mutate(
      orf_isoform_id = paste(isoform_id, collapse = "|"),
      orf_base_id = isoform_id[1]  # First transcript as representative
    ) %>% 
    ungroup()
  
  return(orf_groups)
}


#' Calculate total expression for grouped transcripts
#' @param protein_groups Data frame with protein groups
#' @param orfs Data frame with expression data
#' @return Data frame with aggregated expression values
aggregate_expression <- function(protein_groups, orfs) {
  message("Aggregating expression values for protein groups...")
  
  # Create lookup tables for FL and CPM values
  fl_lookup <- setNames(orfs$FL, orfs$pb_acc)
  cpm_lookup <- setNames(orfs$CPM, orfs$pb_acc)
  
  # Aggregate expression for each protein group
  expression_summary <- protein_groups %>%
    mutate(
      FL = map_dbl(pb_accs, ~sum(fl_lookup[.x], na.rm = TRUE)),
      CPM = map_dbl(pb_accs, ~sum(cmp_lookup[.x], na.rm = TRUE))
    )
  
  return(expression_summary)
}

#' Main analysis function
main <- function() {

  message("=== Starting ORF Database Refinement ===")

  # === STEP 1: Load input data ===
  message("\n--- Loading input files ---")
  orfs <- read_tsv(args$orfs, show_col_types = FALSE)
  message("Loaded ", nrow(orfs), " ORFs")
  
  all_ids <- unique(orfs$pb_acc)
  message("Total unique transcripts: ", length(all_ids))
  
  # === STEP 2: Filter ORFs by quality ===
  message("\n--- Filtering ORFs by quality ---")
  # Apply quality filters
  filtered_results = best_orfs %>% filter(orf_quality %in% c("Plausible ORF", "Clear Best ORF"))
  
  # === STEP 4: Group transcripts by protein sequence ===
  message("\n--- Grouping by protein sequence ---")
  orf_groups <- group_by_protein_sequence(filtered_results, full_fasta)
  
  
  
  
  # === STEP 5: Aggregate expression data ===
  message("\n--- Aggregating expression data ---")
  protein_groups_with_expr <- aggregate_expression(protein_groups, kept_orfs)
  
  # === STEP 6: Merge with ORF metadata ===
  message("\n--- Merging with ORF metadata ---")
  
  # Get metadata for representative (base) transcripts
  orf_metadata <- kept_orfs %>%
    select(pb_acc, coding_score, orf_score, orf_calling_confidence, upstream_atgs, gene) %>%
    distinct()
  
  final_results <- protein_groups_with_expr %>%
    left_join(orf_metadata, by = c("base_acc" = "pb_acc")) %>%
    select(pb_accs_str, base_acc, protein_sequence, coding_score, orf_calling_confidence,
           upstream_atgs, orf_score, gene, FL, CPM) %>%
    rename(pb_accs = pb_accs_str)
  
  # === STEP 7: Write final outputs ===
  message("\n--- Writing output files ---")
  
  # Write refined protein FASTA
  refined_fasta <- paste0(args$name, "_orf_refined.fasta")
  fasta_lines <- paste0(">pb|", final_results$base_acc, "|fullname GN=", final_results$gene, "\n",
                       final_results$protein_sequence)
  writeLines(fasta_lines, refined_fasta)
  
  # Write refined metadata TSV
  refined_tsv <- paste0(args$name, "_orf_refined.tsv")
  write_tsv(final_results, refined_tsv)
  
  # === SUMMARY ===
  message("\n=== Refinement Complete ===")
  message("Final outputs:")
  message("- Refined FASTA: ", refined_fasta)
  message("- Refined metadata: ", refined_tsv)
  message("- Dropout FASTA: dropout_", basename(args$name), "_orf.fasta")
  message("- Dropout metadata: dropout_", basename(args$name), "_orf.tsv")
  message("")
  message("Summary:")
  message("- Input ORFs: ", nrow(orfs))
  message("- Quality filtered: ", nrow(kept_orfs))
  message("- Final protein groups: ", nrow(final_results))
  message("- Redundancy reduction: ", round((1 - nrow(final_results)/nrow(kept_orfs)) * 100, 1), "%")
}

# Run main function
if (!interactive()) {
  main()
}
```

Problem is that we have redundancy...some transcripts have more than one probably ORF
```{r}
#' Translate ORF sequences to proteins and group by identical sequences
#' @param orfs Data frame with ORF coordinates
#' @param transcript_seqs Named vector of transcript sequences
#' @return Data frame with protein sequences and grouped transcript IDs
group_by_protein_sequence <- function(filtered_results, full_fasta) {
  message("Translating ORFs and grouping by protein sequence...")
  
  orf_proteins = filtered_results %>%
    select(isoform_id, ORF_start, ORF_end) %>%
    left_join(full_fasta, by = "isoform_id") %>%
    filter(!is.na(full_sequence)) %>%
    mutate(orf_dna_sequence = substr(full_sequence, ORF_start, ORF_end))
    
  # Vectorized translation for speed
  dna <- DNAStringSet(orf_proteins$orf_dna_sequence)
  proteins <- translate(dna, if.fuzzy.codon = "solve")
  
  orf_proteins %<>%
    mutate(orf_protein_sequence = as.character(proteins)) %>%
    filter(orf_protein_sequence != "") %>%
    select(isoform_id, orf_protein_sequence)
  
  
  # Group transcripts by identical protein sequences
  orf_groups = orf_proteins %>%
    group_by(orf_protein_sequence) %>%
    arrange(isoform_id) %>%
    mutate(
      orf_isoform_id = paste(isoform_id, collapse = "|"),
      orf_base_id = isoform_id[1]  # First transcript as representative
    ) %>% 
    ungroup()
  
  return(orf_groups)
}






```


Compare SQANTI and SQANTI protein output
```{r}
trans = read_tsv("/home/cwp5au/sheynkman/projects/LRP_AML/03_filter_sqanti/filtered_AML_classification.txt")
protein = read_tsv("/home/cwp5au/sheynkman/projects/LRP_AML/allAMLvNBM/09_sqanti_protein/AML.sqanti_protein_classification.tsv")

trans %>% filter(associated_gene == "ENSG00000116954.8")
protein %>% filter(tx_gene == "ENSG00000116954.8")

```

Annotate cds and utr based on ORF
```{r}
# Sample GTF
sample_gtf_path = "~/megan/github/LRP2_lite/results/sqanti/sample_corrected_filtered.gtf"
orfs_file = "results/orf_calling/sample_best_cpat_orfs_mapped.tsv"

gtf = import(sample_gtf_path)

#!/usr/bin/env Rscript

#' GTF CDS Feature Addition Script
#' 
#' Creates GTF file that includes ORF regions as CDS features
#' Input: PacBio GTF, ORF calls, pb_gene mapping
#' Output: PacBio GTF with added CDS features (ORFs)

# =============================================================================
# Load required libraries
# =============================================================================

suppressPackageStartupMessages({
  library(tidyverse)
  library(rtracklayer)
  library(magrittr)
})

# =============================================================================
# Get environment variables and check required files
# =============================================================================

basename         <- Sys.getenv("OUTPUT_BASE_NAME")
sqanti_dir       <- file.path(Sys.getenv("OUTPUT_DIR"), "sqanti") 
cpat_dir         <- file.path(Sys.getenv("OUTPUT_DIR"), "orf_calling") 

sample_gtf_path  <- file.path(sqanti_dir, paste0(basename, "_corrected_filtered.gtf"))
orfs_file        <- file.path(cpat_dir, paste0(basename, "_best_cpat_orfs_mapped.tsv"))

stopifnot("Sample GTF file not found" = file.exists(sample_gtf_path))
stopifnot("Best ORFs file not found" = file.exists(orfs_file))

# =============================================================================
# Helper functions
# =============================================================================

#' Find which exon block contains an ORF coordinate
#' @param orf_coord Position in transcript coordinates
#' @param cumulative_lens Cumulative exon lengths
#' @return List with block index and delta from end
get_first_block_index <- function(orf_coord, cumulative_lens) {
  for (i in seq_along(cumulative_lens)) {
    if (orf_coord <= cumulative_lens[i]) {
      delta <- cumulative_lens[i] - orf_coord
      return(list(index = i, delta = delta))
    }
  }
  return(NULL)
}

#' Trim coordinates to ORF range (positive strand)
#' @param i1 Start block index
#' @param delta1 Distance from end of start block
#' @param i2 End block index
#' @param delta2 Distance from end of end block
#' @param coords_df Data frame with start/end coordinates
#' @return Trimmed coordinates
trim_coords_positive <- function(i1, delta1, i2, delta2, coords_df) {
  orf_coords <- coords_df %>%
    slice(i1:i2) %>%
    mutate(
      start = if_else(row_number() == 1, end - delta1, start),
      end = if_else(row_number() == n(), end - delta2, end)
    )
  return(orf_coords)
}

#' Trim coordinates to ORF range (negative strand)
#' @param i1 Start block index
#' @param delta1 Distance from end of start block
#' @param i2 End block index
#' @param delta2 Distance from end of end block
#' @param coords_df Data frame with start/end coordinates
#' @return Trimmed coordinates
trim_coords_negative <- function(i1, delta1, i2, delta2, coords_df) {
  orf_coords <- coords_df %>%
    slice(i1:i2) %>%
    mutate(
      start = if_else(row_number() == n(), start + delta2, start),
      end = if_else(row_number() == 1, start + delta1, end)
    )
  return(orf_coords)
}

#' Generate GTF with CDS features
#' @param sample_gtf GTF file path
#' @param agg_orfs_file Aggregated ORFs file path
#' @param refined_orfs_file Refined ORFs file path
#' @param pb_gene_file PacBio to gene mapping file path
#' @param output_cds Output GTF file path
#' @return NULL (writes output file)
make_pacbio_cds_gtf <- function(sample_gtf, agg_orfs_file, refined_orfs_file, 
                                pb_gene_file, output_cds) {
  
  # === STEP 1: Load GTF and extract exons ===
  message("\n--- STEP 1: Reading GTF file ---")
  
  gtf <- import(sample_gtf) %>%
    as.data.frame() %>%
    filter(type == "exon") %>%
    select(seqnames, start, end, strand, transcript_id) %>%
    rename(chr = seqnames, acc = transcript_id)
  
  message(paste0("Loaded ", nrow(gtf), " exon features"))
  
  # === STEP 2: Load ORF aggregation data ===
  message("\n--- STEP 2: Reading ORF aggregation data ---")
  
  agg_orfs <- read_tsv(agg_orfs_file, show_col_types = FALSE)
  
  # Create mappings from base accession
  base_acc_to_cpm <- setNames(agg_orfs$CPM, agg_orfs$base_acc)
  base_acc_to_transcripts <- setNames(agg_orfs$pb_accs, agg_orfs$base_acc)
  
  # Filter to representative transcripts only
  repr_pbs <- agg_orfs$base_acc
  gtf %<>% filter(acc %in% repr_pbs)
  
  message(paste0("Filtered to ", n_distinct(gtf$acc), " representative transcripts"))
  
  # === STEP 3: Organize coordinates by transcript ===
  message("\n--- STEP 3: Processing transcript coordinates ---")
  
  pbs <- gtf %>%
    group_by(acc) %>%
    arrange(acc, if_else(strand == "+", start, -start)) %>%
    mutate(
      block_len = end - start + 1,
      cum_block_len = cumsum(block_len)
    ) %>%
    summarise(
      chr = first(chr),
      strand = first(strand),
      coords = list(tibble(start = start, end = end)),
      block_lens = list(block_len),
      cum_block_lens = list(cum_block_len),
      .groups = "drop"
    )
  
  # === STEP 4: Load ORF ranges and gene mappings ===
  message("\n--- STEP 4: Loading ORF coordinates and gene mappings ---")
  
  ranges <- read_tsv(refined_orfs_file, show_col_types = FALSE) %>%
    select(pb_acc, orf_start, orf_end)
  
  pb_to_gene <- read_tsv(pb_gene_file, show_col_types = FALSE)
  pb_gene_map <- setNames(pb_to_gene$gene, pb_to_gene$pb_acc)
  
  message(paste0("Processing ", nrow(ranges), " ORF calls"))
  
  # === STEP 5: Map ORFs to genomic coordinates and write output ===
  message("\n--- STEP 5: Mapping ORFs and writing output GTF ---")
  
  output_lines <- list()
  n_processed <- 0
  n_skipped <- 0
  
  for (i in seq_len(nrow(ranges))) {
    acc <- ranges$pb_acc[i]
    orf_start <- ranges$orf_start[i]
    orf_end <- ranges$orf_end[i]
    
    # Check if transcript exists and has gene mapping
    if (!(acc %in% pbs$acc)) {
      n_skipped <- n_skipped + 1
      next
    }
    
    gene <- pb_gene_map[acc]
    if (is.na(gene) || gene == "-") {
      n_skipped <- n_skipped + 1
      next
    }
    
    # Get transcript information
    pb_info <- pbs %>% filter(acc == !!acc)
    chr <- pb_info$chr
    strand <- pb_info$strand
    coords <- pb_info$coords[[1]]
    cum_lens <- pb_info$cum_block_lens[[1]]
    
    # Get CPM and collapsed transcript names
    cpm <- base_acc_to_cpm[acc]
    if (is.na(cpm)) cpm <- 0
    
    all_transcripts <- base_acc_to_transcripts[acc]
    if (is.na(all_transcripts)) all_transcripts <- acc
    all_transcripts <- str_replace_all(all_transcripts, "\\|", ",")
    
    # Truncate long names
    max_name_length <- 50
    if (nchar(all_transcripts) > max_name_length) {
      transcript_list <- str_split(all_transcripts, ",")[[1]]
      all_transcripts <- paste0(transcript_list[1], "_plus_", 
                                length(transcript_list) - 1, "_others")
    }
    
    # Find ORF blocks
    idx1 <- get_first_block_index(orf_start, cum_lens)
    idx2 <- get_first_block_index(orf_end, cum_lens)
    
    if (is.null(idx1) || is.null(idx2)) {
      n_skipped <- n_skipped + 1
      next
    }
    
    # Trim coordinates to ORF range
    if (strand == "+") {
      orf_coords <- trim_coords_positive(idx1$index, idx1$delta, 
                                         idx2$index, idx2$delta, coords)
    } else {
      orf_coords <- trim_coords_negative(idx1$index, idx1$delta, 
                                         idx2$index, idx2$delta, coords)
    }
    
    # Create transcript ID with gene, transcripts, and CPM
    acc_w_gene_w_cpm <- paste0(gene, "|", all_transcripts, "|", cpm)
    attributes <- paste0('gene_id "', gene, '"; transcript_id "', 
                        acc_w_gene_w_cpm, '";')
    
    # Write exon lines
    for (j in seq_len(nrow(coords))) {
      line <- paste(chr, "hg38", "exon", coords$start[j], coords$end[j], 
                   ".", strand, ".", attributes, sep = "\t")
      output_lines <- c(output_lines, line)
    }
    
    # Write CDS lines
    for (j in seq_len(nrow(orf_coords))) {
      line <- paste(chr, "hg38", "CDS", orf_coords$start[j], orf_coords$end[j], 
                   ".", strand, ".", attributes, sep = "\t")
      output_lines <- c(output_lines, line)
    }
    
    n_processed <- n_processed + 1
  }
  
  # Write output file
  writeLines(unlist(output_lines), output_cds)
  
  # === Summary ===
  message("\n=== PROCESSING COMPLETE ===")
  message(paste0("Successfully processed: ", n_processed, " transcripts"))
  message(paste0("Skipped (no gene mapping or not found): ", n_skipped, " transcripts"))
  message(paste0("Output written to: ", output_cds))
  message(paste0("Total GTF lines written: ", length(output_lines)))
}

# =============================================================================
# Main execution
# =============================================================================

message("\n=== GTF CDS FEATURE ADDITION ===")
message(paste0("Sample GTF: ", sample_gtf_path))
message(paste0("Aggregated ORFs: ", agg_orfs_file))
message(paste0("Refined ORFs: ", refined_orfs_file))
message(paste0("PacBio-Gene mapping: ", pb_gene_file))
message(paste0("Output: ", output_cds_path))

make_pacbio_cds_gtf(
  sample_gtf_path,
  agg_orfs_file,
  refined_orfs_file,
  pb_gene_file,
  output_cds_path
)

message("\nScript completed successfully!")

```


```{r}
all_cds_exons %>% filter(new_start > new_end) # PB.10040.43

#keep = best_orfs
best_orfs = keep
best_orfs %>% filter(strand == "+")


get_cds_coords <- function(best_orfs, sample_exons) {
  
  # Join mapped ORFs with their exons- select all CDS exons
  cds_all = best_orfs %>%
    left_join(sample_exons, by = c("isoform_id", "seqnames", "strand")) %>%
    filter(exon_number >= orf_start_exon & exon_number <= orf_end_exon) %>% # get all CDS exons
    group_by(isoform_id) %>%
    arrange(isoform_id, exon_number) %>%
    mutate(
      is_first_orf_exon = (exon_number == min(exon_number)),
      is_last_orf_exon = (exon_number == max(exon_number))
    ) %>%
    ungroup()
  
  # Trim first and last cds exons based on precise cds start and end
  cds_trimmed = cds_all %>%
    mutate(
      new_start = case_when(
        strand == "+" & is_first_orf_exon ~ start + orf_start_offset - 1,
        strand == "-" & is_last_orf_exon ~ end - orf_end_offset + 1,
        TRUE ~ start
      ),
      new_end = case_when(
        strand == "+" & is_last_orf_exon ~ start + orf_end_offset - 1,
        strand == "-" & is_first_orf_exon ~ end - orf_start_offset + 1,
        TRUE ~ end
      )
    ) %>%
    select(ID, isoform_id, orf_isoform_id, seqnames, exon_number, new_start, new_end, strand)
  
  return(cds_trimmed)
}

cds_trimmed %>% filter(new_start == new_end) %>% left_join(gene_mapping)
all_cds_exons = get_cds_coords(best_orfs, sample_exons)
all_cds_exons

# PB.27.11 (-)
gencode_gtf %>% filter(type == "CDS", transcript_name == "NOC2L-201")

# PB.30.2 (+)
gencode_gtf %>% filter(type == "CDS", transcript_name == "ISG15-203")
gencode_gtf %>% filter(transcript_name == "ISG15-203")

# PB.10453.20
mapped_orfs_classified %>% filter(isoform_id == "PB.10453.48")
cds_all %>% filter(isoform_id == "PB.10453.48")
gencode_gtf %>% filter(type == "CDS", transcript_name == "SELENOT-206")


# TNFAIP8-201, PB.15079.14, PB.15079.29, PB.15079.30
gencode_gtf %>% filter(type == "CDS", transcript_name == "TNFAIP8-201")
cds_trimmed %>% filter(isoform_id == "PB.15079.14")
```


Create a single function for CDS mapping- replace
```{r}
sample_gtf_path = "~/megan/github/LRP2_lite/results/sqanti/sample_corrected_filtered.gtf"
sample_gtf  = import(sample_gtf_path) %>% as.data.frame()

gencode_gtf = import("/project/sheynkman/external_data/GENCODE_v47/gencode.v47.annotation.gtf") %>% as.data.frame()

cpat_results_file = "/home/cwp5au/megan/github/LRP2_lite/results/orf_calling/sample_cpat.ORF_prob.tsv"
orf_coords = read_tsv(cpat_results_file, show_col_types = FALSE) %>%
  separate(ID, into = c("isoform_id", "orf_rank"), sep = "_ORF_", remove = FALSE)

map_orfs_to_genome <- function(orf_coords, sample_gtf, gencode_gtf, full_fasta) {
  
  # Extract exon information from sample GTF
  sample_exons = sample_gtf %>%
    filter(type == "exon") %>%
    select(isoform_id = transcript_id, seqnames, start, end, strand) %>%
    mutate(exon_length = end - start + 1) %>%
    arrange(isoform_id, if_else(strand == "+", start, -start))
  
  # Calculate cumulative exon positions for coordinate mapping
  sample_exons %<>%
    group_by(isoform_id) %>%
    mutate(
      exon_number = row_number(),
      cumulative_length = cumsum(exon_length), 
      prior_length = lag(cumulative_length, default = 0)
    ) %>%
    ungroup()
  
  # Count the number of upstream ORFs
  orf_coords %<>% left_join(full_fasta, by = "isoform_id") %>%
    mutate(upstream_atgs = ifelse(!is.na(full_sequence), 
                                  str_count(substr(full_sequence, 1, ORF_start - 1), "ATG"),
                                  Inf)
    ) %>%
    select(-full_sequence)
  
  # Map ORF Start coordinates to genomic positions- strandedness here is tricky
  mapped_starts = orf_coords %>%
    left_join(sample_exons, by = c("isoform_id")) %>%
    filter(ORF_start > prior_length & ORF_start <= cumulative_length) %>% # which exon has the ORF start?
    mutate(orf_genomic_start = ifelse(strand == "+",
                                  start + (ORF_start - prior_length) - 1,
                                  end - (ORF_start - prior_length) + 1),
           orf_start_exon   = exon_number,
           orf_start_offset = ORF_start - prior_length  # position within this exon
    ) %>%
    select(-start, -end, -exon_length, -exon_number, -cumulative_length, -prior_length)
  
  # Map ORF End coordinates to genomic positions
  mapped_complete = mapped_starts %>%
    left_join(sample_exons, by = c("isoform_id", "seqnames", "strand")) %>%
    filter(ORF_end > prior_length & ORF_end <= cumulative_length) %>% # which exon has the ORF end?
    mutate(
      orf_genomic_end = if_else(strand == "+",
                                 start + (ORF_end - prior_length) - 1,
                                 end - (ORF_end - prior_length) + 1),
      orf_end_exon   = exon_number,
      orf_end_offset = ORF_end - prior_length  # position within this exon
    ) %>%
    select(-start, -end, -exon_length, -exon_number, -cumulative_length, -prior_length)
  
  # Check for GENCODE start codon matches- looking for any match here
  gencode_start_codons = gencode_gtf %>%
    filter(type == "start_codon") %>% 
    mutate(gencode_pos = ifelse(strand == "+", start, end)) %>%
    select(seqnames, strand, gencode_pos) %>%
    distinct() %>%
    mutate(gencode_match = TRUE)
  
  mapped_orfs = mapped_complete %>% 
    left_join(gencode_start_codons, by = c("seqnames", "strand", "orf_genomic_start" = "gencode_pos")) %>%
    mutate(gencode_match = ifelse(is.na(gencode_match), FALSE, gencode_match))
  
  return(mapped_orfs)
}


```

Next function to trim CDS to add to gtf- do this after filtering for best ORF? 
- yes because otherwise, will have multiple ORFs per transcript in the gtf
- and this is how it was in the OG pipeline
```{r}
test = best_orfs

#' Get CDS coordinates by trimming exons to ORF boundaries
#' Uses the exon mapping information to extract only CDS regions
#' @param best_orfs Each row has an ORF with mapping info
#' @param sample_gtf Info on all exons
#' @return Data frame with CDS start/end coordinates
get_cds_coords <- function(best_orfs, sample_exons) {
  
  # Join mapped ORFs with their exons- select all CDS exons
  cds_all = best_orfs %>%
    left_join(sample_exons, by = c("isoform_id", "seqnames", "strand")) %>%
    filter(exon_number >= orf_start_exon & exon_number <= orf_end_exon) %>% # get all CDS exons
    group_by(isoform_id) %>%
    arrange(isoform_id, exon_number) %>%
    mutate(
      is_first_orf_exon = (exon_number == min(exon_number)),
      is_last_orf_exon = (exon_number == max(exon_number))
    ) %>%
    ungroup()
  
  # Trim first and last cds exons based on precise cds start and end
  cds_trimmed = cds_all %>%
    mutate(
      new_start = case_when(
        strand == "+" & is_first_orf_exon ~ start + orf_start_offset - 1,
        strand == "-" & is_last_orf_exon ~ start + orf_end_offset - 1,
        TRUE ~ start
      ),
      new_end = case_when(
        strand == "+" & is_last_orf_exon ~ start + orf_end_offset - 1,
        strand == "-" & is_first_orf_exon ~ end - orf_start_offset + 1,
        TRUE ~ end
      )
    ) %>%
    select(ID, isoform_id, seqnames, exon_number, new_start, new_end, strand)
  
  return(cds_trimmed)
}

get_cds_coords(best_orfs, sample_exons)


gencode_gtf %>% group_by(type) %>% tally()
sample_gtf %>% group_by(type) %>% tally()

```

Make the gtf- dont think this needs to be a function
- can we output a gtf of all transcripts and then one that is aggregated at the ORF level with ORF id?


```{r}
basename = "sample"
gene_mapping
all_cds_exons
sample_gtf

mapped_orfs_classified %>% filter(isoform_id == "PB.100.1")


sample_gtf_path = "~/megan/github/LRP2_lite/results/sqanti/sample_corrected_filtered.gtf"
gencode_gtf = import("/project/sheynkman/external_data/GENCODE_v47/gencode.v47.annotation.gtf") %>% as.data.frame()

sample_gtf  = import(sample_gtf_path) %>% as.data.frame()

cpat_dir = "/home/cwp5au/megan/github/LRP2_lite/results/orf_calling"
cpat_fasta_file = "/home/cwp5au/megan/github/LRP2_lite/results/orf_calling/sample_cpat.ORF_seqs.fa"
cpat_results_file = "/home/cwp5au/megan/github/LRP2_lite/results/orf_calling/sample_cpat.ORF_prob.tsv"
sample_full_fasta_path = "/home/cwp5au/megan/github/LRP2_lite/results/sqanti/sample_corrected_filtered.fasta"
classification_file = "/home/cwp5au/megan/github/LRP2_lite/results/sqanti/sample_classification_filtered.txt"


write_gtf_with_cds <- function(sample_gtf, all_cds_exons, gene_mapping) {
  
  # selecting type individually in case custom gtfs have more type columns- want to reset attributes column
  sample_gtf %<>% 
    select(isoform_id = transcript_id, everything()) %>%
    left_join(gene_mapping %>% select(isoform_id, gencode_gene_id = gene_id, gene_name), by = "isoform_id")
  
  # Create transcript lines (one per transcript)
  transcript_lines = sample_gtf %>%
    filter(type == "transcript") %>%
    select(seqnames, type, start, end, strand, source, isoform_id, gencode_gene_id, gene_name)
  
  # Create exon lines
  exon_lines = sample_gtf %>%
    filter(type == "exon") %>%
    select(seqnames, type, start, end, strand, source, isoform_id, gencode_gene_id, gene_name)
  
  # Create CDS lines
  cds_lines = all_cds_exons %>%
    left_join(gene_mapping %>% select(isoform_id, gencode_gene_id = gene_id, gene_name), by = "isoform_id") %>%
    mutate(type = "CDS", source = sample_gtf$source[1]) %>%
    select(seqnames, type, start = new_start, end = new_end, strand, source, isoform_id, gencode_gene_id, gene_name)
  
  # combine types and add attribute column
  updated_gtf = bind_rows(transcript_lines, exon_lines, cds_lines) %>%
    arrange(isoform_id, start) %>%
    mutate(
      attributes = paste0('gene_id "', gencode_gene_id, '"; transcript_id "', isoform_id, '"; gene_name "', gene_name, '";'),
      score = ".", 
      frame = "."
    ) %>%
    select(seqnames, source, type, start, end, score, strand, frame, attributes)

  message(paste0("Combined ", nrow(transcript_lines), " transcript lines, ",
                 nrow(exon_lines), " exon lines, and ", 
                 nrow(cds_lines), " CDS lines"))
  
  return(updated_gtf)
}




```

Compare SQANTI transcript and protein classification output for AML samples
```{r}
transcript = read_tsv("/home/cwp5au/sheynkman/projects/LRP_AML/03_filter_sqanti/filtered_AML_classification.txt")
transcript %>% group_by(structural_category) %>% tally()
transcript %>% group_by(subcategory) %>% tally()
  
protein = read_tsv("/home/cwp5au/sheynkman/projects/LRP_AML/allAMLvNBM/09_sqanti_protein/AML.sqanti_protein_classification.tsv")

transcript %>% filter(str_detect(isoform, "PB\\.1000\\.")) %>% arrange(isoform)
protein %>% filter(str_detect(pb, "PB\\.1000\\.")) %>% arrange(pb)

transcript %>% filter(isoform == "PB.1000.25")
protein %>% filter(pb == "PB.1000.25")


protein_class = protein %>% select(isoform = pb, tx_cat, pr_splice_cat, tx_subcat, pr_splice_subcat, tx_gene, pr_gene, tx_transcripts, pr_transcripts)
transcript_class = transcript %>% select(isoform, structural_category, associated_gene, associated_transcript, subcategory)

df = protein_class %>% left_join(transcript_class)

# original transcript calls versus current
all(df$tx_cat == df$structural_category)
df %>% filter(tx_cat != structural_category) # fusions, 89 dont match

all(df$tx_subcat == df$subcategory)
df %>% filter(tx_subcat != subcategory) # 14,861


# transcript versus protein
all(df$tx_cat == df$pr_splice_cat)
df %>% filter(tx_cat != pr_splice_cat)

all(df$tx_subcat == df$pr_splice_subcat)
df %>% filter(tx_subcat != pr_splice_subcat)

# match to gencode
all(df$tx_gene == df$associated_gene)
df %>% filter(tx_gene != associated_gene) # 104, includes a lot of fusions

# look at a fusion
transcript %>% filter(isoform == "PB.34191.15")
protein %>% filter(pb == "PB.34191.15")


protein %>% filter(is_nmd == TRUE)
transcript %>% group_by(subcategory) %>% tally() # 97,481
protein %>% group_by(tx_subcat) %>% tally() # 32,029
protein %>% group_by(pr_splice_subcat) %>% tally()


protein %>% filter(pb == "PB.1000.25")

protein %>% filter(tx_cat == "full-splice_match") %>% group_by(pr_splice_cat) %>% tally()
protein %>% filter(tx_cat == "incomplete-splice_match") %>% group_by(pr_splice_cat) %>% tally()
protein %>% filter(tx_cat == "novel_in_catalog") %>% group_by(pr_splice_cat) %>% tally()
protein %>% filter(tx_cat == "novel_not_in_catalog") %>% group_by(pr_splice_cat) %>% tally()


```


5'UTR code
```{r}
# AML test
gencode_gtf_path = "/project/sheynkman/external_data/GENCODE_v47/gencode.v47.annotation.gtf"
sample_cds_gtf_path = "/project/sheynkman/users/megan/github/LRP2_lite/results/orf_calling/sample_corrected_filtered_CDS.gtf"
protein_sqanti_path = "/project/sheynkman/users/megan/github/LRP2_lite/results/protein_sqanti/sample.sqanti_protein_classification.tsv"
basename = "sample"  

# jurkat test
gencode_gtf_path = "~/datasets/jurkat_zenodo5234651/gencode.v35.annotation.chr22.gtf"
sample_cds_gtf_path = "~/datasets/jurkat_zenodo5234651/jurkat_chr22_with_cds.gtf"
protein_sqanti_path = "/project/sheynkman/users/megan/github/LRP2_lite/results/protein_sqanti/jurkat_chr22.sqanti_protein_classification.tsv"
basename = "jurkat" 

#!/usr/bin/env Rscript

#' Protein Classification Script
#' 
#' Part 1: Classifies Long read derived 5' UTRs as either:
#'   - "subset" (N-terminal truncation, using internal start codon)
#'   - "unique" (novel start site, extending into new sequence)
#'   
#' Part 2: Protein Classification
#' Integrates SQANTI-protein output to classify CDS structures:
#'   - pFSM (Full Protein Match): CDS junctions exactly match reference
#'   - pISM (Incomplete Protein Match): CDS junctions are subset of reference
#'   - pNIC (Novel In Catalog): Novel combination of known N/splice/C
#'   - pNNC (Novel Not in Catalog): At least one novel N/splice/C

# =============================================================================
# Load libraries
# =============================================================================

suppressPackageStartupMessages({
  library(GenomicRanges)
  library(txdbmaker)
  library(Biostrings)
  library(tidyverse)
  library(magrittr)
  library(rtracklayer)
})

# =============================================================================
# Get environment variables and check required files
# =============================================================================

basename           <- Sys.getenv("OUTPUT_BASE_NAME")
cpat_dir           <- file.path(Sys.getenv("OUTPUT_DIR"), "orf_calling") 
protein_sqanti_dir <- file.path(Sys.getenv("OUTPUT_DIR"), "protein_sqanti") 
gencode_gtf_path   <- Sys.getenv("GENCODE_GTF_FILE")

sample_cds_gtf_path   <- file.path(cpat_dir, paste0(basename, "_corrected_filtered_CDS.gtf"))
protein_sqanti_path   <- file.path(protein_sqanti_dir, paste0(basename, ".sqanti_protein_classification.tsv"))

stopifnot("GENCODE gtf file not found" = file.exists(gencode_gtf_path))
stopifnot("Sample gtf file not found" = file.exists(sample_cds_gtf_path))
stopifnot("Protein classification file not found" = file.exists(protein_sqanti_path))

# =============================================================================
# Helper Functions
# =============================================================================

#' Classify multiexonic 5' UTR based on junction chain
#' @param tss Integer: TSS coordinate
#' @param strand Character: '+' or '-'
#' @param junc_chain Character: junction chain string
#' @param gc_chains Character vector: GENCODE junction chains for this gene
#' @return Character: "perfect_subset", "known_protruding", or "novel"
classify_multiexonic_utr = function(tss, strand, junc_chain, gc_chains) {
  if (is.na(tss) || junc_chain == "" || length(gc_chains) == 0) return("novel")

  # Split the concatenated chains
  gc_chains = str_split(gc_chains, "\\|\\|\\|")[[1]]
  
  status = "novel"
  
  for (gc_chain in gc_chains) {
    if (!str_detect(gc_chain, fixed(junc_chain))) next
    
    # Junction chain matches - check if TSS protrudes beyond reference
    if (strand == "+") {
      prefix  = str_split(gc_chain, fixed(junc_chain))[[1]][1]
      coords  = str_split(prefix, "_|-")[[1]][1]
      gc_5end = as.integer(coords[length(coords)])
      
      if (is.na(gc_5end)) next
      
      if ((tss + 9) < gc_5end) {
        status = "known_protruding"
      } else {
        return("perfect_subset")
      }
    } else {
      suffix  = str_split(gc_chain, fixed(junc_chain))[[1]][2]
      coords  = str_split(suffix, "_|-")[[1]][1]
      gc_5end = as.integer(coords[2])
      
      if (is.na(gc_5end)) next
      
      if ((tss - 9) > gc_5end) {
        status = "known_protruding"
      } else {
        return("perfect_subset")
      }
    }
  }
  
  return(status)
}

# =============================================================================
# STEP 1: Process GENCODE reference
# =============================================================================

message("\n--- STEP 1: Processing GENCODE reference ---")

# Load GENCODE GTF
gencode_gtf = import(gencode_gtf_path) %>% as.data.frame()

# Identify protein-coding transcripts (have CDS)
pc_transcripts = gencode_gtf %>%
  filter(type == "CDS") %>%
  pull(transcript_name) %>%
  unique()

# Extract exons and create GRanges for overlap checking
gencode_exons = gencode_gtf %>%
  filter(type == "exon", transcript_name %in% pc_transcripts) %>%
  select(seqnames, start, end, gene_id, transcript_name, strand)

gencode_gr = gencode_exons %>%
  select(seqnames, start, end, strand) %>%
  makeGRangesFromDataFrame()

gencode_gr = GenomicRanges::reduce(gencode_gr) # Reduces overlapping exons into merged genomic intervals

# Create exon chain strings per gene for junction matching (could potentially be tweaked to mimic collapse logic)
gencode_chains = gencode_exons %>%
  arrange(gene_id, transcript_name, start) %>%
  group_by(gene_id, transcript_name) %>%
  summarise(
    exon_chain = paste(paste0(start, "-", end), collapse = "_"),
    .groups = "drop_last"  # Keep gene grouping!
  ) %>%
  summarise(chains = paste(exon_chain, collapse = "|||"), .groups = "drop") 

# =============================================================================
# STEP 2: Process long read transcripts
# =============================================================================

message("\n--- STEP 2: Processing Sample transcripts ---")

sample_gtf = import(sample_cds_gtf_path) %>% as.data.frame()

# Create TxDb and extract 5' UTRs directly
txdb      = makeTxDbFromGFF(sample_cds_gtf_path, format = "gtf")
utr_by_tx = fiveUTRsByTranscript(txdb, use.names = TRUE)

# Get transcript metadata (gene name, chromosome, strand)
tx_metadata = sample_gtf %>%
  filter(type == "transcript") %>%
  select(transcript_id, gene_id, seqnames, strand)

# Get CDS Nterm and TSS info 
sample_cds = sample_gtf %>%
  filter(type == "CDS") %>%
  group_by(transcript_id) %>%
  summarise(
    nterm = ifelse(first(strand) == "+", min(start), max(end)),
    .groups = "drop"
  )

sample_exons = sample_gtf %>%
  filter(type == "exon") %>%
  group_by(transcript_id) %>%
  summarise(
    tss = ifelse(first(strand) == "+", min(start), max(end)),
    .groups = "drop"
  )

# Convert 5' UTR GRangesList to tibble with stats
utr_stats = tibble(
  transcript_id    = names(utr_by_tx),
  num_5utr_exons   = lengths(utr_by_tx),
  utr_length       = sum(width(utr_by_tx))
)

# Create chain strings for 5' UTRs
utr_chains = tibble(
  transcript_id = names(utr_by_tx),
  utr_chain = sapply(utr_by_tx, function(gr) {
    if (length(gr) == 0) return("")
    gr = sort(gr)
    paste(paste0(start(gr), "-", end(gr)), collapse = "_")
  })
) %>%
  mutate(
    junc_chain = case_when(
      utr_chain == "" ~ "",
      !str_detect(utr_chain, "_") ~ "",
      TRUE ~ {
        coords = str_extract_all(utr_chain, "\\d+")
        sapply(coords, function(x) {
          if (length(x) <= 2) return("")
          junctions = as.numeric(x[2:(length(x)-1)])
          paste(junctions, collapse = "_")
        })
      }
    )
  )

# Combine all info
utr_info = tx_metadata %>%
  left_join(sample_exons, by = "transcript_id") %>%
  left_join(sample_cds, by = "transcript_id") %>%
  left_join(utr_stats, by = "transcript_id") %>%
  left_join(utr_chains, by = "transcript_id") %>%
  mutate(
    # Handle transcripts with no UTR
    num_5utr_exons = replace_na(num_5utr_exons, 0),
    utr_length = replace_na(utr_length, 0),
    utr_exon_status = case_when(
      num_5utr_exons == 0 ~ "no_utr",
      num_5utr_exons == 1 ~ "monoexonic",
      TRUE ~ "multiexonic"
    )
  )

# =============================================================================
# STEP 3: Classify 5' UTRs
# =============================================================================

message("\n--- STEP 3: Classifying 5' UTRs ---")

# Check TSS overlap with GENCODE using GRanges 
tss_gr = GRanges(
  seqnames = utr_info$seqnames,
  ranges   = IRanges(start = utr_info$tss, end = utr_info$tss),
  strand   = utr_info$strand
)

overlaps = findOverlaps(tss_gr, gencode_gr)
utr_info$tss_in_gc_exons = seq_len(nrow(utr_info)) %in% queryHits(overlaps)

# For monoexonic: calculate distance to nearest GENCODE exon 
distances       = rep(NA_real_, nrow(utr_info))
non_overlapping = which(!utr_info$tss_in_gc_exons & utr_info$utr_exon_status == "monoexonic")

if (length(non_overlapping) > 0) {
  nearest                    = distanceToNearest(tss_gr[non_overlapping], gencode_gr)
  distances[non_overlapping] = mcols(nearest)$distance
}

utr_info$distance_to_gc = distances

# Join GENCODE chains for multiexonic comparison
utr_info = utr_info %>%
  left_join(gencode_chains, by = "gene_id")

# Classify monoexonic UTRs
utr_info = utr_info %>%
  mutate(
    mono_class = case_when(
      utr_exon_status != "monoexonic" ~ NA_character_,
      tss_in_gc_exons ~ "within",
      distance_to_gc >= 10 ~ "protruding",
      TRUE ~ "within"
    )
  )

# Classify multiexonic UTRs (needs rowwise for junction matching)
multi_idx = which(utr_info$utr_exon_status == "multiexonic")

if (length(multi_idx) > 0) {
  multi_results = utr_info[multi_idx, ] %>%
    rowwise() %>%
    mutate(
      multi_class = classify_multiexonic_utr(tss, strand, junc_chain, chains)
    ) %>%
    ungroup() %>%
    pull(multi_class)
  
  utr_info$multi_class            = NA_character_
  utr_info$multi_class[multi_idx] = multi_results
}

# Final classification 
utr_results = utr_info %>%
  mutate(
    junc_cat = case_when(
      is.na(nterm) ~ "no_orf",
      utr_exon_status == "no_utr" ~ "no_utr",
      utr_exon_status == "monoexonic" ~ mono_class,
      utr_exon_status == "multiexonic" ~ multi_class,
      TRUE ~ "unknown"
    ),
    
    # Map to final categories
    utr_cat = case_when(
      junc_cat %in% c("no_orf", "no_utr", "unknown") ~ junc_cat,
      junc_cat %in% c("within", "perfect_subset") ~ "subset",
      junc_cat %in% c("protruding", "known_protruding", "novel") ~ "unique",
      TRUE ~ "unknown"
    )
  ) %>%
  select(isoform_id = transcript_id, num_5utr_exons, utr_exon_status, tss_in_gc_exons, junc_cat, utr_cat)

# Merge with SQANTI protein classification
sqanti_class = read_tsv(protein_sqanti_path, show_col_types = FALSE)

# Adjust C term differences since CPAT includes stop codon in coords and gencode does not
sqanti_class %<>% mutate(
  pr_cterm_diff = ifelse(!is.na(pr_cterm_diff), pr_cterm_diff + 3, pr_cterm_diff),
  pr_cterm_gene_diff = ifelse(!is.na(pr_cterm_gene_diff), pr_cterm_gene_diff - 3, pr_cterm_gene_diff),
  pr_chang = ifelse(!is.na(pr_chang), pr_chang - 3, pr_chang))

utr_output = sqanti_class %>%
  full_join(utr_results, by = "isoform_id")

# write out intermediate file for troubleshooting
#write_tsv(utr_output, paste0("../results/protein_sqanti/", basename, "_5utr_classification.tsv"))


# 5' UTR summary statistics
message("\n--- 5'UTR Classification complete! ---")
message(sprintf("Total transcripts: %d", nrow(utr_output)))
message(sprintf("  subset: %d", sum(utr_output$utr_cat == "subset", na.rm = TRUE)))
message(sprintf("  unique: %d", sum(utr_output$utr_cat == "unique", na.rm = TRUE)))
message(sprintf("  no_orf: %d", sum(utr_output$utr_cat == "no_orf", na.rm = TRUE)))
message(sprintf("  no_utr: %d", sum(utr_output$utr_cat == "no_utr", na.rm = TRUE)))


# =============================================================================
# STEP 4: Protein classification
# =============================================================================

message("\n--- STEP 4: Protein classification ---")

protein_classifications = utr_output %>%
  mutate(
    protein_classification = case_when(
      
      ## Compared to the specific best-matching reference isoform
      # pr_nterm_diff: difference in ORF start positions (ATG)
      # pr_cterm_diff: difference in ORF end positions (stop codon)
      
      ## Compared to the nearest position across ALL isoforms of that gene
      # pr_nterm_gene_diff: distance to nearest CDS start among all isoforms
      # pr_cterm_gene_diff: distance to nearest CDS end among all isoforms
      
      # -----------------------------------------------------------------------------------------------------------
      # SQANTI_Protein, Full Splice Match - CDS splice junctions match exactly (known_splice), but precise N and C term may differ 
      # -----------------------------------------------------------------------------------------------------------
      
      # Case 1: Both ATG and stop codon match reference isoform exactly
      # FPM = Full protein match
      pr_splice_cat == "full-splice_match" & pr_splice_subcat == "multi-exon" & 
        pr_nterm_diff == 0 & pr_cterm_diff == 0 ~ 
        "pFPM,known_nterm_known_splice_known_cterm",
      
      # Case 2: Novel combination - ATG and/or STOP differ from reference isoform, but both exist in other isoforms of the gene
      pr_splice_cat == "full-splice_match" & pr_splice_subcat == "multi-exon" & 
        (pr_nterm_diff != 0 | pr_cterm_diff != 0) & pr_nterm_gene_diff == 0 & pr_cterm_gene_diff == 0 ~ 
        "pNIC,combo_nterm_cterm",
      
      # Case 3: ATG matches the reference exactly, but STOP codon is novel
      pr_splice_cat == "full-splice_match" & pr_splice_subcat == "multi-exon" & 
        pr_nterm_diff == 0 & pr_cterm_diff != 0 ~ 
        "pNNC,known_nterm_known_splice_novel_cterm",
      
      # Case 4 subtypes: ATG doesn't match any isoform (Implicit: pr_nterm_gene_diff != 0), known STOP
      
      # 4a: Protein extends upstream - novel N-terminal amino acids
      pr_splice_cat == "full-splice_match" & pr_splice_subcat == "multi-exon" & 
        pr_nterm_diff != 0 & pr_cterm_diff == 0 & pr_nhang > 0 ~ 
        "pNNC,novel_nterm_known_splice_known_cterm",
      
      # 4b: ATG is downstream/within reference CDS, using internal start codon = N-terminal truncation
      # IPM = Incomplete Protein Match
      pr_splice_cat == "full-splice_match" & pr_splice_subcat == "multi-exon" & 
        pr_nterm_diff != 0 & pr_cterm_diff == 0 & pr_nhang <= 0 & utr_cat == "subset" ~ 
        "pIPM,nterm_truncation",
      
      # 4c: Novel TSS context - ATG position is novel but protein doesn't extend
      # Transcript has extended 5' UTR (novel TSS) creating novel regulatory context
      pr_splice_cat == "full-splice_match" & pr_splice_subcat == "multi-exon" & 
        pr_nterm_diff != 0 & pr_cterm_diff == 0 & pr_nhang <= 0 & utr_cat == "unique" ~ 
        "pNNC,novel_nterm_known_splice_known_cterm",
      
      # Case 5 subtypes: Both ATG and stop codon differ from reference isoform
      
      # 5a: ATG matches another isoform (known), stop is novel
      pr_splice_cat == "full-splice_match" & pr_splice_subcat == "multi-exon" & 
        pr_nterm_diff != 0 & pr_cterm_diff != 0 & pr_nterm_gene_diff == 0 ~ 
        "pNNC,known_nterm_known_splice_novel_cterm",
      
      # 5b: ATG doesn't match any isoform + internal start = truncation
      pr_splice_cat == "full-splice_match" & pr_splice_subcat == "multi-exon" & 
        pr_nterm_diff != 0 & pr_cterm_diff != 0 & pr_nterm_gene_diff != 0 & utr_cat == "subset" ~ 
        "pIPM,nterm_truncation",
      
      # 5c: Novel ATG (protein or TSS context) but end matches known isoform
      pr_splice_cat == "full-splice_match" & pr_splice_subcat == "multi-exon" & 
        pr_nterm_diff != 0 & pr_cterm_diff != 0 & pr_nterm_gene_diff != 0 & 
        utr_cat == "unique" & pr_cterm_gene_diff == 0 ~ 
        "pNNC,novel_nterm_known_splice_known_cterm",
      
      # 5d: Novel ATG and novel STOP
      pr_splice_cat == "full-splice_match" & pr_splice_subcat == "multi-exon" & 
        pr_nterm_diff != 0 & pr_cterm_diff != 0 & pr_nterm_gene_diff != 0 & 
        utr_cat == "unique" & pr_cterm_gene_diff != 0 ~ 
        "pNNC,novel_nterm_known_splice_novel_cterm",
      
      # FSM orphan catch-all
      pr_splice_cat == "full-splice_match" & pr_splice_subcat == "multi-exon" ~ 
        "orphan_fsm",
      
      # ---------------------------------------------------------------------------------------
      # SQANTI_Protein, Incomplete Splice Match (ISM) - CDS junctions are a subset of reference
      # (e.g., 5' fragment, 3' fragment, intron retention in CDS)
      # ---------------------------------------------------------------------------------------
      
      # Case 1: ATG matches reference isoform, stop codon is novel
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff == 0 & pr_cterm_diff != 0 & pr_cterm_gene_diff != 0 ~ 
        "pNNC,known_nterm_known_splice_novel_cterm",
      
      # Case 2: Novel combination - ATG matches reference isoform, stop codon matches another isoform
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff == 0 & pr_cterm_diff != 0 & pr_cterm_gene_diff == 0 ~ 
        "pNIC,combo_nterm_cterm",
      
      # Case 3: Novel combination - Both ATG and stop codon match (incomplete CDS splicing only)
      # Likely intron retention within CDS causing unique splicing pattern
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff == 0 & pr_cterm_diff == 0 ~ 
        "pNIC,known_nterm_combo_splice_known_cterm",
      
      # Case 4: Novel combination - ATG differs from reference but matches another isoform, stop codon matches
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff != 0 & pr_cterm_diff == 0 & pr_nterm_gene_diff == 0 ~ 
        "pNIC,combo_nterm_cterm",
      
      # Case 5 subtypes - ATG doesnt match any isoform (novel), stop codon matches
      
      # Case 5a: ATG is downstream (internal start codon) - N-terminal truncation
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff != 0 & pr_cterm_diff == 0 & pr_nterm_gene_diff != 0 & 
        pr_nhang <= 0 & utr_cat == "subset" ~ 
        "pIPM,nterm_truncation",
      
      # Case 5b: Novel TSS context - ATG position novel but protein doesn't extend
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff != 0 & pr_cterm_diff == 0 & pr_nterm_gene_diff != 0 & 
        pr_nhang <= 0 & utr_cat == "unique" ~ 
        "pNNC,novel_nterm_known_splice_known_cterm",
      
      # Case 5c: Protein extends upstream - novel N-terminal amino acids
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff != 0 & pr_cterm_diff == 0 & pr_nterm_gene_diff != 0 & pr_nhang > 0 ~ 
        "pNNC,novel_nterm_known_splice_known_cterm",
      
      # Case 6: Novel Combination - Both ATG and stop match other isoforms
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff != 0 & pr_cterm_diff != 0 & pr_nterm_gene_diff == 0 & pr_cterm_gene_diff == 0 ~ 
        "pNIC,combo_nterm_cterm",
      
      # Case 7: ATG matches another isoform, stop codon is novel
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff != 0 & pr_cterm_diff != 0 & pr_nterm_gene_diff == 0 & pr_cterm_gene_diff != 0 ~ 
        "pNNC,known_nterm_known_splice_novel_cterm",
      
      # Case 8 subtypes - ATG doesn't match any isoform, stop codon varies
      
      # Case 8a: Internal start codon - truncation
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff != 0 & pr_cterm_diff != 0 & pr_nterm_gene_diff != 0 & utr_cat == "subset" ~ 
        "pIPM,nterm_truncation",
      
      # Case 8b: Novel ATG (protein or TSS context), stop matches another isoform
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff != 0 & pr_cterm_diff != 0 & pr_nterm_gene_diff != 0 & 
        utr_cat == "unique" & pr_cterm_gene_diff == 0 ~ 
        "pNNC,novel_nterm_known_splice_known_cterm",
      
      # Case 8c: Both ATG and stop codon are novel (don't match any isoform)
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_diff != 0 & pr_cterm_diff != 0 & pr_nterm_gene_diff != 0 & 
        utr_cat == "unique" & pr_cterm_gene_diff != 0 ~ 
        "pNNC,novel_nterm_known_splice_novel_cterm",
      
      # ISM orphan catch-all
      pr_splice_cat == "incomplete-splice_match" & !str_detect(pr_splice_subcat, "mono-exon") ~ 
        "orphan_ism",
      
      # ------------------------------------------------------------------------------------------
      # SQANTI_Protein, Novel in Catalog (NIC) - combo_splice (no reference isoform to compare to)
      # ------------------------------------------------------------------------------------------
      
      # Case 1: Both ATG and stop codon match an annotated isoform
      pr_splice_cat == "novel_in_catalog" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_gene_diff == 0 & pr_cterm_gene_diff == 0 ~ 
        "pNIC,known_nterm_combo_splice_known_cterm",
      
      # Case 2: ATG matches an isoform, stop codon is novel
      pr_splice_cat == "novel_in_catalog" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_gene_diff == 0 & pr_cterm_gene_diff != 0 ~ 
        "pNNC,known_nterm_combo_splice_novel_cterm",
      
      # Case 3: ATG doesn't match - truncation
      pr_splice_cat == "novel_in_catalog" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_gene_diff != 0 & utr_cat == "subset" ~ 
        "pIPM,nterm_truncation",
      
      # Case 4: Novel ATG (protein or TSS context), stop codon matches
      pr_splice_cat == "novel_in_catalog" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_gene_diff != 0 & utr_cat == "unique" & pr_cterm_gene_diff == 0 ~ 
        "pNNC,novel_nterm_combo_splice_known_cterm",
      
      # Case 5: Both ATG and stop codon are novel
      pr_splice_cat == "novel_in_catalog" & !str_detect(pr_splice_subcat, "mono-exon") & 
        pr_nterm_gene_diff != 0 & utr_cat == "unique" & pr_cterm_gene_diff != 0 ~ 
        "pNNC,novel_nterm_combo_splice_novel_cterm",
      
      # NIC orphan catch-all
      pr_splice_cat == "novel_in_catalog" & !str_detect(pr_splice_subcat, "mono-exon") ~ 
        "orphan_nic",
      
      # ----------------------------------------------------------------------------------------------
      # SQANTI_Protein, Novel Not in Catalog (NNC) - novel_splice (no reference isoform to compare to)
      # ----------------------------------------------------------------------------------------------
    
      # Case 1: Both ATG and stop codon match an annotated isoform
      pr_splice_cat == "novel_not_in_catalog" & pr_nterm_gene_diff == 0 & pr_cterm_gene_diff == 0 ~ 
        "pNNC,known_nterm_novel_splice_known_cterm",
      
      # Case 2: ATG matches an isoform, stop codon is novel
      pr_splice_cat == "novel_not_in_catalog" & pr_nterm_gene_diff == 0 & pr_cterm_gene_diff != 0 ~ 
        "pNNC,known_nterm_novel_splice_novel_cterm",
      
      # Case 3: ATG doesn't match - truncation
      pr_splice_cat == "novel_not_in_catalog" & pr_nterm_gene_diff != 0 & utr_cat == "subset" ~ 
        "pIPM,nterm_truncation",
      
      # Case 4: Novel ATG (protein or TSS context), stop codon matches
      pr_splice_cat == "novel_not_in_catalog" & pr_nterm_gene_diff != 0 & 
        utr_cat == "unique" & pr_cterm_gene_diff == 0 ~ 
        "pNNC,novel_nterm_novel_splice_known_cterm",
      
      # Case 5: Both ATG and stop codon are novel
      pr_splice_cat == "novel_not_in_catalog" & pr_nterm_gene_diff != 0 & 
        utr_cat == "unique" & pr_cterm_gene_diff != 0 ~ 
        "pNNC,novel_nterm_novel_splice_novel_cterm",
      
      # NNC orphan catch-all
      pr_splice_cat == "novel_not_in_catalog" ~ 
        "orphan_nnc",
      
      # ------------------------------
      # Mono-exon cases
      # ------------------------------
      
      # Perfect mono-exon match
      (pr_splice_subcat == "mono-exon" | pr_splice_subcat == "mono-exon_by_intron_retention") & 
        pr_splice_cat == "full-splice_match" & pr_nterm_diff == 0 & pr_cterm_diff == 0 ~ 
        "pFPM,mono-exon",
      
      # Intergenic mono-exon
      (pr_splice_subcat == "mono-exon" | pr_splice_subcat == "mono-exon_by_intron_retention") & 
        pr_splice_cat == "intergenic" ~ 
        "intergenic,mono-exon",
      
      # Other mono-exon cases (orphan)
      (pr_splice_subcat == "mono-exon" | pr_splice_subcat == "mono-exon_by_intron_retention") ~ 
        "orphan_monoexon,mono-exon",
      
      # ------------------------------
      # Miscellaneous multi-exon
      # ------------------------------
      
      pr_splice_subcat == "multi-exon" & pr_splice_cat == "intergenic" ~ 
        "intergenic,multi-exon",
      
      pr_splice_subcat == "multi-exon" & pr_splice_cat == "genic" ~ 
        "genic,multi-exon",
      
      pr_splice_subcat == "multi-exon" & pr_splice_cat == "antisense" ~ 
        "antisense,multi-exon",
      
      pr_splice_subcat == "multi-exon" & pr_splice_cat == "fusion" ~ 
        "fusion,multi-exon",
      
      # ------------------------------
      # Unclassified
      # ------------------------------
      
      TRUE ~ ""
    )
  ) %>%
  # Split protein_classification into base and subset
  separate(protein_classification, 
           into = c("protein_classification_base", "protein_classification_subset"), 
           sep = ",", 
           fill = "right", 
           remove = FALSE)

message("\n--- Protein classification complete! ---")
message("\nBy classification base:")
message(sprintf("  pFPM: %d", sum(pb_classifications$protein_classification_base == "pFPM", na.rm = TRUE)))
message(sprintf("  pIPM: %d", sum(pb_classifications$protein_classification_base == "pIPM", na.rm = TRUE)))
message(sprintf("  pNIC: %d", sum(pb_classifications$protein_classification_base == "pNIC", na.rm = TRUE)))
message(sprintf("  pNNC: %d", sum(pb_classifications$protein_classification_base == "pNNC", na.rm = TRUE)))
message(sprintf("  intergenic: %d", sum(pb_classifications$protein_classification_base == "intergenic", na.rm = TRUE)))
message(sprintf("  genic: %d", sum(pb_classifications$protein_classification_base == "genic", na.rm = TRUE)))
message(sprintf("  antisense: %d", sum(pb_classifications$protein_classification_base == "antisense", na.rm = TRUE)))
message(sprintf("  fusion: %d", sum(pb_classifications$protein_classification_base == "fusion", na.rm = TRUE)))
message(sprintf("  orphan: %d", sum(grepl("orphan", pb_classifications$protein_classification_base), na.rm = TRUE)))

message("\nAnalysis complete!")
```

```{r}
sample = read_tsv("~/megan/github/LRP2_lite/results/protein_sqanti/sample_5utr_classification.tsv")
sample %>% group_by(utr_exon_status, junc_cat) %>% tally()
sample %>% group_by(utr_cat) %>% tally()

# check ORF file for orf coords, incldues stop codon
orfs = read_tsv("~/megan/github/LRP2_lite/results/orf_calling/sample_best_orfs_mapped.tsv")


names(sqanti_class)
test = sqanti_class %>% mutate(
  pr_cterm_diff = ifelse(!is.na(pr_cterm_diff), pr_cterm_diff + 3, pr_cterm_diff),
  pr_cterm_gene_diff = ifelse(!is.na(pr_cterm_gene_diff), pr_cterm_gene_diff - 3, pr_cterm_gene_diff),
  pr_chang = ifelse(!is.na(pr_chang), pr_chang - 3, pr_chang))

test %>% group_by(pr_cterm_diff) %>% tally() %>% arrange(desc(n)) # 75710, equals 0 so no diff and 51276 NA (NIC and NNC)
test %>% group_by(pr_cterm_gene_diff) %>% tally() %>% arrange(desc(n)) # 99117, equals 0 so no diff and 650 NA
test %>% group_by(pr_chang) %>% tally() %>% arrange(desc(n)) #=
test

```


```{r}
# add fasta to orf calling module
best_orfs = read_tsv("../results/orf_calling/sample_best_orfs_mapped.tsv", show_col_types = FALSE)
best_orfs %>% filter(gene_id == "ENSG00000157191.20")
unique_orf = best_orfs %>% distinct(gene_id, gene_name, orf_isoform_id, orf_protein_sequence)
protein_seqs = AAStringSet(unique_orf$orf_protein_sequence)
names(protein_seqs) = paste0(unique_orf$orf_isoform_id, "|", unique_orf$gene_id, "|", unique_orf$gene_name)
writeXStringSet(protein_seqs, "../results/orf_calling/sample_")


```

What is the agreement between transcript and protein annotation?
Issue in the protein sqanti code
I think it selects one transcript even if multiple transcripts match
some duplication in gene names
```{r}
sqanti = read_tsv("~/megan/github/LRP2_lite/results/sqanti/sample_classification_filtered.txt")
sqanti %>% filter(associated_gene == "ENSG00000157191.20")

protein_classifications = read_tsv("~/megan/github/LRP2_lite/results/protein_sqanti/sample.sqanti_protein_classification.tsv") # 148,829
protein_classifications %>% filter(tx_gene != pr_gene) # 1,717
protein_classifications %>% filter(tx_gene == pr_gene)
protein_classifications %>% separate_rows(pr_gene, sep = ",")
protein_classifications %>% separate_rows(tx_gene, sep = ",")
protein_classifications %>% filter(grepl(",", tx_gene))
protein_classifications %>% filter(grepl(",", pr_gene))

# add section in protein classification script
protein_classifications = read_tsv("~/megan/github/LRP2_lite/results/protein_sqanti/sample_protein_classification.tsv")
gencode_gtf_path = "/project/sheynkman/external_data/GENCODE_v47/gencode.v47.annotation.gtf"
gencode_gtf = import(gencode_gtf_path) %>% as.data.frame()

message("\n--- STEP 5: Reconcile gene ids and names ---")

gene_mapping = gencode_gtf %>% distinct(gene_id, gene_name)
ensg_lookup  = setNames(gene_mapping$gene_name, gene_mapping$gene_id)

# Just do direct assignment - no functions at all
protein_classifications$tx_gene_name = ensg_lookup[protein_classifications$tx_gene]
protein_classifications$pr_gene_name = ensg_lookup[protein_classifications$pr_gene]

# For the columns with commas, manually fix them (this is much faster than ifelse)
has_comma_tx = which(grepl(",", protein_classifications$tx_gene))
has_comma_pr = which(grepl(",", protein_classifications$pr_gene))

message("Transcripts mapping to multiple genes at tx level: ", length(has_comma_tx))
message("Transcripts mapping to multiple genes at pr level: ", length(has_comma_pr))

for (i in has_comma_tx) {
  ensgs = strsplit(protein_classifications$tx_gene[i], ",")[[1]]
  protein_classifications$tx_gene_name[i] = paste(unique(ensg_lookup[ensgs]), collapse = ",")
}

for (i in has_comma_pr) {
  ensgs = strsplit(protein_classifications$pr_gene[i], ",")[[1]]
  protein_classifications$pr_gene_name[i] = paste(unique(ensg_lookup[ensgs]), collapse = ",")
}

# Reconciliation of gene ids
matching = protein_classifications$tx_gene == protein_classifications$pr_gene
protein_classifications$reconciled_gene_id = protein_classifications$pr_gene  
protein_classifications$reconciled_gene_name = protein_classifications$pr_gene_name

# Only need to reconcile non-matching rows
non_matching = which(!matching)

message("Rows that don't match and need reconciliation:", length(non_matching), "\n")

for (i in non_matching) {
  tx_ids  = strsplit(protein_classifications$tx_gene[i], ",")[[1]]
  pr_ids  = strsplit(protein_classifications$pr_gene[i], ",")[[1]]
  
  tx_name = strsplit(protein_classifications$tx_gene_name[i], ",")[[1]]
  pr_name = strsplit(protein_classifications$pr_gene_name[i], ",")[[1]]
  
  match_id   = intersect(tx_ids, pr_ids)
  match_name = intersect(tx_name, pr_name)
  
  if (length(match_id) > 0) {
    protein_classifications$reconciled_gene_id[i] = match_id[1]
  } else {
    protein_classifications$reconciled_gene_id[i] = pr_ids[1]
  }
  
  if (length(match_name) > 0) {
    protein_classifications$reconciled_gene_name[i] = match_name[1]
  } else {
    protein_classifications$reconciled_gene_name[i] = pr_name[1]
  }
}

# check examples- check na
protein_classifications %>% 
  filter(grepl(",", tx_gene)) %>%
  select(tx_gene, pr_gene, tx_gene_name, pr_gene_name, reconciled_gene_id, reconciled_gene_name)

protein_classifications %>% 
  filter(grepl(",", pr_gene)) %>%
  select(tx_gene, pr_gene, tx_gene_name, pr_gene_name, reconciled_gene_id, reconciled_gene_name)

```

Filtering SQANTI protein
```{r}
min_junctions_after_stop_codon = 0
pclass_base_to_keep = c('pFPM', 'pNIC')
sample_cds_fasta_path = "~/megan/github/LRP2_lite/results/orf_calling/sample_best_orfs_collapsed.fa"


# Subclass lookup for short names, important for browser tracks
subclass_lookup = c(
  'known_nterm_novel_splice_known_cterm' = 'kn_ns_kc',
  'known_nterm_known_splice_known_cterm' = 'kn_ks_kc',
  'known_nterm_combo_splice_known_cterm' = 'kn_cs_kc',
  'known_nterm_novel_splice_novel_cterm' = 'kn_ns_nc',
  'known_nterm_combo_splice_novel_cterm' = 'kn_cs_nc',
  'known_nterm_known_splice_novel_cterm' = 'kn_ks_nc',
  'novel_nterm_known_splice_known_cterm' = 'nn_ks_kc',
  'novel_nterm_known_splice_novel_cterm' = 'nn_ks_nc',
  'novel_nterm_novel_splice_known_cterm' = 'nn_ns_kc',
  'novel_nterm_combo_splice_novel_cterm' = 'nn_cs_nc',
  'novel_nterm_combo_splice_known_cterm' = 'nn_cs_kc',
  'novel_nterm_novel_splice_novel_cterm' = 'nn_ns_nc',
  'combo_nterm_cterm'   = 'combo',
  'nterm_truncation'    = 'ntrunc',
  'multi-exon'          = 'multi',
  'mono-exon'           = 'mono'
)

full_protein = protein_classifications %>%
  filter(!is.na(protein_classification)) %>%
  mutate(
    filter_status = case_when(
      
      # Check 1: Filter if high NMD signal (applies to ALL)
      num_junc_after_stop_codon > min_junctions_after_stop_codon ~ "NMD",
      
      # Check 2: Filter if problematic patterns
      grepl('intergenic|antisense|fusion|orphan|genic', protein_classification) ~ "atypical",
      
      # Check 3: Filter if NOT in keep base classification list
      !(protein_classification_base %in% pclass_base_to_keep) ~ "pc_base",
      
      # Keep everything else
      TRUE ~ "keep"
    ),
    
    # abbreviate
    psubclass_short = subclass_lookup[protein_classification_subset]

  )

# Output full table with filter status
full_protein %<>%
  dplyr::select(isoform_id, tx_gene, pr_gene, reconciled_gene_id, reconciled_gene_name, num_junc_after_stop_codon, num_5utr_exons,
         tclass = tx_cat, pclass = protein_classification_base, psubclass = protein_classification_subset, psubclass_short, filter_status)
  
write_tsv(full_protein, file.path(protein_sqanti_dir, paste0(basename, "_protein_classification_all.tsv")))

# Output filtered text file
filtered_protein = full_protein %>% filter(filter_status == "keep")
write_tsv(filtered_protein, file.path(protein_sqanti_dir, paste0(basename, "_protein_high_confidence.txt")))

# Output filtered gtf 
gtf          = import(sample_cds_gtf_path)
gtf_filtered = gtf[gtf$transcript_id %in% filtered_protein$isoform_id]
gtf_meta     = as.data.frame(mcols(gtf_filtered)) # pull metadata

new_meta = filtered_protein %>% 
  mutate(transcript_id = paste(reconciled_gene_name, isoform_id, pclass, sep = "|")) %>%
  select(isoform_id, gene_id = reconciled_gene_id, gene_name = reconciled_gene_name, transcript_id)

updated_meta = gtf_meta %>% 
  select(source, type, score, phase, isoform_id = transcript_id, ORF_id) %>%
  left_join(new_meta, by = "isoform_id") %>%
  select(-ORF_id, everything(), ORF_id)

mcols(gtf_filtered) = updated_meta
export(gtf_filtered, file.path(protein_sqanti_dir, paste0(basename, "_protein_high_confidence.gtf")), format="gtf")

# Output filtered fasta, and adjust header pb|PB.1001.26|fullname GN=MYCBP
fa = readDNAStringSet(sample_cds_fasta_path)

filtered_fa_seqs = data.frame(
  header   = names(fa),
  sequence = as.character(fa),
  length   = width(fa),
  row.names = NULL
) %>%
  separate(header,
           into = c("isoform_id", "old_gene_id", "old_gene_name"),
           sep = "\\|") %>%
  separate_rows(isoform_id, sep = ",") %>%
  select(isoform_id, sequence, length) %>%
  distinct() %>%
  filter(isoform_id %in% filtered_protein$isoform_id)

# combine with reconciled naming, update orf ids post filtering
fasta_out = filtered_fa_seqs %>% 
  left_join(select(filtered_protein, isoform_id, reconciled_gene_id, reconciled_gene_name)) %>%
  group_by(sequence, length) %>%
  arrange(isoform_id) %>%
  summarize(orf_isoform_id = paste(isoform_id, collapse = ","), gene_name = paste(unique(reconciled_gene_name), collapse = ",")) %>%
  ungroup()

protein_seqs        = AAStringSet(fasta_out$sequence)
names(protein_seqs) = paste0(fasta_out$orf_isoform_id, "|", fasta_out$gene_name)
writeXStringSet(protein_seqs, file.path(protein_sqanti_dir, paste0(basename, "_protein_high_confidence.fa")))

```




```{r}


# testing
test %>% group_by(protein_classification_base, filter_status) %>% tally()
test %>% group_by(protein_classification_subset) %>% tally()

protein_classifications %>% filter(protein_classification_base == "pIPM") %>% arrange(num_junc_after_stop_codon)
protein_classifications %>% filter(num_junc_after_stop_codon > min_junctions_after_stop_codon) %>%
  group_by(protein_classification_base) %>% tally()

protein_classifications %>% filter(protein_classification_base == "pFPM" & num_junc_after_stop_codon > min_junctions_after_stop_codon)
full_protein %>% filter(is.na(protein_classification_subset))
full_protein %>% filter(is_nmd == FALSE) %>% group_by(num_junc_after_stop_codon) %>% tally()
full_protein %>% filter(filter_status == "NMD" & is_nmd == TRUE) # 25562

# numbers are off because of AA sequences mapping to different genes?
fasta_out
test = fa[keep_seqs]
data.frame(header = names(test), sequence = as.character(test), row.names = NULL) %>%
  group_by(sequence) %>% tally() %>% arrange(desc(n))
data.frame(header = names(test), sequence = as.character(test), row.names = NULL) %>% 
  filter(sequence == "MSGRGKGGKGGKGGAKRHRKVRDNGTKARRARRGGVKRSGYTRGVKVNVRDAVTYTHAKRKTVTAMDVVYAKRGRTYGGG")

```



Jurkat testing
```{r}
og_gtf = import("~/datasets/jurkat_zenodo5234651/jurkat_chr22.collapsed.gff")
gtf = import("../jurkat_test_results/pacbio_isoseq/jurkat_chr22_merged.collapsed.gff")
og_gtf %>% as.data.frame() %>% filter(type == "transcript") %>% arrange(start, end)
gtf %>% as.data.frame() %>% filter(type == "transcript") %>% arrange(start, end)


sqanti = read_tsv("../jurkat_test_results/sqanti/jurkat_chr22_classification.txt")

sample = read_tsv("../results/sqanti/sample_classification.txt")

```


```{r}
metadata      = fread(metadata_path)
sample_lookup = setNames(metadata$name, metadata$bam_id)

classification_file = "~/megan/github/LRP2_lite/results/sqanti_transcript/sample_classification.txt"
sqanti_df = read_tsv(classification_file, show_col_types = FALSE)

colnames(sqanti_df) = sub("^FL\\.", "", colnames(sqanti_df))
colnames(sqanti_df) = ifelse(colnames(sqanti_df) %in% names(sample_lookup),
                           paste0(sample_lookup[colnames(sqanti_df)], "_counts"),
                           colnames(sqanti_df))

counts_cols = grep("_counts$", colnames(sqanti_df), value = TRUE)
sqanti_df %<>%
  mutate(across(all_of(counts_cols), 
                ~ (.x / sum(.x)) * 1e6,
                .names = "{gsub('_counts', '_cpm', .col)}"))

sqanti_df_full = sqanti_df %>%
  filter(!is.na(associated_gene), str_starts(associated_gene, "ENS"))

sqanti_df_full %<>% 
  left_join(gencode_gene, by = "associated_gene") %>% 
  left_join(gencode_df, by = c("associated_gene", "associated_transcript", "gene_type", "gene_name"))

# combine with hashids, and selct tpm columns
hashids = read_tsv(mapping_file)

sqanti_ids = sqanti_df_full %>% 
  select(isoform_id = isoform, ensg_gene_id = associated_gene, gene_name, enst_transcript_id = associated_transcript, transcript_name, gene_type, protein_id, ends_with("_counts"), ends_with("_cpm"))

all_ids = hashids %>% select(isoform_id = transcript_id, hash_id) %>%
  inner_join(sqanti_ids, by = "isoform_id") 
all_ids %>% distinct(isoform_id)
all_ids %>% distinct(ensg_gene_id, hash_id)

write_tsv(all_ids, "~/megan/github/LRP2_lite/results/sqanti_transcript/sample_all_hashids_with_cpm.txt")

```




























