---
title: "LRP2 Lite Testing"
output: html_document
---

```{r}
source("/project/sheynkman/users/megan/github/megan-schertzer/ms_functions.R")

```

What does sqanti classification file have?
```{r}
all = read_tsv("../results/sqanti/sample_classification.txt")

gtf_file_path = "/project/sheynkman/external_data/GENCODE_v47/gencode.v47.annotation.gtf"

gr <- import(gtf_file_path, format = "gtf")
gtf_data <- as.data.table(gr)

reference_df = gtf_data %>% filter(type == "transcript") %>% select(gene_id, gene_type, gene_name, transcript_id, transcript_type, transcript_name, protein_id)
tail(reference_df)
```



sqanti_filter.R
```{r}
#!/usr/bin/env Rscript

#' SQANTI3 Transcript Filtering Script
#' 
#' R script to filter SQANTI3 transcripts based on quality criteria:
#' - Protein-coding genes only
#' - Remove transcripts with internal priming 
#' - Remove template switching artifacts
#' - Filter by SQANTI structural categories
#'
#' Outputs both filtered files and dropout files with reasons.

# =============================================================================
# Load required libraries
# =============================================================================

suppressPackageStartupMessages({
  library(tidyverse)
  library(rtracklayer)
  library(Biostrings) 
  library(argparse)
  library(data.table)
})

# =============================================================================
# Define SQANTI structural categories
# =============================================================================

STRUCTURAL_CATEGORIES <- list(
  strict = c("novel_not_in_catalog", "novel_in_catalog", 
             "incomplete-splice_match", "full-splice_match"),
  all = c("antisense", "novel_not_in_catalog", "novel_in_catalog",
          "incomplete-splice_match", "full-splice_match", "genic",
          "intergenic", "fusion", "genic_intron")
)

# =============================================================================
# Helper functions
# =============================================================================

#' Apply filter and track dropouts
#' @param df Data frame to filter
#' @param condition Logical condition for filtering
#' @param reason Reason for dropout
#' @param dropout_tracker List to store dropout reasons
#' @return Filtered data frame
apply_filter <- function(df, condition, reason, dropout_tracker) {
  pre_ids <- df$isoform
  df_filtered <- df[condition, ]
  post_ids <- df_filtered$isoform
  
  # Track dropped transcripts
  dropped <- setdiff(pre_ids, post_ids)
  dropout_tracker[[reason]] <- dropped
  
  message(paste0("Filter '", reason, "': dropped ", length(dropped), " transcripts"))
  return(df_filtered)
}

#' Save filtered sequences to FASTA
#' @param fasta_file Input FASTA file
#' @param keep_ids IDs to keep
#' @param output_file Output FASTA file
save_filtered_fasta <- function(fasta_file, keep_ids, output_file) {
  # Read FASTA
  sequences <- readDNAStringSet(fasta_file)
  
  # Filter sequences
  keep_mask <- names(sequences) %in% keep_ids
  filtered_sequences <- sequences[keep_mask]
  
  # Write filtered FASTA
  writeXStringSet(filtered_sequences, output_file)
  message(paste0("Saved ", length(filtered_sequences), " sequences to ", basename(output_file)))
}

#' Save filtered GTF
#' @param gtf_file Input GTF file
#' @param keep_ids Transcript IDs to keep
#' @param output_file Output GTF file
save_filtered_gtf <- function(gtf_file, keep_ids, output_file) {
  # Read GTF
  gtf <- import(gtf_file)
  
  # Filter by transcript_id
  filtered_gtf <- gtf[gtf$transcript_id %in% keep_ids]
  
  # Write filtered GTF
  export(filtered_gtf, output_file)
  message(paste0("Saved ", length(unique(filtered_gtf$transcript_id)), " transcripts to ", basename(output_file)))
}

# =============================================================================
# Command line arguments and environmental variables
# =============================================================================

parser <- ArgumentParser(description = "Filter SQANTI3 transcripts using environment variables")
  
# Default is to filter
parser$add_argument("--filter_protein_coding", default = TRUE, action = "store_true",
                    help = "Filter transcripts from protein-coding genes")
parser$add_argument("--filter_internal_priming", default = TRUE, action = "store_true",
                    help = "Filter transcripts that are the result of internal priming")
parser$add_argument("--filter_RTS", default = TRUE, action = "store_true",
                    help = "Filter template switching artifacts")
  
# Filtering parameters
parser$add_argument("--percent_polyA_threshold", default = 95, type = "double",
                    help = "Max percent A content downstream")
parser$add_argument("--structural_level", default = "strict",
                    choices = c("strict", "all"), help = "SQANTI Structural category level")
  
# Get environment variables set in pipeline config
basename <- Sys.getenv("OUTPUT_BASE_NAME")
dir <- file.path(Sys.getenv("OUTPUT_DIR"), "sqanti") 
gencode_gtf <- Sys.getenv("GENCODE_GTF_FILE")

# =============================================================================
# Check required files
# =============================================================================

classification_file <- paste0(dir, "/", basename, "_classification.txt")
stopifnot("File not found" = file.exists(classification_file))
  
sqanti_gtf <- paste0(dir, "/", basename, "_corrected.gtf")
stopifnot("File not found" = file.exists(sqanti_gtf))
  
sqanti_fasta <- paste0(dir, "/", basename, "_corrected.fasta")
stopifnot("File not found" = file.exists(sqanti_fasta))

# Create additional output directory
dropout_dir <- file.path(dir, "dropout")
dir.create(dropout_dir, recursive = TRUE, showWarnings = FALSE)

# =============================================================================
# Main code
# =============================================================================

message("Loading SQANTI3 classification data...")
  
# Load and process gencode gtf file
gencode <- import(gencode_gtf, format = "gtf")
gencode_df <- as.data.table(gencode)

gencode_df %<>% filter(type == "transcript") %>% select(gene_id, gene_type, gene_name, transcript_id, transcript_type, transcript_name, protein_id)
gencode_gene = gencode_df %>% select(associated_gene = gene_id, gene_type, gene_name) %>% distinct()
  
# Load and preprocess classification data, merge with gencode annotation info
sqanti_df <- read_tsv(classification_file, show_col_types = FALSE) %>%
  # Remove rows without associated genes and keep only ENS gene IDs
  filter(!is.na(associated_gene), 
          str_starts(associated_gene, "ENS"))
  
sqanti_df %<>% left_join(gencode_gene)
  
original_ids <- sqanti_df$isoform
message(paste0("Starting with ", length(original_ids), " transcripts"))
  
# Initialize dropout tracking
dropout_tracker <- list()
  
# Apply filters sequentially
if (args$filter_protein_coding) {
  message("Applying protein-coding gene filter...")
  sqanti_df <- apply_filter(
    sqanti_df,
    sqanti_df$gene_type == "protein_coding",
    "not_protein_coding",
    dropout_tracker
  )
}
  
if (args$filter_internal_priming) {
  message("Applying internal priming filter...")
  sqanti_df <- apply_filter(
    sqanti_df,
    sqanti_df$perc_A_downstream_TTS <= args$percent_polyA_threshold,
    "internal_priming",
    dropout_tracker
  )
}
  
if (args$filter_RTS) {
  message("Applying template switching filter...")
  sqanti_df <- apply_filter(
    sqanti_df,
    !sqanti_df$RTS_stage,  # Keep only FALSE (no RTS)
    "Template_switching_artifact",
    dropout_tracker
  )
}
  
# Apply structural category filter
message("Applying structural category filter...")
allowed_categories <- STRUCTURAL_CATEGORIES[[args$structural_level]]
sqanti_df <- apply_filter(
  sqanti_df,
  sqanti_df$structural_category %in% allowed_categories,
  "structural_category_filtered",
  dropout_tracker
)
  
# Calculate final results
kept_ids <- sqanti_df$isoform
all_dropout_ids <- setdiff(original_ids, kept_ids)
  
message(paste0("Final: ", length(kept_ids), " kept, ", length(all_dropout_ids), " dropped"))

# =============================================================================
# Write output files and generate summary
# =============================================================================

message("Writing output files...")
write_tsv(sqanti_df, file.path(dir, paste0(basename, "_classification_filtered.txt")))
  
save_filtered_fasta(sqanti_fasta, kept_ids, file.path(dir, paste0(basename, "_corrected_filtered.fasta")))
  
save_filtered_gtf(sqanti_gtf, kept_ids, file.path(dir, paste0(basename, "_corrected_filtered.gtf")))
  
# Save dropout files
dropout_reasons_df <- map_dfr(names(dropout_tracker), ~ {
  tibble(isoform = dropout_tracker[[.x]], dropout_reason = .x)
})

write_tsv(dropout_reasons_df, file.path(dropout_dir, paste0(basename, "_dropout_transcripts.tsv")))
save_filtered_fasta(sqanti_fasta, all_dropout_ids, file.path(dropout_dir, paste0(basename, "_corrected_dropout.fasta")))
save_filtered_gtf(sqanti_gtf, all_dropout_ids, file.path(dropout_dir, paste0(basename, "_corrected_dropout.gtf")))
  
# Print summary
cat("\n=== FILTERING SUMMARY ===\n")
cat(paste0("Sample: ", basename, "\n"))
cat(paste0("Original transcripts: ", length(original_ids), "\n"))
cat(paste0("Final transcripts: ", length(kept_ids), "\n"))
cat(paste0("Retention rate: ", round(100 * length(kept_ids) / length(original_ids), 1), "%\n"))
cat("\nDropout breakdown:\n") 

dropout_summary <- dropout_reasons_df %>%
  count(dropout_reason, sort = TRUE)
  
for (i in seq_len(nrow(dropout_summary))) {
  cat(paste0("  ", dropout_summary$dropout_reason[i], ": ", dropout_summary$n[i], "\n"))
    }


```

Why are there so many RT switching in the AML long read?
- Inspect support depth
- canonical or non-canonical junctions
- microhomology length distribution > 6 are more likely to be real

SQANTI’s RT-switching rule is purely sequence-based: if a junction is flanked by microhomology, it gets flagged. Many true biological junctions happen to sit near repeats, so you can get multiple reads flagged even though they’re real isoforms. This is the most common explanation when you see >1 read support.

A lot of these seem real to me
```{r}
# this is my analysis in LRP2_lite
all = read_tsv("../results/sqanti/sample_classification.txt") # 839,763
switch = all %>% filter(RTS_stage == "TRUE") # 542,120

# expression
ggplot(all, aes(x = RTS_stage, y = FL.BioSample_1)) + geom_violin() + theme_bw() + scale_y_log10()
switch %>% filter(FL.BioSample_1 > 1) # 153,179
switch %>% filter(FL.BioSample_1 > 5) # 27,193

# canonical splice sites, GT-AG
switch %>% group_by(all_canonical) %>% tally() # 509,234/542,120 are non-canonical
switch %>% filter(FL.BioSample_1 > 1) %>% group_by(all_canonical) %>% tally() # 138,400/153,179 are non-canonical
switch %>% filter(FL.BioSample_1 > 5) %>% group_by(all_canonical) %>% tally() # 20,923/27,193 are non-canonical

# structural category
switch %>% group_by(structural_category) %>% tally() # 355,038 are novel not in catalog

# consider all factors
switch %>% filter(FL.BioSample_1 > 1, all_canonical == "canonical") %>% group_by(structural_category) %>% tally() # 14,779


```
Compare to Emily's analysis- similar percentage
```{r}
em = read_tsv("~/sheynkman/projects/LRP_AML/02_sqanti/AML_classification.txt") # 1,689,238
em_switch = em %>% filter(RTS_stage == "TRUE") # 929,199

```

Compare to MDS Mouse data
```{r}
mds = read_tsv("~/sheynkman/projects/LRP_Mohi_project/02_sqanti/MDS_classification.txt") # 838,021
mds_switch = mds %>% filter(RTS_stage == "TRUE") # 443,272

# expression
mds_switch %>% filter(FL.BioSample_1 > 1) # 74,490
mds_switch %>% filter(FL.BioSample_1 > 5) # 15,122

# canonical splice sites, GT-AG
mds_switch %>% group_by(all_canonical) %>% tally() # 440,301/443,272 are non-canonical
mds_switch %>% filter(FL.BioSample_1 > 1) %>% group_by(all_canonical) %>% tally() 
mds_switch %>% filter(FL.BioSample_1 > 5) %>% group_by(all_canonical) %>% tally() 

# structural category
mds_switch %>% group_by(structural_category) %>% tally() # most are novel not in catalog

# consider all factors
mds_switch %>% filter(FL.BioSample_1 > 1, all_canonical == "canonical") %>% group_by(structural_category) %>% tally() # 1,359

```


From 05_orf_calling
- I dont think expression is useful here. Create an ORF-id that concatanates the transcript ids but keep each row a transcript. can aggregate on ORF ids later.
- dont call orfs separately, call all together

```{r}
#!/usr/bin/env Rscript

#' ORF Analysis Script - R Version
#' 
#' Map ORF coordinates to genomic positions
#' Compares to GENCODE start codon annotations
#' Selects the best ORF for each transcript
#'
#' Outputs 

# =============================================================================
# Load required libraries
# =============================================================================

suppressPackageStartupMessages({
  library(tidyverse)
  library(rtracklayer)
  library(Biostrings)
})

# =============================================================================
# Get environment variables
# =============================================================================

basename <- Sys.getenv("OUTPUT_BASE_NAME")
sqanti_dir <- file.path(Sys.getenv("OUTPUT_DIR"), "sqanti") 
cpat_dir <- file.path(Sys.getenv("OUTPUT_DIR"), "orf_calling") 
gencode_gtf_path <- Sys.getenv("GENCODE_GTF_FILE")
coding_threshold <- as.numeric(Sys.getenv("CPAT_CODING_THRESHOLD", "0.364"))

# =============================================================================
# Check required files
# =============================================================================

cpat_fasta_file <- file.path(cpat_dir, paste0(basename, "_cpat.ORF_seqs.fa"))
stopifnot("Input ORF FASTA file not found" = file.exists(cpat_fasta_file))

cpat_results_file <- file.path(cpat_dir, paste0(basename, "_cpat.ORF_prob.tsv"))
stopifnot("CPAT results file not found" = file.exists(cpat_results_file))

sample_full_fasta_path <- file.path(sqanti_dir, paste0(basename, "_corrected_filtered.fasta"))
stopifnot("Sample full fasta file not found" = file.exists(sample_full_fasta_path))

sample_gtf_path <- file.path(sqanti_dir, paste0(basename, "_corrected_filtered.gtf"))
stopifnot("Sample gtf file not found" = file.exists(sample_gtf_path))

classification_file <- file.path(sqanti_dir, paste0(basename, "_classification_filtered.txt"))
stopifnot("SQANTI classification file not found" = file.exists(classification_file))


# =============================================================================
# Helper functions
# =============================================================================

#' Check if ORF ends with stop codon
#' @param orf_fasta_file Path to ORF FASTA file
#' @return Data frame with ID and has_stop_codon columns
check_stop_codons <- function(orf_fasta_file) {
  message("Checking stop codons in ORF fasta sequences...")
  
  stop_codons <- c("TAG", "TAA", "TGA")
  orfs <- readDNAStringSet(orf_fasta_file)
  
  tibble(
    ID = str_split_fixed(names(orfs), "\t", 2)[,1],
    sequence = as.character(orfs)) %>%
    mutate(seq_length = nchar(sequence),
           last_codon = if_else(seq_length >= 3, str_sub(sequence, -3, -1), ""),
           has_stop_codon = last_codon %in% stop_codons) %>%
    select(ID, has_stop_codon)
}

#' Map ORF coordinates to genomic positions
#' This is a simplified version that focuses on the core mapping logic
#' @param orf_coords ORF coordinate data
#' @param sample_gtf GTF data for sample transcripts
#' @param gencode_gtf GTF data for GENCODE reference
#' @param sample_fasta Sample transcript sequences
#' @return Mapped ORF data with genomic coordinates

sample_gtf = import("../results/sqanti/sample_corrected_filtered.gtf") %>% as.data.frame()
gencode_gtf = import("/project/sheynkman/external_data/GENCODE_v47/gencode.v47.annotation.gtf") %>% as.data.frame()
 
map_orfs_to_genome <- function(orf_coords, sample_gtf, gencode_gtf, full_fasta) {
  message("Mapping ORFs to genomic coordinates...")
  
  # Extract exon information from sample GTF
  sample_exons <- sample_gtf %>%
    filter(type == "exon") %>%
    select(transcript_id, seqnames, start, end, strand) %>%
    mutate(exon_length = end - start + 1) %>%
    arrange(transcript_id, if_else(strand == "+", start, -start))
  
  # Calculate cumulative exon positions for coordinate mapping
  sample_exons %<>%
    group_by(transcript_id) %>%
    mutate(cumulative_length = cumsum(exon_length), prior_length = lag(cumulative_length, default = 0)) %>%
    ungroup()
  
  # Map ORF coordinates to genomic positions and pull number of upstream ATGs- strandedness here is tricky
  mapped_orfs <- orf_coords %>%
    left_join(sample_exons, by = c("isoform_id" = "transcript_id")) %>%
    filter(ORF_start > prior_length & ORF_start <= cumulative_length) %>%
    left_join(full_fasta, by = "isoform_id") %>%
    mutate(genomic_start = ifelse(strand == "+",
                                  start + (ORF_start - prior_length) - 1,
                                  end - (ORF_start - prior_length) + 1)) %>%
    mutate(upstream_atgs = ifelse(!is.na(full_sequence), 
                                  str_count(substr(full_sequence, 1, ORF_start - 1), "ATG"),
                                  Inf)) %>%
    select(-full_sequence)
  
  # Check for GENCODE start codon matches- looking for any match here
  gencode_start_codons <- gencode_gtf %>%
    filter(type == "start_codon") %>% 
    mutate(gencode_pos = ifelse(strand == "+", start, end)) %>%
    select(seqnames, strand, gencode_pos) %>%
    distinct() %>%
    mutate(gencode_match = TRUE)
    
  mapped_orfs %<>% left_join(gencode_start_codons, by = c("seqnames", "strand", "genomic_start" = "gencode_pos")) %>%
    mutate(gencode_match = ifelse(is.na(gencode_match), FALSE, gencode_match))
  
  return(mapped_orfs)
}

#' Select best ORF for each transcript
#' Logic: Prefer GENCODE matches with fewest upstream ATGs, otherwise highest coding score
#' @param mapped_orfs Data frame of mapped ORFs
#' @return Data frame with best ORF per transcript
call_best_orfs <- function(mapped_orfs) {
  message("Calling best ORFs for each transcript...")
  
  # ORF quality scoring parameters
  # but these don't seem to be used
  coding_threshold  # CPAT protein-coding threshold
  atg_shift <- 10          # Sigmoid parameters for ATG penalty
  atg_growth <- 0.5
  
  # require the ORF to have a stop codon
  # per transcript, if there is a gencode match, pick the one with the fewest upstream ATGs
  mapped_orfs %<>% 
    mutate(atg_penalty = 1 - 1/(1 + exp(-atg_growth * (upstream_atgs - atg_shift))), 
           composite_score = Coding_prob * (1 - atg_penalty)) %>%
    mutate(orf_quality = ifelse(!has_stop_codon, "No Stop Codon", 
                                ifelse(Coding_prob <= coding_threshold, "Low Quality ORF", "Plausible ORF")))
  
  best_orfs = mapped_orfs %>% 
    filter(orf_quality == "Plausible ORF") %>%
    group_by(isoform_id) %>% 
    arrange(desc(gencode_match), upstream_atgs, desc(Coding_prob)) %>%
    slice_head(n = 1) %>%
    ungroup()
  
  mapped_orfs %<>%
    mutate(orf_quality = ifelse(ID %in% best_orfs$ID, "Clear Best ORF", orf_quality))
  
  return(mapped_orfs)
}

# ===== MAIN ANALYSIS FUNCTION =====
cpat_fasta_file = "/home/cwp5au/megan/github/LRP2_lite/results/orf_calling/sample.ORF_seqs.fa"
cpat_results_file = "/home/cwp5au/megan/github/LRP2_lite/results/orf_calling/sample.ORF_prob.tsv"
sample_full_fasta_path = "/home/cwp5au/megan/github/LRP2_lite/results/sqanti/sample_corrected_filtered.fasta"
classification_file = "/home/cwp5au/megan/github/LRP2_lite/results/sqanti/sample_classification_filtered.txt"

main <- function() {

  # === STEP 1: Load all input data ===
  message("\n--- STEP 1: Loading input data ---")
  
  # cpat results
  cpat_results = read_tsv(cpat_results_file, show_col_types = FALSE) %>%
    separate(ID, into = c("isoform_id", "orf_rank"), sep = "_ORF_", remove = FALSE)
  stop_codon_status <- check_stop_codons(cpat_fasta_file)
  orf_coords <- left_join(cpat_results, stop_codon_status, by = "ID")
  
  # Load GTF files
  sample_gtf <- import(sample_gtf_path) %>% as.data.frame()
  gencode_gtf <- import(gencode_gtf_path) %>% as.data.frame()
  
  # Load sample fasta with full sequence
  full <- readDNAStringSet(sample_full_fasta_path)
  full_fasta = tibble(
    isoform_id = names(full),
    full_sequence = as.character(full)
  )
  
  # Load sqanti classification files
  classification <- read_tsv(classification_file, show_col_types = FALSE)
  
  # === STEP 2: Map ORFs to genomic coordinates ===
  message("\n--- STEP 2: Mapping ORFs to genome ---")
  
  mapped_orfs <- map_orfs_to_genome(orf_coords, sample_gtf, gencode_gtf, full_fasta)

  # === STEP 3: Call best ORFs ===
  message("\n--- STEP 3: Calling best ORFs ---")
  
  best_orfs <- call_best_orfs(mapped_orfs)
  original_n_transcripts = n_distinct(best_orfs$isoform_id)
  n_best_orfs = best_orfs %>% filter(orf_quality == "Clear Best ORF") %>% distinct(isoform_id) %>% nrow()
  
  write_tsv(best_orfs, file.path(cpat_dir, paste0(basename, "_all_cpat_orfs_mapped.tsv")))
  
  # === SUMMARY ===
  message("\n=== ANALYSIS COMPLETE ===")
  message(paste0("Started with ORFs called for ", original_n_transcripts, " transcripts."))
  message(paste0("After filtering, ", n_best_orfs, " transcripts have a 'Clear Best ORF'")) 
  message(paste0("All mapped ORFs: ", basename, "_all_cpat_orfs_mapped.tsv"))
  message("Analysis completed successfully!")
}

# Run main function
if (!interactive()) {
  main()
}


```

Checking strand here
```{r}
# testing
mapped_orfs %>% filter(gencode_match == TRUE) %>% group_by(strand) %>% tally()

# negative stranded gene
sample_exons %>% filter(grepl("^PB\\.27\\.", transcript_id)) %>% arrange(desc(end))
test_neg = orf_coords %>% filter(ID == "PB.27.42_ORF_1") %>%
    left_join(sample_exons, by = c("isoform_id" = "transcript_id")) %>%
    filter(ORF_start > prior_length & ORF_start <= cumulative_length) %>%
    left_join(full_fasta, by = "isoform_id") %>%
    mutate(genomic_start = ifelse(strand == "+",
                                  start + (ORF_start - prior_length) - 1,
                                  end - (ORF_start - prior_length) + 1)) %>%
    mutate(upstream_atgs = ifelse(!is.na(full_sequence), 
                                  str_count(substr(full_sequence, 1, ORF_start - 1), "ATG"),
                                  Inf)) %>%
    select(-full_sequence)

gencode_test = gencode_gtf %>%
    filter(type == "start_codon") %>% 
    mutate(gencode_pos = ifelse(strand == "+", start, end)) %>%
    select(gene_id, transcript_id, seqnames, strand, gencode_pos) %>%
    distinct()

gencode_test %>% filter(strand == "-", gencode_pos == 959240)
gencode_gtf %>% filter(gene_id == "ENSG00000188976.11", type == "transcript")

# positive stranded gene
orf_coords %>% filter(ID == "PB.100.1_ORF_1") %>%
    left_join(sample_exons, by = c("isoform_id" = "transcript_id")) %>%
    filter(ORF_start > prior_length & ORF_start <= cumulative_length) %>%
    left_join(full_fasta, by = "isoform_id") %>%
    mutate(genomic_start = ifelse(strand == "+",
                                  start + (ORF_start - prior_length) - 1,
                                  end - (ORF_start - prior_length) + 1)) %>%
    mutate(upstream_atgs = ifelse(!is.na(full_sequence), 
                                  str_count(substr(full_sequence, 1, ORF_start - 1), "ATG"),
                                  Inf)) %>%
    select(-full_sequence)
```

Refine ORFs
```{r}
#!/usr/bin/env Rscript

# ORF Database Refinement Script - R Version
# Groups transcripts with identical protein sequences and filters by quality

suppressPackageStartupMessages({
  library(tidyverse)
  library(Biostrings)
  library(argparse)
})


#' Translate ORF sequences to proteins and group by identical sequences
#' @param orfs Data frame with ORF coordinates
#' @param transcript_seqs Named vector of transcript sequences
#' @return Data frame with protein sequences and grouped transcript IDs
group_by_protein_sequence <- function(filtered_results, full_fasta) {
  message("Translating ORFs and grouping by protein sequence...")
  
  orf_proteins = filtered_results %>%
    select(isoform_id, ORF_start, ORF_end) %>%
    left_join(full_fasta, by = "isoform_id") %>%
    filter(!is.na(full_sequence)) %>%
    mutate(orf_dna_sequence = substr(full_sequence, ORF_start, ORF_end))
    
  # Vectorized translation for speed
  dna <- DNAStringSet(orf_proteins$orf_dna_sequence)
  proteins <- translate(dna, if.fuzzy.codon = "solve")
  
  orf_proteins %<>%
    mutate(orf_protein_sequence = as.character(proteins)) %>%
    filter(orf_protein_sequence != "") %>%
    select(isoform_id, orf_protein_sequence)
  
  
  # Group transcripts by identical protein sequences
  orf_groups = orf_proteins %>%
    group_by(orf_protein_sequence) %>%
    arrange(isoform_id) %>%
    mutate(
      orf_isoform_id = paste(isoform_id, collapse = "|"),
      orf_base_id = isoform_id[1]  # First transcript as representative
    ) %>% 
    ungroup()
  
  return(orf_groups)
}


#' Calculate total expression for grouped transcripts
#' @param protein_groups Data frame with protein groups
#' @param orfs Data frame with expression data
#' @return Data frame with aggregated expression values
aggregate_expression <- function(protein_groups, orfs) {
  message("Aggregating expression values for protein groups...")
  
  # Create lookup tables for FL and CPM values
  fl_lookup <- setNames(orfs$FL, orfs$pb_acc)
  cpm_lookup <- setNames(orfs$CPM, orfs$pb_acc)
  
  # Aggregate expression for each protein group
  expression_summary <- protein_groups %>%
    mutate(
      FL = map_dbl(pb_accs, ~sum(fl_lookup[.x], na.rm = TRUE)),
      CPM = map_dbl(pb_accs, ~sum(cmp_lookup[.x], na.rm = TRUE))
    )
  
  return(expression_summary)
}

#' Main analysis function
main <- function() {

  message("=== Starting ORF Database Refinement ===")

  # === STEP 1: Load input data ===
  message("\n--- Loading input files ---")
  orfs <- read_tsv(args$orfs, show_col_types = FALSE)
  message("Loaded ", nrow(orfs), " ORFs")
  
  all_ids <- unique(orfs$pb_acc)
  message("Total unique transcripts: ", length(all_ids))
  
  # === STEP 2: Filter ORFs by quality ===
  message("\n--- Filtering ORFs by quality ---")
  # Apply quality filters
  filtered_results = best_orfs %>% filter(orf_quality %in% c("Plausible ORF", "Clear Best ORF"))
  
  # === STEP 4: Group transcripts by protein sequence ===
  message("\n--- Grouping by protein sequence ---")
  orf_groups <- group_by_protein_sequence(filtered_results, full_fasta)
  
  
  
  
  # === STEP 5: Aggregate expression data ===
  message("\n--- Aggregating expression data ---")
  protein_groups_with_expr <- aggregate_expression(protein_groups, kept_orfs)
  
  # === STEP 6: Merge with ORF metadata ===
  message("\n--- Merging with ORF metadata ---")
  
  # Get metadata for representative (base) transcripts
  orf_metadata <- kept_orfs %>%
    select(pb_acc, coding_score, orf_score, orf_calling_confidence, upstream_atgs, gene) %>%
    distinct()
  
  final_results <- protein_groups_with_expr %>%
    left_join(orf_metadata, by = c("base_acc" = "pb_acc")) %>%
    select(pb_accs_str, base_acc, protein_sequence, coding_score, orf_calling_confidence,
           upstream_atgs, orf_score, gene, FL, CPM) %>%
    rename(pb_accs = pb_accs_str)
  
  # === STEP 7: Write final outputs ===
  message("\n--- Writing output files ---")
  
  # Write refined protein FASTA
  refined_fasta <- paste0(args$name, "_orf_refined.fasta")
  fasta_lines <- paste0(">pb|", final_results$base_acc, "|fullname GN=", final_results$gene, "\n",
                       final_results$protein_sequence)
  writeLines(fasta_lines, refined_fasta)
  
  # Write refined metadata TSV
  refined_tsv <- paste0(args$name, "_orf_refined.tsv")
  write_tsv(final_results, refined_tsv)
  
  # === SUMMARY ===
  message("\n=== Refinement Complete ===")
  message("Final outputs:")
  message("- Refined FASTA: ", refined_fasta)
  message("- Refined metadata: ", refined_tsv)
  message("- Dropout FASTA: dropout_", basename(args$name), "_orf.fasta")
  message("- Dropout metadata: dropout_", basename(args$name), "_orf.tsv")
  message("")
  message("Summary:")
  message("- Input ORFs: ", nrow(orfs))
  message("- Quality filtered: ", nrow(kept_orfs))
  message("- Final protein groups: ", nrow(final_results))
  message("- Redundancy reduction: ", round((1 - nrow(final_results)/nrow(kept_orfs)) * 100, 1), "%")
}

# Run main function
if (!interactive()) {
  main()
}
```

Problem is that we have redundancy...some transcripts have more than one probably ORF
```{r}
#' Translate ORF sequences to proteins and group by identical sequences
#' @param orfs Data frame with ORF coordinates
#' @param transcript_seqs Named vector of transcript sequences
#' @return Data frame with protein sequences and grouped transcript IDs
group_by_protein_sequence <- function(filtered_results, full_fasta) {
  message("Translating ORFs and grouping by protein sequence...")
  
  orf_proteins = filtered_results %>%
    select(isoform_id, ORF_start, ORF_end) %>%
    left_join(full_fasta, by = "isoform_id") %>%
    filter(!is.na(full_sequence)) %>%
    mutate(orf_dna_sequence = substr(full_sequence, ORF_start, ORF_end))
    
  # Vectorized translation for speed
  dna <- DNAStringSet(orf_proteins$orf_dna_sequence)
  proteins <- translate(dna, if.fuzzy.codon = "solve")
  
  orf_proteins %<>%
    mutate(orf_protein_sequence = as.character(proteins)) %>%
    filter(orf_protein_sequence != "") %>%
    select(isoform_id, orf_protein_sequence)
  
  
  # Group transcripts by identical protein sequences
  orf_groups = orf_proteins %>%
    group_by(orf_protein_sequence) %>%
    arrange(isoform_id) %>%
    mutate(
      orf_isoform_id = paste(isoform_id, collapse = "|"),
      orf_base_id = isoform_id[1]  # First transcript as representative
    ) %>% 
    ungroup()
  
  return(orf_groups)
}






```










